### 2020-04-07 Paper.Rmd
First commit. Adding two data sets, the master, and the paper template.

---
title: "Paper"
author: "Ana Karina"
date: "4/7/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-04-07 master.R

@@ -1,0 +1,3 @@
+# Master code 
+
+source ("src/XXX.R")

### 2020-04-09 Paper.Rmd
Second commit. Adding paper, references, and first draft of the process for analysing hand to saliva data.

@@
 output: html_document
+bibliography: references.bib
 ---
@@
 knitr::opts_chunk$set(echo = TRUE)
+require(tidyverse)
 ```
 
-## R Markdown
+# Risk assessment for indirect transmission of SARS-CoV-2
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+References 
+@Adhikari2019
+@Chen2004
+@Ong2020
+@Beale2020
+@Kampf2020
+@Kwok2015
+@Wu2020
+@Yang2020
+@Yeo2020
+@VanDoremalen2020
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+## Introduction
 
-```{r cars}
-summary(cars)
+Direct transmission (sneeze, cough, droplet inhalation) and indirect transmission (contact with surfaces and possible contaminated air).
+
+Hand washing has been proposed as one of the primary strategies to control transmission (REF).
+Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020.
+But there is not conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
+
+Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.
+
+Air samples negative, but surface samples positive @Ong2020.
+
+Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
+
+“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.
+
+"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020
+
+Human coronaviruses can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020. Therefore, touching contaminated surfaces offers a risk for exposure.
+
+SARS-CoV-1 transmission has been associated with nosocomial spread.
+Transmission data suggests that, although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004.
+
+A study of an outbreak of SARS CoV in Tawian showed that, out of the 232 SARS patients in the hospital, 31 patients (13 %)  were hospital acquired transmissions.
+Out of the 31 cases, most of the transmission occurred by direct contact with a known SARS patient, but 6/31 (~20%) of the transmissions could not be traced to direct contact with SARS patients.
+Sampling of surfaces for the RNA of coronavirus showed that surfaces such as drinking water buttons, beds and chairs were contaminated with coronavirus.
+Thus, the researchers postulated that indirect contact could have been the source of transmission @Chen2004.
+
+### QMRA 
+A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus.
+They only considered the risk of infection due from droplets and aer(osolized particles, since that is the main transmission route.
+Most of the uncertainty associated with the model was due to the concentration of virus in the saliva.
+The researchers evaluated possible interventions and showed that the use of mask contributed significantly to the reduction of the risk (>90%), as well as increasing air exchange (up to 58% risk reduction).
+
+### Dose-response for SARS coronavirus
+“For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”
+
+\begin{equation}
+P_{inf}=1 - e^{-kd}
+\end{equation}
+
+```{r dose-response, include=FALSE}
+# Dose-Response Parameter (exponential): (0.00246 (0.00135; 0.00459)),5th, 50th, and 95th percentile values, Huang (2013), wikiQMRA,
+
 ```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-04-09 src/process_handtosaliva.R

@@ -1,0 +1,13 @@
+
+# Processing data of virus transfer from hands to saliva
+
+# Data taken from Pitol, 2017. Available at <https://zenodo.org/record/1454505#.Xo3wMZNKjBI>
+
+# dt = drying time for average droplet size (min), assumtion. 
+
+dt= 10
+
+TE_hm = m * TE_hm_d + (1-m) * TE_hm_w
+if t = 0, m=0,
+if 
+

### 2020-04-10 Paper.Rmd
Third commit, adding more data sets and scripts.

@@
 Direct transmission (sneeze, cough, droplet inhalation) and indirect transmission (contact with surfaces and possible contaminated air).
+
+Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.
+
+Coronaviruses spread between humans through direct (e.g., respiratory droplets) and indirect contacts (e.g., hand-to-surface, fomite-to-face).
+Although the use of face masks, and hand hygiene has been suggested as the primary interventions to reduce coronavirus transmission, little is known regarding the contribution of indirect (hand-to-surface, fomite-to-face) transmission.
 
@@
 
+The dose-response relationship was obtain from QMRA wiki <http://qmrawiki.org/experiments/sars>.
+The model is based on animal dose-response relationship.
+No human dose-response relationship is available for SARS Coronavirus.
+It is generally accepted to use animal models
+
 \begin{equation}
@@
 ```
+
+## Limitations
+Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.
+
+Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.
+
+Since no randomized control trials will be runned and there are limits to the conclusions we can reach using epidemiological data, risk assesments provide an attempt to understand the risk associated with different activities.
+The results of this paper could inform policy.
+For example, trafick lights in the crossing stations could clausured momentaniely, the same than train buttons,
+
+The dependence of temperature and humidity in the transfer rate of coronavirus is unclear.
+If temperature and humidity may play a role in non-dominant transmission pathways, such as indirect transfer, it would be hard to observe epidemiological studies.
+It is known that the survival of viruses on surfaces is dependent on environmental factors such as temperature and humidity (REF).
+Similarly, the transfer rate of viruses varies with humidity (REF).
+NEvertheless, if only a small fraction of the transmission occurs through indirect transfer we wont see it.
+
+We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countribution of each route.

### 2020-04-10 src/process_TEhandtosaliva.R

@@ -1,0 +1,45 @@
+
+# Processing data of virus transfer from hands to saliva
+# Data taken from Pitol, 2017. Available at <https://zenodo.org/record/1454505#.Xo3wMZNKjBI>
+
+# setup -------------------------------------------------------------------
+# external libraries
+# library(tidyverse) 
+
+# read and process data  --------------------------------------------------
+infile = "data/raw/TEhandtosaliva.csv"
+#outfile = "data/processed/TEhandtosaliva.csv"
+
+# read data 
+df_TE_hm <- read.csv (file=infile, header = TRUE, sep=",")
+
+#-----------------
+
+# dryingtime [min] = drying time for average droplet size, assumtion. 
+# TE_hm_d [%] = transfer efficiency from hand to mucous membrane (dry)
+# TE_hm_w [%] = transfer efficiency from hand to mucous membrane (wet)
+# TE_hm [%]= transfer efficiency from hand to mucous membrane (dry and wet)
+
+TE_hm_d <- mean(df_TE_hm$pt_dry)
+sd_TE_hm_d <- sd(df_TE_hm$pt_dry)
+
+TE_hm_w <- mean(df_TE_hm$pt_wet)
+sd_TE_hm_d <- sd(df_TE_hm$pt_wet)
+
+# Calculating the TE_hm as a function of time 
+dryingtime= 10
+
+if t <= dryingtime, m(t) = t/dryingtime 
+if t > dryingtime, m(t)=1
+
+TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
+sd_TE_hm(t)=
+
+    
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)
+
+    
+    

### 2020-04-10 src/process_TEsurfacetohand.R

@@ -1,0 +1,20 @@
+
+# Processing data of virus transfer from surface to hand 
+# Data obtained from the article (Lopez, 2013, doi: 10.1128/AEM.01030-13)
+
+# setup -------------------------------------------------------------------
+# external libraries
+
+# read and process data  --------------------------------------------------
+infile = "data/raw/TEsurfacetohand.csv"
+outfile = "data/processed/TEsurfacetohand.csv"
+
+# read data 
+df_TE_sh <- read.csv (file=infile, header = TRUE, sep=",")
+
+
+
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)

### 2020-04-10 src/process_decay.R

@@ -1,0 +1,25 @@
+
+# Processing data decay rate of SARS-CoV-2 on different surfaces
+
+# Data obtained from the Suplementary Appendix of the article 
+
+
+# setup -------------------------------------------------------------------
+# external libraries
+
+# read and process data  --------------------------------------------------
+infile = "data/raw/decay.csv"
+#outfile = "data/processed/decay.csv"
+
+# read data 
+df_decay <- read.csv (file=infile, header = TRUE, sep=",")
+
+# Equations for first order decay
+# C(t) = C(0) e^ (-kt)
+# halflife = ln2 / k
+# k = ln2 / halflife
+
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)

### 2020-04-10 src/process_handtosaliva.R

@@ -1,13 +1,0 @@
-
-# Processing data of virus transfer from hands to saliva
-
-# Data taken from Pitol, 2017. Available at <https://zenodo.org/record/1454505#.Xo3wMZNKjBI>
-
-# dt = drying time for average droplet size (min), assumtion. 
-
-dt= 10
-
-TE_hm = m * TE_hm_d + (1-m) * TE_hm_w
-if t = 0, m=0,
-if 
-

### 2020-04-13 src/process_concentrationsaliva.R
Adding data set for the concentration of viruses in infected individuals

@@ -1,0 +1,19 @@
+# Processing data of concentration of SARS-CoV-2 samples in sputum (taken as saliva)
+# Data obtained from: XXX
+
+
+# setup -------------------------------------------------------------------
+# external libraries
+
+# read and process data  --------------------------------------------------
+infile = "data/raw/concentrationsaliva.csv"
+#outfile = "data/processed/decay.csv"
+
+# read data 
+df_C_sp <- read.csv (file=infile, header = TRUE, sep=",")
+
+
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)

### 2020-04-15 src/process_concentrationsaliva.R
Updating Paper with new reference and change in the decay data set

@@ -1,6 +1,7 @@
 # Processing data of concentration of SARS-CoV-2 samples in sputum (taken as saliva)
-# Data obtained from: XXX
-
+# Data obtained from: 
+    #"Viral load of SARS-CoV-2 in clinical samples" (Pan, 2020) 
+    #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020)
 
 # setup -------------------------------------------------------------------
 # external libraries
@@ -10,7 +11,7 @@ infile = "data/raw/concentrationsaliva.csv"
 #outfile = "data/processed/decay.csv"
 
 # read data 
-df_C_sp <- read.csv (file=infile, header = TRUE, sep=",")
+df_C_sp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
 
 
 # output -------------------------------------------------------------------

### 2020-04-16 Paper.Rmd
Updating Paper with new reference and change in the decay data set
Adding first draft of ATM dataset

@@
 @VanDoremalen2020
+
+Risk assessment parameters
+@AuYeung2008  -> fractional surface area of finger
+@
 
@@
 
+He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
+with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case.
+More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"
+
 ### Dose-response for SARS coronavirus

### 2020-04-20 src/Risk.R
Calculating Risks

@@ -1,0 +1,48 @@
+
+require(tidyverse)
+
+## Parameters Used
+
+# Reading the data frames 
+source ("src/process_concentrationsaliva.R") # Wölfel 2020 and Pan 2020
+source ("src/process_TEhandtosaliva.R") # Pitol 2017
+source ("src/process_TEsurfacetohand.R") # Lopez 2013
+source ("src/process_ATM_observation.R") # Observation ATM
+source ("src/process_decay.R") # van Doremalen 2020
+
+# Simulation parameters
+simNum = 100 #number of simulations
+
+# Constant parameters
+GC_TCID50 <- 0.01 # Proportion of infective gene copies [GC:TCID50]
+cough_surf_ang <- pi/2 
+t_k <- 2
+x_k <- 4
+
+# Functions
+f_Cs0 <- function (Ncough_min, Csp, Vs) {
+    Cs_0 <- (Ncough_min * Csp * Vs)
+    return(Cs_0)
+}
+
+# Parameters from distributions and datasets
+Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
+Ncough_min <- rnorm(simNum, 0.57, 1) # Number of coughs per minute, 
+Csp <- sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Concentration of virus in sputum [GC/ml] 
+TEsh_steel <- rnorm(simNum, df_TE_sh$pt_mean[14], df_TE_sh$pt_sd[14]) # stainless steel RH=[40-65%]
+
+
+df <- as.data.frame(cbind(Vs, Ncough_min, Csp, TEsh_steel))
+df$Cs_0 <- NA
+df$Cs_1 <- NA
+
+
+for (i in 1:simNum )
+{
+    df$Cs_0[i] <- f_Cs0(Ncough_min[i], Csp[i], Vs[i])
+    df$Cs_1[i] <- df$Cs_0[i] + 1
+}
+
+
+
+

### 2020-04-20 src/process_ATM_observation.R

@@ -1,0 +1,14 @@
+# Processing data obtained from the ATM observation study
+
+# read and process data  --------------------------------------------------
+infile = "data/raw/ATM_observation.csv"
+#outfile = "data/processed/XX.csv"
+
+# read data 
+df_ATM <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
+
+
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)

### 2020-04-20 Paper.Rmd
Updating info

@@
 
+Part of the WHO COVID-19 strategy is to "engaging and educating the public to ensure protective measures are implemented effectively as changes are made to existing policies and restrictions." @WHO2020a
+It is necesary to transition into more flexible measures.
+
 ### QMRA 
@@
 We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countribution of each route.
+
+#### Viral shedding and infection potential 
+Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
+The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF).
+
+In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.

### 2020-04-20 src/process_TEhandtosaliva.R

@@ -27,13 +27,13 @@ TE_hm_w <- mean(df_TE_hm$pt_wet)
 sd_TE_hm_d <- sd(df_TE_hm$pt_wet)
 
 # Calculating the TE_hm as a function of time 
-dryingtime= 10
+# dryingtime= 10
 
-if t <= dryingtime, m(t) = t/dryingtime 
-if t > dryingtime, m(t)=1
+# if t <= dryingtime, m(t) = t/dryingtime 
+# if t > dryingtime, m(t)=1
 
-TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
-sd_TE_hm(t)=
+# TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
+# sd_TE_hm(t)=
 
     
 # output -------------------------------------------------------------------

### 2020-04-20 src/process_concentrationsaliva.R

@@ -11,7 +11,7 @@ infile = "data/raw/concentrationsaliva.csv"
 #outfile = "data/processed/decay.csv"
 
 # read data 
-df_C_sp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
+df_Csp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
 
 
 # output -------------------------------------------------------------------

### 2020-05-03 Paper.Rmd
Updating the analysis

--- 2020-04-16 Paper.Rmd
@@
 
-# Risk assessment for indirect transmission of SARS-CoV-2
+# Risk of infection from touching surfaces contaminated with SARS-CoV-2
 
-References 
-@Adhikari2019
-@Chen2004
-@Ong2020
-@Beale2020
-@Kampf2020
-@Kwok2015
-@Wu2020
-@Yang2020
-@Yeo2020
-@VanDoremalen2020
-
-Risk assessment parameters
-@AuYeung2008  -> fractional surface area of finger
-@
+## Abstract
 
@@
 
-Direct transmission (sneeze, cough, droplet inhalation) and indirect transmission (contact with surfaces and possible contaminated air).
+If no contrameasures are taken, viruses can spread thorugh the space populated by the human host at a high speed, as we have observed @BarOn2020.
 
-Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.
-
-Coronaviruses spread between humans through direct (e.g., respiratory droplets) and indirect contacts (e.g., hand-to-surface, fomite-to-face).
-Although the use of face masks, and hand hygiene has been suggested as the primary interventions to reduce coronavirus transmission, little is known regarding the contribution of indirect (hand-to-surface, fomite-to-face) transmission.
-
-Hand washing has been proposed as one of the primary strategies to control transmission (REF).
-Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020.
-But there is not conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
-
-Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.
-
-Air samples negative, but surface samples positive @Ong2020.
-
-Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
-
-“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.
-
-"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020
-
-Human coronaviruses can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020. Therefore, touching contaminated surfaces offers a risk for exposure.
+The CDC and WHO have highligthed two main transmission routes: direct transmission or person-to-person transmission (respiratory droplets produced while talking, caughin, or sneezing), and indirect transmission or surface mediated transmission (touching a contaminated surface followed by touching mucose membrains in the face such as mouth or nose) @WHO2020 @CDC2020.
 
@@
 Transmission data suggests that, although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004.
-
 A study of an outbreak of SARS CoV in Tawian showed that, out of the 232 SARS patients in the hospital, 31 patients (13 %)  were hospital acquired transmissions.
@@
 
-### QMRA 
+Therefore, touching contaminated surfaces offers a risk for exposure. 
+Inspite its importance, there is not a lot of information regarding the magnitud of risk associeted with touching contaminated surfaces @BarOn2020.
+
+Hand hygiene interventions and surface desinfection have been hihgligthed as one of the primary strategies to control transmission @Mcclellan2020.
+Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020.
+But there is not conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
+
 A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus.
@@
 
-He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
-with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case.
-More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"
+## Materials and Methods 
+### Hazard Identification
+This risk assessment is designed to explore the risks associated indirect transmission of SARS-CoV-2.
 
-### Dose-response for SARS coronavirus
+Air samples negative, but surface samples positive @Ong2020.
+Human coronaviruses can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020.
+
+### Exposure Assessment 
+Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.
+
+Two expossure scenarios were modeled. 
+*Scenario 1. Single touch.*
+Frequently touched buttons (Traffic light buttons, train buttons). Supermarket products.
+
+*Scenatio 2. Multiple touches.*
+ATM machines.
+
+(Insert diagram)
+
+#### SARS-CoV-2 Concentration in saliva samples
+
+*Conversing viral RNA data to infectious viruses*
+There was a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
+Therefore, we had to translate the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
+Althoug not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF).
+To do this we ...
+
+*Viral shedding and infection potential* 
+Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
+The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF).
+In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
+
+### Dose Response for SARS-CoV
 “For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”
@@
 No human dose-response relationship is available for SARS Coronavirus.
-It is generally accepted to use animal models
+It is generally accepted to use animal models in QMRA if human data is not available (REF)
 
@@
 
+### Risk Characterization 
+Monte Carlo simulation method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.
+
+### Sensitivity analysis
+
+A sensitivity analysis was performed to investigate hot the variabilty and uncertainty of the parameters in the model onfluenced the estimated risks.
+The sensitivity was estimated using Spearman correlation coefficients between the inputs and outputs of the model.
+
+( Insert table of parameters )
+
+## Results and Discussion
 ```{r dose-response, include=FALSE}
-# Dose-Response Parameter (exponential): (0.00246 (0.00135; 0.00459)),5th, 50th, and 95th percentile values, Huang (2013), wikiQMRA,
 
@@
 
-## Limitations
-Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.
-
-Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.
-
 Since no randomized control trials will be runned and there are limits to the conclusions we can reach using epidemiological data, risk assesments provide an attempt to understand the risk associated with different activities.
 The results of this paper could inform policy.
-For example, trafick lights in the crossing stations could clausured momentaniely, the same than train buttons,
+For example, trafick lights in the crossing stations could clausured momentaniely, the same than train buttons.
 
@@
 Similarly, the transfer rate of viruses varies with humidity (REF).
-NEvertheless, if only a small fraction of the transmission occurs through indirect transfer we wont see it.
+Nevertheless, if only a small fraction of the transmission occurs through indirect transfer we wont see it.
+
+The sensitivity analysis shows that XXX parapeters are responsible for XX % of the risk variation of the model. Collecting more data on these parameters could reduce uncertainty.
+
+It is important to notice that the effectiveness of interventions such as hand washing depends not only on many factors which include: the hand-wasihng technique, the suceptibility of the virus to phisical and chemical desinfection, and the relative contribution of the indirect transfer of viruses to the overal infection risk.
+As it has been pointed out, thand hygiene should be promoted but the risk of overemphasizing it without more intormation could turn out to be negative (Yang, 2020).
+...
+("risk compensation"?).
+
+### Limitations
+Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.
 
 We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countribution of each route.
+
+Even though this risk assessment model relies on many assumptions, some with high uncertainty around them, and the risk caucluation is hihgly simplidied, this model can: provide insigts into the possible effects that different interventions can have in reducing the spread of SARS-CoV-2, serve as a first step in the pathway of building more complex models that allow exploring the relative contribution that different transmission routes could have in spreddin the disease.
+
+## Supporting Information
+## Author Information
+## Aknowledgements 
+## References
+
+#### More info that could be added
+
+Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.
+
+Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
+“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.
+
+"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020
+
+Part of the WHO COVID-19 strategy is to "engaging and educating the public to ensure protective measures are implemented effectively as changes are made to existing policies and restrictions." @WHO2020a
+
+He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
+with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case.
+More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"
+
+Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.

### 2020-05-03 src/Risk.R

@@ -1,4 +1,4 @@
-
+## Set-up
 require(tidyverse)
 
 ## Parameters Used
@@ -15,24 +15,45 @@ simNum = 100 #number of simulations
 
 # Constant parameters
 GC_TCID50 <- 0.01 # Proportion of infective gene copies [GC:TCID50]
-cough_surf_ang <- pi/2 
-t_k <- 2
-x_k <- 4
+cough_ang <- pi/2 # 
+t <- 2 # Time 
+x <- 4
 
-# Functions
+# Functions (relationships between variables)
 f_Cs0 <- function (Ncough_min, Csp, Vs) {
     Cs_0 <- (Ncough_min * Csp * Vs)
     return(Cs_0)
 }
+
+f_Cs1 <- function(Cs_0, n, t) {
+    Cs_1 <- (Cs_0 * exp(-nt))
+    return(Cs_1)
+}
+
+f_Nf <- function(Cs_1, SAf, TEsh) {
+    Cs_1 <- (Cs_0 * exp(-nt))
+    return(Cs_1)
+}
+
+f_TE_hm <- function(t, TE_hm_d, TE_hm_w, sd_TE_hm_d, sd_TE_hm_w) {
+    m <- t/dryingtime 
+    TE_hm_mean <- m * TE_hm_d + (1-m) * TE_hm_w
+    TE_hm_sd <-  m * sd_TE_hm_d + (1-m) * sd_TE_hm_w # Check how to compute mean of sd
+    return(TE_hm_mean, TE_hm_sd)
+}
+
 
 # Parameters from distributions and datasets
 Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
 Ncough_min <- rnorm(simNum, 0.57, 1) # Number of coughs per minute, 
 Csp <- sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Concentration of virus in sputum [GC/ml] 
 TEsh_steel <- rnorm(simNum, df_TE_sh$pt_mean[14], df_TE_sh$pt_sd[14]) # stainless steel RH=[40-65%]
+t <- runif(simNum, 0, 30) # Decay time in the surface
+t_0.5_steel <- runif(simNum, df_decay$lowCI[4], df_decay$highCI[4]) # Must correct this
+n <- 
 
-
[…]

### 2020-05-03 src/analyse_risk .R

@@ -1,0 +1,74 @@
+
+
+# setup -------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+
+# read and process data  --------------------------------------------------
+source ("src/process_concentrationsaliva.R") # Wölfel 2020 and Pan 2020
+#source ("src/process_TEhandtosaliva.R") # Pitol 2017
+source ("src/process_TEsurfacetohand.R") # Lopez 2013
+source ("src/process_ATM_observation.R") # Observation ATM
+source ("src/process_decay.R") # van Doremalen 2020
+
+##    Scenario 1 : Single touch, ATM machine 
+
+# Simulation parameters
+simNum = 100 #number of simulations
+
+# Constant parameters
+GC_TCID50 <- 0.01 # Proportion of infective gene copies [GC:TCID50]
+cough_ang <- pi/2 # 
+t <- 2 # Time since inoculation 
+t_k <- 5 # Time doing the activity
+x <- 4 # Inoculation distance
+
+# Dose-Response Parameter (exponential): 
+# (0.00246 (0.00135; 0.00459)),5th, 50th, and 95th percentile values, Huang (2013), wikiQMRA, 
+
+# Parameters from distributions and datasets
+Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
+Ncough_min <- rnorm(simNum, 0.57, 1) # Number of coughs per minute, Leung (2020)
+Csp <- sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Concentration of virus in sputum [GC/ml] 
+TEsh_stl <- rnorm(simNum, df_TE_sh$pt_mean[14], df_TE_sh$pt_sd[14]) # TE from surface to hand for stainless steel RH=[40-65%]
+t_0.5_stl <- runif(simNum, df_decay$lowCI[4], df_decay$highCI[4]) # Must correct this
+FSAsf <- runif(simNum, 0.5, 1) # Fraction of the finger in contact with the ATM --> correct this
+k <- rnorm(simNum, 2.46*10^-3, 2.46*10^-3) # correct, use the median and quartiles to feed a normal
+
+# Parameters from two distributions 
+#TEhm <- f_TEhm(t)
[…]

### 2020-05-03 src/process_TEhandtosaliva.R

@@ -9,31 +9,41 @@
 # read and process data  --------------------------------------------------
 infile = "data/raw/TEhandtosaliva.csv"
 #outfile = "data/processed/TEhandtosaliva.csv"
+#function TEhm
 
 # read data 
 df_TE_hm <- read.csv (file=infile, header = TRUE, sep=",")
 
 #-----------------
 
-# dryingtime [min] = drying time for average droplet size, assumtion. 
+# dt [min] = drying time for average droplet size, assumtion. 
 # TE_hm_d [%] = transfer efficiency from hand to mucous membrane (dry)
 # TE_hm_w [%] = transfer efficiency from hand to mucous membrane (wet)
 # TE_hm [%]= transfer efficiency from hand to mucous membrane (dry and wet)
 
-TE_hm_d <- mean(df_TE_hm$pt_dry)
-sd_TE_hm_d <- sd(df_TE_hm$pt_dry)
+TE_hm_d <- mean(df_TE_hm$pt_dry) # mean TEhm when dry
+sd_TE_hm_d <- sd(df_TE_hm$pt_dry) # sd TEhm when dry
 
-TE_hm_w <- mean(df_TE_hm$pt_wet)
-sd_TE_hm_d <- sd(df_TE_hm$pt_wet)
+TE_hm_w <- mean(df_TE_hm$pt_wet)  # mean TEhm when wet
+sd_TE_hm_d <- sd(df_TE_hm$pt_wet)  # sd TEhm when wet
 
-# Calculating the TE_hm as a function of time 
-# dryingtime= 10
+# Creating a function to estimate the TEhm as a function of time after inoculation (t)
+dt = 10
+sinNum = 100
+t = 2
 
-# if t <= dryingtime, m(t) = t/dryingtime 
-# if t > dryingtime, m(t)=1
+## Question, can t be a variable with uncertainty? 
 
-# TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
-# sd_TE_hm(t)=
+f_TEhm <- function (t, simNum, dt) {
+    TEhm <- rnorm(simNum, , )
+}
+
+
+ if t <= dryingtime, m(t) = t/dryingtime 
+ if t > dryingtime, m(t)=1
+
+ TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
+ sd_TE_hm(t)=
 
     
 # output -------------------------------------------------------------------

### 2020-05-24 More info to add.Rmd
New datasets and equations

---
title: "More info to add"
author: "Ana Karina"
date: "5/4/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## More info for the paper

#### More info that could be added

GC:Infectivity
@Yan2018 significant correlation between virus RNA copies and infectious virus recovered from breathe and swabs of influenza patients,

WHO recomends:
washing hands often with soap and water
cleaning frequently touched surfaces and objects.

Data on the number of CoV-2 viruses emitted while coughing, speaking, breathing, or sneasing need to be better understaood.

"the main mode of transmission is through close exposure to an infected person who coughs or exhales droplets in your presence." NY times

“The good news is that the novel coronavirus comes from a family that can't take the heat. Coronaviruses in general are enveloped in a coat of fat and protein that tends to lose its shape at high temperatures, a process likened to melting that effectively disables the virus. They also tend to survive longest in conditions of low humidity."

"quantitative measurement of infectious respiratory viruses in the air are needed to evaluate the infectiousness of respiratory viruses over the aerosol route"

"Droplets smaller than 10 μm can remain suspended for many minutes"

unmasked and are coughing, talking or just came in from a jog and are breathing heavily.

Find a table of Inactivation of coronaviruses by different types of biocidal agents in carrier tests in article --> @Kampf2020

modeling-based approaches have a potential role to play in reducing the impact of the ongoing COVID-19 outbreak, and QMRA parameters obtained from previous studies on relevant respiratory viruses help to inform risk assessments of SARS-CoV-2.

@WHO2020b
"Recommendation to Member States to improve hand hygiene practices widely to help prevent the transmission of the COVID-19 virus by:
1. Providing universal access to public hand hygiene stations and making their use obligatory on entering and leaving any public or private commercial building and any public transport facility.
2. Improving access to hand hygiene facilities and practices in health care facilities."

Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.

Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.

"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020

Part of the WHO COVID-19 strategy is to "engaging and educating the public to ensure protective measures are implemented effectively as changes are made to existing policies and restrictions." @WHO2020a

He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case. More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"

Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.

Air samples negative, but surface samples positive @Ong2020.

Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015. 
__________

Ratio.
For flu virus, "There is a weak, but significant, correlation between virus RNA copies and infectious virus." As shown by Yan et al. RNA copies and infectious flu virus recovered of symptomatic influenza patients. Infectious virus and RNA copies (r = 0.34 and 0.58, spearman correlation for breath samples and swab samples). Ratio estimated using Figure 2.C. (swab samples, r=0.58). It looks like at least a 1:1000 correlation, if not higher. @Yan2018

Effectiveness of hand-washing
A systematic review of 67 studies conducted to assess the effectiveness of differnet physical interventions to interrupt or reduce the spread of acute respiratory viruses showed that hygienic measures such as handwashing can reduce the spread of respiratory virus @Jefferson2011.

___________

"The literature suggests the volume of saliva expelled per cough or sneeze could be different
(Duguid, 1946; Loudon & Roberts, 1966; Nicas & Jones, 2009; Papineni & Rosenthal, 1997). If a
mean volume (VSaliva) for coughs was calculated by multiplying the average droplet diameter, by
the average number of droplets expelled per cough with respect to the droplet size distribution, a
total saliva volume expelled per cough of 0.0065 ml would result. However, a higher mean volume
of saliva expelled per cough was reported to be 0.044 ml (Nicas & Jones, 2009) so, the later larger
volume for coughs was used to produce the most conservative estimates. For the volume of saliva
expelled during a sneeze, a volume of 0.44 ml was used. This value was also previously used in
the Nicas and Jones model (Nicas & Jones, 2009). The combination of three exposure times and
two volumes of saliva expelled for coughing or sneezing events led to six different simulations.
The six possible scenarios evaluated can be described as below:
 (1) and (2): An instantaneous exposure with one cough or one sneeze, corresponding to an
expelled volume of 0.044 or 0.44 ml (Nicas & Jones, 2009), followed by a direct surface and face
touch (base scenario);
 (3) and (4): A 15 min exposure duration resulting in 3 coughs of 0.044 ml each or 3 sneezes of
0.44 ml each, 15 surface touches and 3.75 face touches;
 (5) and (6): A 30 min exposure period with 6 coughs of 0.044 mL each or 6 sneezes of 0.44 ml
each, 30 surface touches and 3.75 face touches; "

### 2020-05-24 Paper.Rmd
New datasets and equations

@@
 ---
-title: "Paper"
+title: "Draft 1"
 author: "Ana Karina"
@@
 # Risk of infection from touching surfaces contaminated with SARS-CoV-2
-
 ## Abstract
-
 ## Introduction
 
+extensive environmental contamination has been found in settings with COVID-19 patients @Ong2020 @Ye2020, coronaviruses persist on inanimate surface for extended periods of time @Kampf2020
+
 If no contrameasures are taken, viruses can spread thorugh the space populated by the human host at a high speed, as we have observed @BarOn2020.
+
+Surfaces play a significant role in the spread of respiratory viral diseases @Boone2007
+
+Potential role of aerosol transmission of SARS-CoV-2, sampling the air @Liu2020
 
@@
 
+Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002.
+People touch their faces frequently (REF).
 Therefore, touching contaminated surfaces offers a risk for exposure.
@@
 ### Hazard Identification
-This risk assessment is designed to explore the risks associated indirect transmission of SARS-CoV-2.
-
-Air samples negative, but surface samples positive @Ong2020.
-Human coronaviruses can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020.
+When an individual infected with SARS-CoV-2 coughs, sneezes, or speaks, there could be particles of saliva containing the virus that make their way to the proximal surfaces.
+SARS-CoV-2 can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020.
+Viruses are known to transfer from surfaces to the skin upon contact  @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991.
+It is also well known that people frequently touch their faces and the mucouse membranes within the face (nose, eyes and mouth) @Kwok2015.
+Therefore, indirect transmission of SARS-CoV-2 a relevant transmission pathway.
+Despite its relevance, there are no studies quantifying the risks associated with indirect transmission of SARS-CoV-2 between surfaces and people.
+Hence, this risk assessment is designed to explore the risks associated indirect (fomite mediated) transmission of SARS-CoV-2.
+Since the probability of illness due to infection is, up to date, not known (REF), the end point of this risk assessment is probability of infection.
 
 ### Exposure Assessment 
-Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.
-
 Two expossure scenarios were modeled. 
 *Scenario 1. Single touch.*
-Frequently touched buttons (Traffic light buttons, train buttons). Supermarket products.
+Frequently touched buttons (Traffic light buttons, train buttons). High touched surfaces and Supermarket products.
 
@@
 
+#### Parameters table
+| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
+|:--:|:--:|:--:|:--:|:--:|:--:|
+| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020, RT-PCR data from 9, 2, and 2 patients with COVID-19|
+| GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
+| V~slv/cgh~ | mL | Volume of saliva expelled per cough | |  | |
+| N~cgh/min~ |min^-1^ | Number of caughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV| 
+|t~i~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
+|x~i~| m | Inoculation distance for activity *i* | [...] | Point values | Assumed |
+| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246(0.00098) | Normal | http://qmrawiki.org/experiments/sars |
+| n | min~-1~ |Decay rate for surface *j*| [...] | Normal | @VanDoremalen2020, CoV-2 infectivity assays, 40% RH and 21-23°C |
+| |  ||  | | |
+
 #### SARS-CoV-2 Concentration in saliva samples
 
-*Conversing viral RNA data to infectious viruses*
-There was a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
-Therefore, we had to translate the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
+There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
+Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
 Althoug not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF).
 To do this we ...
-
-*Viral shedding and infection potential* 
-Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
-The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF).
 In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
 
-### Dose Response for SARS-CoV
+#### SARS-CoV-2 Concentration in surfaces
+
+\begin{equation}
+C(0)=\frac {V_{slv/cgh} C_{sp}}{4 \pi x_{i}^{2}  0.2}
+\end{equation}
+
+\begin{equation}
+C(t)= C(0)  e^{-nt}
+\end{equation}
+
+#### Dose Response for SARS-CoV
 “For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”
@@
 No human dose-response relationship is available for SARS Coronavirus.
-It is generally accepted to use animal models in QMRA if human data is not available (REF)
+It is generally accepted to use animal models in QMRA if human data is not available @Haas2014, @Haas2015.
 
@@
 
+Seasonal factors such as temperature, humidity and sunlight are known to influence the spread of respiratory viruses such as influenza (REF).
+The role that those factos have on the spread of SARS-CoV-2 remains unknown (REF), althouh it is know that the temperature and humidity influences the viability of the virus (REF), as well as SARS-CoV-2 survive longer at lower temperature (REF).
+
+It has been shown that seasonal environmental factors such as temperature and humidity influence respiratory virus survival and transmission @Moriyama2020.
+
+Humidity influences the size of the droplets generated by the infected individual, the size of the droplets influence both the lifetime of the virus in the environment. 
+The deposition of droplets onto surfaces depends on the particle size.
+
+Nevertheless, our results suggest that fomite mediated transmission will be ..
+influenced by those factors.
+The extend that this influence will have in the droplet dynanimcs by influencing evaporation rate of the droplets (REF).
+transmission of CoV will depend on the relative contribution of indirect transmission pathway.
+
 ### Limitations
+
+Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
+The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF).
+In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
+
 Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.
@@
 
+Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used.
+However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines.
+
 ## Supporting Information
+
 ## Author Information
+
 ## Aknowledgements
+
 ## References
-
-#### More info that could be added
-
-Large droplets (>= 5um) that fall rapidly near the source contaminate proxime surfaces, making indirect human-surface transmission.
-
-Fecal-oral route has been suggested as a potential transmission route for SARS-CoV-2 @Yeo2020.
-“Toilet bowl and sink samples were positive, suggesting that viral shedding in stool” @Ong2020.
-
-"“aerosol and fomite transmission of SARS-CoV-2 is plausible, since the virus can remain viable and infectious in aerosols for hours and on surfaces up to days (depending on the inoculum shed).” @VanDoremalen2020
-
-Part of the WHO COVID-19 strategy is to "engaging and educating the public to ensure protective measures are implemented effectively as changes are made to existing policies and restrictions." @WHO2020a
-
-He et al. analized the patterns of viral sheddin of 94 COVID-19 patients. They also suggested substantial presymptomatic transmission @He2020. "we have estimated that viral shedding of patients
-with laboratory-confirmed COVID-19 peaked on or before symptom onset, and a substantial proportion of transmission probably occurred before first symptoms in the index case.
-More inclusive criteria for contact tracing to capture potential transmission events 2 to 3 days before symptom onset should be urgently considered for effective control of the outbreak"
-
-Matrices: saliva, mucus, nasal secretions, lower respiratory tack secretions, urine, blood.

### 2020-05-24 src/analyse_risk .R

@@ -1,14 +1,13 @@
-
-
 # setup -------------------------------------------------------------------
 # external libraries
 library(tidyverse) 
+library(truncnorm)
 
 # read and process data  --------------------------------------------------
-source ("src/process_concentrationsaliva.R") # Wölfel 2020 and Pan 2020
-#source ("src/process_TEhandtosaliva.R") # Pitol 2017
-source ("src/process_TEsurfacetohand.R") # Lopez 2013
-source ("src/process_ATM_observation.R") # Observation ATM
+source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020
+source ("src/process_TEhandtosaliva.R") # Pitol 2017 (dry transfer from hand to saliva)
+source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity?)
+source ("src/process_ATM_observation.R") # Observation ATM 
 source ("src/process_decay.R") # van Doremalen 2020
 
 ##    Scenario 1 : Single touch, ATM machine 
@@ -16,37 +15,34 @@ source ("src/process_decay.R") # van Doremalen 2020
 # Simulation parameters
 simNum = 100 #number of simulations
 
-# Constant parameters
-GC_TCID50 <- 0.01 # Proportion of infective gene copies [GC:TCID50]
-cough_ang <- pi/2 # 
-t <- 2 # Time since inoculation 
-t_k <- 5 # Time doing the activity
-x <- 4 # Inoculation distance
-
-# Dose-Response Parameter (exponential): 
-# (0.00246 (0.00135; 0.00459)),5th, 50th, and 95th percentile values, Huang (2013), wikiQMRA, 
-
 # Parameters from distributions and datasets
+GC_TCID50 <- runif(simNum, 100, 1000) # Proportion of infective gene copies [GC:TCID50],  
 Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
-Ncough_min <- rnorm(simNum, 0.57, 1) # Number of coughs per minute, Leung (2020)
-Csp <- sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Concentration of virus in sputum [GC/ml] 
[…]

### 2020-05-24 src/process_ATM_observation.R

@@ -4,9 +4,19 @@
 infile = "data/raw/ATM_observation.csv"
 #outfile = "data/processed/XX.csv"
 
+# Library
+library(MASS) 
+
 # read data 
 df_ATM <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
+t_ATM_mean <- mean(df_ATM$t_ATM)
+t_ATM_sd <-  sd(df_ATM$t_ATM)    
 
+# ATM visits per hour
+visits <- df_ATM$visits_hr
+visits <- visits[!is.na(visits)]
+visits_mean <- mean(visits)
+visits_sd <- sd(visits)
 
 # output -------------------------------------------------------------------
 

### 2020-05-24 src/process_TEhandtosaliva.R

@@ -1,5 +1,4 @@
-
-# Processing data of virus transfer from hands to saliva
+# Processing data of virus transfer from hands to saliva, based on MS2
 # Data taken from Pitol, 2017. Available at <https://zenodo.org/record/1454505#.Xo3wMZNKjBI>
 
 # setup -------------------------------------------------------------------
@@ -8,8 +7,7 @@
 
 # read and process data  --------------------------------------------------
 infile = "data/raw/TEhandtosaliva.csv"
-#outfile = "data/processed/TEhandtosaliva.csv"
-#function TEhm
+# outfile = "data/processed/TEhandtosaliva.csv"
 
 # read data 
 df_TE_hm <- read.csv (file=infile, header = TRUE, sep=",")
@@ -21,35 +19,19 @@ df_TE_hm <- read.csv (file=infile, header = TRUE, sep=",")
 # TE_hm_w [%] = transfer efficiency from hand to mucous membrane (wet)
 # TE_hm [%]= transfer efficiency from hand to mucous membrane (dry and wet)
 
-TE_hm_d <- mean(df_TE_hm$pt_dry) # mean TEhm when dry
-sd_TE_hm_d <- sd(df_TE_hm$pt_dry) # sd TEhm when dry
+TE_hm_d <- mean(df_TE_hm$pt_dry)/100 # mean TEhm when inoculum is dry
+sd_TE_hm_d <- sd(df_TE_hm$pt_dry)/100 # sd TEhm when inoculim is dry
 
-TE_hm_w <- mean(df_TE_hm$pt_wet)  # mean TEhm when wet
-sd_TE_hm_d <- sd(df_TE_hm$pt_wet)  # sd TEhm when wet
+TE_hm_w <- mean(df_TE_hm$pt_wet)/100  # mean TEhm when inoculum is wet
+sd_TE_hm_w <- sd(df_TE_hm$pt_wet)/100  # sd TEhm when inoculum is wet
 
 # Creating a function to estimate the TEhm as a function of time after inoculation (t)
-dt = 10
-sinNum = 100
-t = 2
-
-## Question, can t be a variable with uncertainty? 
-
-f_TEhm <- function (t, simNum, dt) {
-    TEhm <- rnorm(simNum, , )
-}
+# Function normal with limits
 
 
- if t <= dryingtime, m(t) = t/dryingtime 
- if t > dryingtime, m(t)=1
-
- TE_hm(t) = m(t) * TE_hm_d + (1-m (t)) * TE_hm_w
[…]

### 2020-05-24 src/process_TEsurfacetohand.R

@@ -11,8 +11,18 @@ outfile = "data/processed/TEsurfacetohand.csv"
 
 # read data 
 df_TE_sh <- read.csv (file=infile, header = TRUE, sep=",")
+df_TE_sh$pt_mean <- df_TE_sh$pt_mean / 100
+df_TE_sh$pt_sd <- df_TE_sh$pt_sd / 100
+df_TE_sh$pt_min <- df_TE_sh$pt_min / 100
+df_TE_sh$pt_max <- df_TE_sh$pt_max / 100
 
+# Extracting data for steel at high humidity (40-65)
+TE_sh_stl_mean <- df_TE_sh[14,3]
+TE_sh_stl_sd <- df_TE_sh[14,4]
 
+# Extracting data for plastic at high humidity (40-65)
+TE_sh_pl_mean <- df_TE_sh[10,3]
+TE_sh_pl_sd <- df_TE_sh[10,4]
 
 # output -------------------------------------------------------------------
 

### 2020-05-24 src/process_concentrationsaliva.R

@@ -1,14 +1,17 @@
-# Processing data of concentration of SARS-CoV-2 samples in sputum (taken as saliva)
+# Processing data of concentration of SARS-CoV-2 samples in sputum (taken as saliva), and samples
+# taken from the uper respiratory track (urt). 
+
 # Data obtained from: 
     #"Viral load of SARS-CoV-2 in clinical samples" (Pan, 2020) 
     #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020)
+    #"Viral load kinetics of SARS-CoV-2 infection in first two patients in Korea" (Kim, 2020)
 
 # setup -------------------------------------------------------------------
 # external libraries
 
 # read and process data  --------------------------------------------------
 infile = "data/raw/concentrationsaliva.csv"
-#outfile = "data/processed/decay.csv"
+# outfile = "data/processed/decay.csv"
 
 # read data 
 df_Csp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml

### 2020-05-24 src/process_decay.R

@@ -1,23 +1,50 @@
-
 # Processing data decay rate of SARS-CoV-2 on different surfaces
 
-# Data obtained from the Suplementary Appendix of the article 
-
+# Data obtained from the Suplementary Appendix of:
+# van Doremalen 2020 "Aerosol and Surface Stability of SARS-CoV-2 as Compared with SARS-CoV-1"
 
 # setup -------------------------------------------------------------------
 # external libraries
 
 # read and process data  --------------------------------------------------
 infile = "data/raw/decay.csv"
-#outfile = "data/processed/decay.csv"
+# outfile = "data/processed/decay.csv"
 
 # read data 
 df_decay <- read.csv (file=infile, header = TRUE, sep=",")
 
-# Equations for first order decay
-# C(t) = C(0) e^ (-kt)
-# halflife = ln2 / k
-# k = ln2 / halflife
+# Converting hrs to min
+df_decay$halflife <- df_decay$halflife * 60
+df_decay$lowCI <- df_decay$lowCI * 60
+df_decay$highCI <- df_decay$highCI * 60
+
+# Parameters for steel
+steel <- subset(df_decay, material == "steel", select=c(halflife, lowCI, highCI))
+k_stl_mean <- steel$halflife
+k_stl_97.5 <- steel$highCI
+k_stl_2.5  <- steel$lowCI   
+k_stl_sd  <- round((k_stl_97.5 - k_stl_2.5)/(2 *qnorm(0.975)), 5)
+
+# Parameters for plastic
+plastic <- subset(df_decay, material == "plastic", select=c(halflife, lowCI, highCI))
+k_pl_mean <- plastic$halflife
+k_pl_97.5 <- plastic$highCI
+k_pl_2.5  <- plastic$lowCI   
+k_pl_sd  <- round((k_pl_97.5 - k_pl_2.5)/(2 *qnorm(0.975)), 5)
+
+# Parameters for cooper
+copper <- subset(df_decay, material == "copper", select=c(halflife, lowCI, highCI))
+k_cu_mean <- copper$halflife
+k_cu_97.5 <- copper$highCI
+k_cu_2.5  <- copper$lowCI   
+k_cu_sd  <- round((k_cu_97.5 - k_cu_2.5)/(2 *qnorm(0.975)), 5)
+
+# Parameters for cardboard
+cardboard <- subset(df_decay, material == "cardboard", select=c(halflife, lowCI, highCI))
[…]

### 2020-05-24 src/process_doseresponse.R

@@ -1,0 +1,13 @@
+
+# Processing the data from the Dose Response paramaters
+
+# Data obtained from QMRAwiki http://qmrawiki.org/experiments/sars
+# Assuming estimate = median = mean (normal distribution)
+Q_50 <- 0.00246 
+Q_5 <-  0.00135   
+Q_95 <- 0.00459
+
+# Estimating SD form percentiles 
+k_sd <- round((Q_95 - Q_5)/(2 *qnorm(0.95)), 5)
+k_mean <- Q_5
+

### 2020-05-24 src/analyse_risk .R
correcting equations

@@ -20,8 +20,8 @@ GC_TCID50 <- runif(simNum, 100, 1000) # Proportion of infective gene copies [GC:
 Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
 Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # Number of coughs per minute, Leung (2020)
 TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # Transfer Efficiency from hand to saliva, Pitol 2017
-
 FSAsf <- runif(simNum, 0.5, 1) # Fraction of the finger in contact with the ATM --> correct this
+FSAfm <- runif(simNum, 0.5, 1) # Fraction of the finger in contact with the mouth --> correct this
 Csp <- 10^sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Needs correction! Replace 10^0 with 0? Change distribution? Wölfel 2020, Pan 2020, Kim 2020
 
 # Activity specific parameters
@@ -32,20 +32,22 @@ Csp <- 10^sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Needs correction! R
 t_ATM <-  rtruncnorm(simNum, 0, Inf, t_ATM_mean, t_ATM_sd)  # Time at ATM, 20 hours of observational studies
 t_btw_ATM <- rtruncnorm(simNum, 0, Inf, 60/visits_mean, 60/visits_sd) # (Change distribution at the end) Time between ATM visits, observational studies
 x <- 0.4 # Inoculation distance (m), assumed based on observations
-k_stl <- rnorm(simNum, k_stl_mean, k_stl_sd) # Half life of CoV-2 in steel
+k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 in steel
 n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface
 TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
+k <- rnorm(simNum, k_mean, k_sd)
 
[…]

### 2020-05-24 src/process_ATM_observation.R

@@ -15,7 +15,7 @@ t_ATM_sd <-  sd(df_ATM$t_ATM)
 # ATM visits per hour
 visits <- df_ATM$visits_hr
 visits <- visits[!is.na(visits)]
-visits_mean <- mean(visits)
+visits_mean <- mean(visits) # Change the distribution once I have all data
 visits_sd <- sd(visits)
 
 # output -------------------------------------------------------------------

### 2020-05-24 src/analyse_risk .R
...

@@ -43,7 +43,7 @@ df <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, FSAsf, FSAfm, TEhm, t_
 # Calcuated parameters
 for (i in 1:simNum)
 {
-    df$Cs_0[i] <- (df$Csp [i] / df$Vs[i]) / ((4 *pi* x^2)/0.1) #correct with 10% of the area of sphere 
+    df$Cs_0[i] <- ((df$Csp [i]/GC_TCID50[i]) / df$Vs[i]) / ((4 *pi* x^2)/0.1) #correct with 10% of the area of sphere 
     df$Cs_1[i] <- df$Cs_0[i] * exp(- n_stl[i] * t_btw_ATM[i])
     df$Nf[i]   <- df$Cs_1[i] * df$FSAsf [i] * df$TEsh_stl[i]
     df$Dose[i] <- df$Nf[i] * df$FSAfm [i] * df$TEhm [i]
@@ -59,9 +59,14 @@ c_Csp <- cor.test(df$P_inf, df$Csp, method = "spearman", exact=F)
 c_Vs <- cor.test(df$P_inf, df$Vs, method = "spearman", exact=F)   
 c_t_btw_ATM <- cor.test(df$P_inf, df$t_btw_ATM, method = "spearman", exact=F)  
 c_n <- cor.test(df$P_inf, df$n_stl, method = "spearman", exact=F)  
+c_TEsh <- cor.test(df$P_inf, df$TEsh_stl, method = "spearman", exact=F)  
+c_GC_inf <- cor.test(df$P_inf, df$GC_TCID50, method = "spearman", exact=F)  
 
-corRes <- data.frame(type = c("TEhm", "Csp", "Vs", "t_btw_ATM","n" ),
-                     rho = c(c_TEhm$estimate, c_Csp$estimate, c_Vs$estimate, c_t_btw_ATM$estimate, c_n$estimate))
+
+
+corRes <- data.frame(type = c("TEhm", "Csp", "Vs", "t_btw_ATM","n", "TEsh", "GC_inf"),
+                     rho = c(c_TEhm$estimate, c_Csp$estimate, c_Vs$estimate, c_t_btw_ATM$estimate, c_n$estimate, 
+                             c_TEsh$estimate, c_GC_inf$estimate))
 
 ggplot(data = corRes, aes(x = type, y = rho)) + geom_bar(stat = "identity") 
 + theme_minimal() + xlab("") + theme(axis.text.x = element_text(angle = 10))

### 2020-05-28 More info to add.Rmd
Latest changes

@@
 
-GC:Infectivity
-@Yan2018 significant correlation between virus RNA copies and infectious virus recovered from breathe and swabs of influenza patients,
+The subjects touched their eyes, nostrils, and lops an average of 16(11) times an hour @Nicas2008
 
@@
 
+__________
+
+Often, parameters such as virus transfer are not available for the virus in question, therefore surrogates are used
+
 ___________

### 2020-05-28 Paper.Rmd
Latest changes

@@
 ## Introduction
+close contact route (large droplet), airborne (small droplets), fomites.
+An investigation of a nosocomial outbreak of COVID-19 reported at Augustine’s Hospital in Durban, South Africa suggest that indirect contact through health care workers and fomites were the predominant transmission pathways @Lessells2020.
+
+*Risk assessment models*
+Mathematical models have been used to understand SARS-CoV and MERS-CoV @Adhikari2019 @Lei2017.
+Let et al @Lei2017  developed a model to understand the relative contribution of different transmission routes (airborne, close contact, and fomites) on the risk of infection for influenza A H1N1, SARS-CoV, and norovirus.
+The simulation results suggest that around 50% of the CoV infections were due to indirect (fomite mediated) transmission.
+A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus.
+They only considered the risk of infection due from droplets and aerosolized particles, since that is the main transmission route.
+Most of the uncertainty associated with the model was due to the concentration of virus in the saliva.
+The researchers evaluated possible interventions and showed that the use of mask contributed significantly to the reduction of the risk (>90%), as well as increasing air exchange (up to 58% risk reduction).
 
@@
 
-If no contrameasures are taken, viruses can spread thorugh the space populated by the human host at a high speed, as we have observed @BarOn2020.
+If no countermeasures are taken, viruses can spread through the space populated by the human host at a high speed, as we have observed @BarOn2020.
 
@@
 
-The CDC and WHO have highligthed two main transmission routes: direct transmission or person-to-person transmission (respiratory droplets produced while talking, caughin, or sneezing), and indirect transmission or surface mediated transmission (touching a contaminated surface followed by touching mucose membrains in the face such as mouth or nose) @WHO2020 @CDC2020.
+The CDC and WHO have highlighted two main transmission routes: direct transmission or person-to-person transmission (respiratory droplets produced while talking, coughing, or sneezing), and indirect transmission or surface-mediated transmission (touching a contaminated surface followed by touching mucose membranes in the face such as mouth or nose) @WHO2020 @CDC2020.
 
-SARS-CoV-1 transmission has been associated with nosocomial spread.
-Transmission data suggests that, although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004.
-A study of an outbreak of SARS CoV in Tawian showed that, out of the 232 SARS patients in the hospital, 31 patients (13 %)  were hospital acquired transmissions.
+Although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004.
+A study of an outbreak of SARS CoV in Taiwan showed that, out of the 232 SARS-CoV patients in the hospital, 31 patients (13 %)  were hospital-acquired transmissions.
 Out of the 31 cases, most of the transmission occurred by direct contact with a known SARS patient, but 6/31 (~20%) of the transmissions could not be traced to direct contact with SARS patients.
@@
 Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002.
-People touch their faces frequently (REF).
-Therefore, touching contaminated surfaces offers a risk for exposure.
-Inspite its importance, there is not a lot of information regarding the magnitud of risk associeted with touching contaminated surfaces @BarOn2020.
+People touch their faces frequently @Nicas2008.
+Therefore, touching contaminated surfaces offers a risk of exposure.
+Despite its importance, there is not a lot of information regarding the magnitude of risk associated with touching contaminated surfaces @BarOn2020.
 
-Hand hygiene interventions and surface desinfection have been hihgligthed as one of the primary strategies to control transmission @Mcclellan2020.
+Hand hygiene interventions and surface disinfection have been highlighted as one of the primary strategies to control transmission @Mcclellan2020.
 Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020.
-But there is not conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
-
-A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus.
-They only considered the risk of infection due from droplets and aer(osolized particles, since that is the main transmission route.
-Most of the uncertainty associated with the model was due to the concentration of virus in the saliva.
-The researchers evaluated possible interventions and showed that the use of mask contributed significantly to the reduction of the risk (>90%), as well as increasing air exchange (up to 58% risk reduction).
+But there is no conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
 
@@
 |:--:|:--:|:--:|:--:|:--:|:--:|
-| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020, RT-PCR data from 9, 2, and 2 patients with COVID-19|
+| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020 @To2020, RT-PCR data from 9, 2, 2, and 23 patients with COVID-19|
 | GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
@@
 #### SARS-CoV-2 Concentration in saliva samples
-
 There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
 Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
-Althoug not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF).
-To do this we ...
-In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
+We used data on the viral loads of the saliva or sputum of an infected person within the first 14 days of symptom onset @Wolfel2020 @Pan2020 @Kim2020 @To2020.
+
+Althoug translating copy numbers into infectivity data is not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF).
+
+In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016.
+Both, influenza viruses and coronaviruses are enveloped, RNA viruses with similar sizes.
+
+*Conversion factor RT-PCR data to infectivity*
+Although the ratio of infectivity to gene copy numbers is, to our knowledge, available for coronavirus, there is information of this ratio for the samples taken from patients for other respiratory viruses such as different strains of influenza virus @Yan2018 @Ip2016.
+
+GC:Infectivity
+@Yan2018 significant correlation between virus RNA copies and infectious virus recovered from breathe and swabs of influenza patients,
 
 #### SARS-CoV-2 Concentration in surfaces
+
+"viru transferred between two surfaces is often formulated as the product of contact frequency, transfer efficiency, surface concentration, and contact area"
 
@@
 #### Dose Response for SARS-CoV
+
+To "To estimate the infection risk based on the dose received,"
 “For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”
@@
 
+_____
+about the paper, we can say something like "Although the relative importance of each transmission route remains poorly understood, there has been a growing consensus that contaminated fomites play a critical role in the spread of viruses." @Boone2007
+
+The relative contribution of surface-mediated transmission as compared with other transmission routes such as direct transmission, air, or thorugh the feca-oral route remains to be addressed.
+QMRA can be used to incorporate different transmission routes and answer this question.
+
 ## Supporting Information

### 2020-05-28 src/analyse_risk .R

@@ -36,6 +36,7 @@ k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 i
 n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface
 TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
 k <- rnorm(simNum, k_mean, k_sd)
+angle <- 0.5 #fix, based in the images of @2Bourouiba2014
 
 # Creating data fram with all the variables and numbers simulated from distributions ATM   
 df <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, FSAsf, FSAfm, TEhm, t_ATM, t_btw_ATM, x, n_stl, TEsh_stl))

### 2020-05-28 src/process_concentrationsaliva.R

@@ -5,6 +5,8 @@
     #"Viral load of SARS-CoV-2 in clinical samples" (Pan, 2020) 
     #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020)
     #"Viral load kinetics of SARS-CoV-2 infection in first two patients in Korea" (Kim, 2020)
+    #"Temporal profiles of viral load in posterior oropharyngeal saliva samples and serum antibody 
+        # responses during infection by SARS-CoV-2: an observational cohort study" (To, 2020)
 
 # setup -------------------------------------------------------------------
 # external libraries
@@ -15,7 +17,8 @@ infile = "data/raw/concentrationsaliva.csv"
 
 # read data 
 df_Csp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
-
+df_Csp <- subset(df_Csp, day_corr<15)
+df_Csp <- na.omit(df_Csp)
 
 # output -------------------------------------------------------------------
 

### 2020-06-04 More info to add.Rmd
Correction of Finger surface area, addition of new information.

@@
 
+______________
+Regarding the volume of saliva expelled per cough. 
+Papers:
+
+@Chabrelie2018 : Table1. Volume of saliva expelled per cough or sneeze (0.044 or 0.44 [mL]). Point value. Taken from multiple sources.
+
+@Nicas2005: "Original article of particle size and volume, but no aparent 0.044 value in there" This is the one that Nicas2009 cites when it comes to past calculation.
+It also says the following "Duguid reported that a sneeze emits about 200- fold more particles than a cough."
+
+@Adhikari2019: 0.044 value plus minus 10%, uniform, citing Nicas 2009.
+
+@Nicas2009: "The number and size distri- bution of cough particles adhere to values presented in a previous analysis, and an estimated 0.044 mL of fluid (saliva) is emitted per cough." citing Nicas2005
+
 ___________

### 2020-06-04 Paper.Rmd
Correction of Finger surface area, addition of new information.

@@
 | GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
-| V~slv/cgh~ | mL | Volume of saliva expelled per cough | |  | |
+| V~slv/cgh~ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | @Nicas2009, volume = 0.044, @Adhikari2019, assumed uniform $\pm$ 10%|
 | N~cgh/min~ |min^-1^ | Number of caughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV| 
-|t~i~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
-|x~i~| m | Inoculation distance for activity *i* | [...] | Point values | Assumed |
-| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246(0.00098) | Normal | http://qmrawiki.org/experiments/sars |
-| n | min~-1~ |Decay rate for surface *j*| [...] | Normal | @VanDoremalen2020, CoV-2 infectivity assays, 40% RH and 21-23°C |
-| |  ||  | | |
+|t~ATM~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
+|x~ATM~| m | Inoculation distance for activity *i* | [...] | Point value |  This study. Estimated based on activity ATM: 20 hours of observational data |
+| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246, 0.00135, 0.00459 | Triangular | Data obtained from QMRA wiki http://qmrawiki.org/experiments/sars, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 5th, 50th, and 95th percentiles as min, mode, and max |
+| n~stl~ | min^-1^ |Decay rate in steel | [...] | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+| n~pl~ | min^-1^ |Decay rate in plastic | [...] | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+| TE~sh_stl~ | % | Transfer efficiency from surface (steel) to hand | Normal | 37.4 (16) | @Lopez2013, transfer efficiency of viruses (MS2) from surfaces to skin at 40-65% humidity | 
+| TE~sh_pl~ | % | Transfer efficiency from surface (plastic) to hand | Normal | 79.5 (21.2) | @Lopez2013, transfer efficiency of viruses (MS2) from acrylic to skin at 40-65% humidity | 
+| TE~hm~ | - | Transfer efficiency from hand to mucouse membrane | Normal | 0.201(0.063) | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
+| A~sf~ | cm^2^ | Finger surface area |Uniform | 3.8-6.6 | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
+| FSA~fm~ | - | Fractional surface area | Uniform | 0.5-0.8 | Assumed |
 
@@
 
+Although not perfect, a risk assessment framework could be use to conduct a hazzard assessment. Since offices will need to re-open 
+https://www.osha.gov/shpguidelines/hazard-Identification.html.
+
+A more comprehensive model such as the models described elsewhere (REF), could potentially help elucidating the relative contribution of different expossure routes, and inform the population regading the most sensitive measures they could take given the particularities of the situation (e.g., air conditionated in offices)
+
 ## Supporting Information

### 2020-06-04 src/analyse_risk_2.R

@@ -27,16 +27,19 @@ FSAfm <- runif(simNum, 0.5, 0.8) # Assumed
 ##- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
 # Activity specific parameters
 
+#---------------------------------------------------------------------------------------------------
+#----------------------------- Cough on hand -------------------------------------------------------
+
 #                           ---- ATM------   Material : steel
 
 # The TEsh and inactivation on surface were steel specific.
 t_ATM <-  rtruncnorm(simNum, 0, Inf, t_ATM_mean, t_ATM_sd)  # Time at ATM, 20 hours of observational studies
 t_btw_ATM <- rtruncnorm(simNum, 0, Inf, 60/visits_mean, 60/visits_sd) # Time between ATM visits, observational studies
-x <- 40 # Inoculation distance (cm), assumed based on observations
+x <- 5 # Inoculation distance (cm), mouth to hand
 k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 in steel
 n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface
 TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
-angle <- 0.1 #fix, based in the images of @2Bourouiba2014
+angle <- 0.9 # assumed
 
 # Creating data fram with all the variables and numbers simulated from distributions ATM   
 df_risk_ATM <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, Asf, FSAfm, TEhm, t_ATM, t_btw_ATM, n_stl, TEsh_stl))
@@ -44,7 +47,8 @@ df_risk_ATM <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, Asf, FSAfm, T
 # Calcuated parameters
 for (i in 1:simNum)
 {
[…]

### 2020-06-04 src/process_ATM_observation.R

@@ -1,4 +1,6 @@
 # Processing data obtained from the ATM observation study
+        # 20 hours of observation divided in morning, afternoon and evening 
+        # 2 different ATMs
 
 # read and process data  --------------------------------------------------
 infile = "data/raw/ATM_observation.csv"
@@ -8,15 +10,15 @@ infile = "data/raw/ATM_observation.csv"
 library(MASS) 
 
 # read data 
-df_ATM <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
+df_ATM <- read.csv (file=infile, header = TRUE, sep=",") 
+
+# Time spent in the ATM (normally distributed)
 t_ATM_mean <- mean(df_ATM$t_ATM)
 t_ATM_sd <-  sd(df_ATM$t_ATM)    
 
-# ATM visits per hour
+# Number of ATM visits per hour (not normally distributed, Re-Sample)
 visits <- df_ATM$visits_hr
 visits <- visits[!is.na(visits)]
-visits_mean <- mean(visits) # Change the distribution once I have all data
-visits_sd <- sd(visits)
 
 # output -------------------------------------------------------------------
 

### 2020-06-04 src/process_Asf.R

@@ -1,0 +1,41 @@
+# Processing finger area taken from different papers 
+
+#  ------------------------------Method 1 ----------------------------------
+# Method used by Julian et. all 2018. @Julian2018
+# Fractional surface area taken from @AuYeung2008 (adult front partial finger)
+# Average hand surface area taken from EPA 2011 @EPA2011
+
+FA_min <- 0.04
+FA_max <- 0.06    
+
+# Hand surface area (average of the area of 2 hands, EPA 2011 page xiv), 1000 <- m2 to cm2
+HSA_women <- (0.089 * 10000)/2
+HSA_men <-  (0.107  * 10000) /2
+
+HSA <- mean(HSA_women:HSA_men) # average men/women area (cm2) 
+
+# Area of finger (uniform using min and max of front partial finger)
+#Asf_min <- HSA * FA_min
+#Asf_max <- HSA * FA_max
+
+#  ------------------------------Method 2 ----------------------------------
+#   Data from the paper @Chabrelie2018, cm2. He used a combination of data from elsewhere
+Asf_median <- 4.58 #@Chabrelie2018
+
+Asf_murai <- 6.66 # volar area of the index finger, 7 fingers from cadavers, @Murai1996
+Asf_peters <- (3.55 + 4.25)/2  # Index finger distal phalanx female and male, @Peters2009a
+Asf_sahmel <- 3.8 # Six participants, 3 fingers each, @Sahmel2015
+
+
+Asf_min <- Asf_sahmel
+Asd_max <- Asf_murai
+
+#  ------------------------------Method 3 ----------------------------------
+
+
+
+
+# output -------------------------------------------------------------------
+
+# save file
+# write.csv (df_XXX, file= outfile)

### 2020-06-04 src/process_TEhandtosaliva.R

@@ -26,8 +26,6 @@ TE_hm_w <- mean(df_TE_hm$pt_wet)/100  # mean TEhm when inoculum is wet
 sd_TE_hm_w <- sd(df_TE_hm$pt_wet)/100  # sd TEhm when inoculum is wet
 
 # Creating a function to estimate the TEhm as a function of time after inoculation (t)
-# Function normal with limits
-
 
 # output -------------------------------------------------------------------
 

### 2020-06-04 src/process_concentrationsaliva.R

@@ -5,20 +5,23 @@
     #"Viral load of SARS-CoV-2 in clinical samples" (Pan, 2020) 
     #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020)
     #"Viral load kinetics of SARS-CoV-2 infection in first two patients in Korea" (Kim, 2020)
-    #"Temporal profiles of viral load in posterior oropharyngeal saliva samples and serum antibody 
-        # responses during infection by SARS-CoV-2: an observational cohort study" (To, 2020)
+    #"Temporal profiles of viral load in posterior oropharyngeal saliva samples and serum antibody responses during infection by SARS-CoV-2: an observational cohort study" (To, 2020)
 
 # setup -------------------------------------------------------------------
 # external libraries
 
 # read and process data  --------------------------------------------------
 infile = "data/raw/concentrationsaliva.csv"
-# outfile = "data/processed/decay.csv"
+
 
 # read data 
 df_Csp <- read.csv (file=infile, header = TRUE, sep=",") # C_sp is viral load in log10 copies/ml
 df_Csp <- subset(df_Csp, day_corr<15)
 df_Csp <- na.omit(df_Csp)
+
+df_Csp <- mutate(df_Csp, Csp = 10^ log10_C)
+df_Csp$Csp[df_Csp$Csp == 1] <- 0 # replacing 10^0 by 0
+
 
 # output -------------------------------------------------------------------
 

### 2020-06-04 src/process_decay.R

@@ -18,28 +18,28 @@ df_decay$halflife <- df_decay$halflife * 60
 df_decay$lowCI <- df_decay$lowCI * 60
 df_decay$highCI <- df_decay$highCI * 60
 
-# Parameters for steel
+# Parameters for steel, using median as mean, and assuming sd based on CI
 steel <- subset(df_decay, material == "steel", select=c(halflife, lowCI, highCI))
 k_stl_mean <- steel$halflife
 k_stl_97.5 <- steel$highCI
 k_stl_2.5  <- steel$lowCI   
 k_stl_sd  <- round((k_stl_97.5 - k_stl_2.5)/(2 *qnorm(0.975)), 5)
 
-# Parameters for plastic
+# Parameters for plastic, using median as mean, and assuming sd based on CI
 plastic <- subset(df_decay, material == "plastic", select=c(halflife, lowCI, highCI))
 k_pl_mean <- plastic$halflife
 k_pl_97.5 <- plastic$highCI
 k_pl_2.5  <- plastic$lowCI   
 k_pl_sd  <- round((k_pl_97.5 - k_pl_2.5)/(2 *qnorm(0.975)), 5)
 
-# Parameters for cooper
+# Parameters for cooper, using median as mean, and assuming sd based on CI
 copper <- subset(df_decay, material == "copper", select=c(halflife, lowCI, highCI))
 k_cu_mean <- copper$halflife
 k_cu_97.5 <- copper$highCI
 k_cu_2.5  <- copper$lowCI   
 k_cu_sd  <- round((k_cu_97.5 - k_cu_2.5)/(2 *qnorm(0.975)), 5)
 
-# Parameters for cardboard
+# Parameters for cardboard, using median as mean, and assuming sd based on CI
 cardboard <- subset(df_decay, material == "cardboard", select=c(halflife, lowCI, highCI))
 k_crdb_mean <- cardboard$halflife
 k_crdb_97.5 <- cardboard$highCI

### 2020-06-04 src/process_doseresponse.R

@@ -2,6 +2,10 @@
 # Processing the data from the Dose Response paramaters
 
 # Data obtained from QMRAwiki http://qmrawiki.org/experiments/sars
+    # Parameter developed from 2 different studies: 
+        # @
+        # @
+
 # Assuming estimate = median = mean (normal distribution)
 Q_50 <- 0.00246 
 Q_5 <-  0.00135   

### 2020-06-12 Paper.Rmd
Correction of parameters, and first draft of materials and methods

--- 2020-05-28 Paper.Rmd
@@
 ## Introduction
+
 close contact route (large droplet), airborne (small droplets), fomites.
@@
 Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002.
-People touch their faces frequently @Nicas2008.
+People touch their faces frequently @Nicas2008 @Kwok2015.
 Therefore, touching contaminated surfaces offers a risk of exposure.
@@
 
+Relevance of fomites in the context of SARS-CoV-2 (Stanford review) @Castano2020
+
 ## Materials and Methods 
 ### Hazard Identification
-When an individual infected with SARS-CoV-2 coughs, sneezes, or speaks, there could be particles of saliva containing the virus that make their way to the proximal surfaces.
-SARS-CoV-2 can persist at room temperature on surfaces like plastic, metal, or glass for up to 9 days @Kampf2020.
-Viruses are known to transfer from surfaces to the skin upon contact  @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991.
-It is also well known that people frequently touch their faces and the mucouse membranes within the face (nose, eyes and mouth) @Kwok2015.
-Therefore, indirect transmission of SARS-CoV-2 a relevant transmission pathway.
-Despite its relevance, there are no studies quantifying the risks associated with indirect transmission of SARS-CoV-2 between surfaces and people.
-Hence, this risk assessment is designed to explore the risks associated indirect (fomite mediated) transmission of SARS-CoV-2.
-Since the probability of illness due to infection is, up to date, not known (REF), the end point of this risk assessment is probability of infection.
+Whenever an infected individual coughs, sneezes or speaks, droplets containing pathogens make their way to the proximal surfaces [@Bourouiba2014].
+It has been shown that SARS-CoV-2 can persist on surfaces like plastic, metal, or glass at room temperature for hours to days [@VanDoremalen2020].
+Additionally, viruses are known to transfer from surfaces to the hands upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991], as well as from hands to the lips and saliva of individuals [@Rusin2002; @Pitol2017].
+Given the fact that people frequently touch their faces and the mucous membranes within the face (nose, eyes and mouth) [@Kwok2015; @Nicas2008], we have identified the indirect (surface-mediated) transfer of SARS-CoV-2 as a relevant transmission pathway.
+This risk assessment is designed to explore the risks of infection associated with the surface-mediated transmission of SARS-CoV-2.
 
 ### Exposure Assessment 
-Two expossure scenarios were modeled. 
-*Scenario 1. Single touch.*
-Frequently touched buttons (Traffic light buttons, train buttons). High touched surfaces and Supermarket products.
+Only one potential exposure pathway was analysed: the indirect transmission of viruses through contaminated surfaces.
+Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine).
 
-*Scenatio 2. Multiple touches.*
-ATM machines.
+*Scenario 1.
+Single touch buttons.* Frequently touched buttons such as traffic light buttons, train buttons, and elevator buttons are of risk concern due to their frequent use by multiple people.
+In this scenario, we modelled the risk of touching a contaminated button followed by touching mucous membranes on the face.
+The material of the button was assumed to be either plastic or steel.
+The sequence of events leading to the viral inoculation of the susceptible individual was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger, 4) virus transfer from the finger to the mucous membranes in the face (eyes, mouth, nose).
 
-(Insert diagram)
+*Scenario 2.
+Multiple touches.* ATM/cash machines are another example of surfaces that are frequently visited by individuals and are not frequently cleaned, as they are located in public spaces.
+For this risk assessment model, the ATM was assumed to be made of steel.
+The sequence of events leading to virus inoculation was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger and from finger to surface through multiple contacts, 4) virus transfer from the finger to the mucous membranes in the face (eyes, mouth, nose).
 
-#### Parameters table
-| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
-|:--:|:--:|:--:|:--:|:--:|:--:|
-| C~sp~ | Gene Copies/ml | Concentration of SARS-CoV-2 in sputum or saliva | log~10~ [4,4.8...] | ReSample | @Wolfel2020 @Pan2020 @Kim2020 @To2020, RT-PCR data from 9, 2, 2, and 23 patients with COVID-19|
-| GC:TCID~50~ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | @Ip2016, mean RT-PCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
-| V~slv/cgh~ | mL | Volume of saliva expelled per cough | |  | |
-| N~cgh/min~ |min^-1^ | Number of caughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV| 
-|t~i~ | min |Time spend doing the activity *i* | [...] | Normal | This study. Activity ATM: 20 hours of observational data |
-|x~i~| m | Inoculation distance for activity *i* | [...] | Point values | Assumed |
-| k | PFU^-1^ | Parameter of the exponential dose-response| 0.00246(0.00098) | Normal | http://qmrawiki.org/experiments/sars |
-| n | min~-1~ |Decay rate for surface *j*| [...] | Normal | @VanDoremalen2020, CoV-2 infectivity assays, 40% RH and 21-23°C |
-| |  ||  | | |
+####  Concentration and transfer of SARS-CoV-2
+Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input.
+The virus titers used in the risk assessment (Table 1) were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020].
+The RT-qPCR method estimates the number of copies of viral RNA in the clinical samples.
+Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses.
+In risk assessments where no data on the concentration of infectious viruses is available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses (REF).
+The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2.
+In the absence of SARS-CoV-2 data, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016 (Table 1).
+Both, influenza viruses and coronaviruses, are enveloped, RNA viruses with similar sizes.
 
-#### SARS-CoV-2 Concentration in saliva samples
-There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
-Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
-We used data on the viral loads of the saliva or sputum of an infected person within the first 14 days of symptom onset @Wolfel2020 @Pan2020 @Kim2020 @To2020.
-
-Althoug translating copy numbers into infectivity data is not ideal, this is an approach often used in risk assessment models, given the complexity of obtaining infectivity data of a number of viral  pathogens (REF).
-
-In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for seasonal A(H1N1), influenza A(H3N2), and influenza B @Ip2016.
-Both, influenza viruses and coronaviruses are enveloped, RNA viruses with similar sizes.
-
-*Conversion factor RT-PCR data to infectivity*
-Although the ratio of infectivity to gene copy numbers is, to our knowledge, available for coronavirus, there is information of this ratio for the samples taken from patients for other respiratory viruses such as different strains of influenza virus @Yan2018 @Ip2016.
-
-GC:Infectivity
-@Yan2018 significant correlation between virus RNA copies and infectious virus recovered from breathe and swabs of influenza patients,
-
-#### SARS-CoV-2 Concentration in surfaces
-
-"viru transferred between two surfaces is often formulated as the product of contact frequency, transfer efficiency, surface concentration, and contact area"
+The concentration of viruses in the surface after inoculation (through cough) was estimated using the following equation:
 
 \begin{equation}
-C(0)=\frac {V_{slv/cgh} C_{sp}}{4 \pi x_{i}^{2}  0.2}
+C(0)=\frac { C_{sp}V_{s}}{\frac {4 \pi x_{i}^{2}}{a}}
 \end{equation}
+
+where $C(0)$ [virus/cm^2^] is the concentration of virus in the surface, $V_{s}$ [mL/cough] is the volume of saliva expelled per cough, $C_{sp}$ [virus/mL] is the concentration of infective viruses in the saliva or sputum, $x_i$[cm] is the inoculation distance or the distance between the surface and the infected individual at the time of the cough for activity $i$, and $a$ is the XXX.
+This equation assumes that the droplets are expelled in a conical shape and a proportion of the spheric part of the cone is in contact with the surface.
+After surface inoculation, the concentration of the virus on the surface at time = $t$ was estimated using the following equation:
 
@@
 
-#### Dose Response for SARS-CoV
-
-To "To estimate the infection risk based on the dose received,"
-“For translating this animal dose–response relationship to a human dose– response relationship, a generally accepted assumption that a death end point for an animal model may be used for examining the human risk of infection was applied (Haas, Rose, & Gerba, 2014)”
-
-The dose-response relationship was obtain from QMRA wiki <http://qmrawiki.org/experiments/sars>.
-The model is based on animal dose-response relationship.
-No human dose-response relationship is available for SARS Coronavirus.
-It is generally accepted to use animal models in QMRA if human data is not available @Haas2014, @Haas2015.
+where $C(t)$[virus/cm^2^] is the concentration of infective viruses at time $t$, $C(0)$[virus/cm^2^] is the initial concentration of virus in the surface, $t$[min] is the time lapsed after surface inoculation, and $n$[m^{-1}^] is the decay rate of the virus in the selected surface (metal or plastic).
+The number of viruses transferred from the surface to the finger $N_f$ [virus] was estimated using the following equation:
 
 \begin{equation}
-P_{inf}=1 - e^{-kd}
+N_f = C(t)  TE_{sh} A_f
 \end{equation}
+
+where $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands, $C(t)$ [virus/cm^2^] is the concentration of virus in the surface and $A_f$[cm^2^] is the area of the finger.
+For multiple touch scenarios the number of virus in the finger multiple contacts was calculated as follows:
+
+The dose of viruses that entered the susceptible individual through facial membranes, $D$ [virus], was estimated as follows:
+
+\begin{equation}
+D = N_f FSA_{fm} TE_{hm}
+\end{equation}
+
+where $N_f$[virus] is the number of virus in the finger, $FSA_{fm}$ [%] is the fractional surface area between the finger and the mucous membranes, and $TE_{hm}$ [%] is the transfer efficiency of the virus from the hand to the mucous membranes.
+
+#### Dose-response
+In the case of highly virulent pathogens, where no prophylaxis is available, it is common practice to use data of animal trials to generate dose-response models.
+The dose-response relationship for SARS was obtained from the QMRA wiki <http://qmrawiki.org/experiments/sars>.
+This model is based on two studies [@DeAlbuquerque2006; @DeDiego2008] of SARS-CoV and Murine hepatitis virus (MHV-1) infection in mice.
+MHV is a commonly used surrogate for human coronaviruses.
+
+The risk of infection was estimated using the following exponential dose-response model:
+
+\begin{equation}
+P_{inf}=1 - e^{-kD}
+\end{equation}
+
+where $P_{inf}$ [%] is the probability of infection, $k$ [-] is the infectivity parameter of the exponential model and $D$ [virus] is the dose of virus that entered the individual through mucous membranes.
+
+#### Model parameters used in the risk assessment 
+| Parameter | Units | Description | Input values / Equation | Distribution | Reference and comments |  
+|:--:|:--:|:--:|:--:|:--:|:--:|
+| $C_{spGC}$ | gene copies/mL | Concentration of SARS-CoV-2 in sputum or saliva | Log~10~ gene copies/mL (data set) | ReSample | [@Wolfel2020; @Pan2020; @Kim2020; @To2020] RT-qPCR data from 9, 2, 2, and 23 patients with COVID-19|
+| $GC:TCID_{50}$ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | [@Ip2016] Mean RT-qPCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
+| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
+| $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
+|$t_{ATM}$ | min |Time spent on the ATM | 0.75(0.38) | Normal | This study. Otained from 20 hours of observational data |
+|$t_{btw-ATM}$ | min |Time between visits | 8.4(7.1) | Normal | This study. Otained from 20 hours of observational data |
+|$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
+|$x_{btn}$| cm | Button inoculation distance | 50 | Point value | Assumed |
+|$t_{0.5-stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+|$t_{0.5-pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $\frac {ln(2)}{t_{0.5}}$ | - | calculated |
+| $TE_{sh-stl}$ | % | Transfer efficiency of virus between steel to hand | 37.4(16) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
+| $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
+| $TE_{hm}$ | - | Transfer efficiency from hand to mucouse membranes | 0.201(0.063) | Normal | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
+| $A_{f}$ | cm^2^ | Finger surface area | 3.8-6.6 | Uniform | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
+| $FSA_{fm}$ | - | Fractional surface area between finger and mucous membranes | 0.5-0.8 | Uniform | Assumed | 
+| $k$ | PFU^-1^ | Parameter of the exponential dose-response| 0.00107, 0.00135, 0.00680 | Triangular | Data obtained from QMRAwiki, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 0.5th, 50th, and 99.5th percentiles as min, mode, and max |
+|:--:|:--:|:--:|:--:|:--:|:--:|
+| N~cgh/min~ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV|
 
@@
 
-( Insert table of parameters )
-
 ## Results and Discussion
@@
 ```
+
+### Factors influencing fomite mediate virus risk
+There are number of factors influencing indirect risk of SARS-CoV-2 transmission @Moriyama2020, some of these factors are highly correlated with each other.
+Seasonal and environmental factors such as temperature, humidity and sunlight are examples of them.
+Relative humidity (RH) influences the size of the droplets expelled in the cough, virus survival on surfaces @Casanova2010, and virus transfer between hands and surfaces @Lopez2013.
+Lopez et al showed that XXX.
+This risk assessment was performed using the transfer efficiency of viruses at a relative humidity of (X-X).
+If we were to exchange this variable for the percentage transfer of a low relative humidity setting we will have a lower inactivation, as the lower the RH, the higher the survival @Casanova2010, but we would have a significantly higher transfer of viruses between surfaces and skin (i.e., XX ).
+Overall, this change would predict a XX% decrease in the indirect risk of transmission with a dicrease in RH.
+Temperature and sunlight are known to correlate negatively to virus survival in surfaces @Chin2020, @Casanova2010.
+In addition to environmental factors, behavioural factors such as hygiene practices and social behaviour are known to influence fomite mediated transmission (REF).
+
+One important parameter not considered in the model is the susceptible individual immune system and how the environmental factors influence it.
+The immune system plays an important role in the probability of infection @Moriyama2020.
+
+Surfaces: SARS-CoV-2 is more stable in smooth surfaces @Chin2020.
+
+Surfaces are easy to clean.
+SARS-CoV-2 is susceptible to disintectants including household belach and ethanol @Chin2020 @Castano2020.
+Although disinfection can be achieved easily using a variaty of disinfectants, public surfaces that are frequently contacted but not frequently cleanned present a challenge.
+Strategies to minimize the contact with surfaces such as trafic light buttons could therefore be beneficial to reduce the risk of expossure.
+
+### Sensitivity analysis
+The factors that influence the most the risk of ilness where the same in all the scenations: 1) concentration of virus in the saliva, 2) ..., 3)...
+
+“A dX change in input results in a dY change in outcome”
 
@@
 
+*Dose Response* 
+The use of animals models in QMRA is a generally accepted @Haas2014, @Haas2015. 
+It has been shown that different viruses of the family of coronaviruses have significantly different dose-response relationships, the difference in the ID~50~ or dose require to observe a 50% response can be orders of magnitude different beween different coroaviruses [@Watanabe2010].
+Therefore, a dose-response relationship developed specifically for SARS-CoV-2 would increase the accuracy of the results of the present and future risk assessments.
+The dose response relationship used herein predicts a 50% response when the dose is 282 PFU.
+*xx*
+
+*Virus inoculation* 
+More detailed information regarding the number and size of particles expelled per couugh could be used to increase accuracy of risk assessments @Lee2019.
+This QMRA had big assumptions regarding the inoculation of viruses on the surfaces.
+Regardless of the assumptions, we expect the results not to change significantly with an increase in accurace of virus inoculation calculation, as we don't expect it to be wrong more than one order of magnitude, as compared with the variation expected in the concentration of viruses on the saliva for symptomatic patients within the first two weeks of symptom onsent, which goes from 0 to ~10^X gene copies/mL.
+*xx*
+
+*Virus concentration on infected individuals*
+
+There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
+Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
+
 Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
@@
 In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
+Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used.
+However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines.
+*xx*
+
+*Transmission routes*
+We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countributionof each route. 
+*x*
 
@@
 
-We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countribution of each route.
-
+*Conclusion*
 Even though this risk assessment model relies on many assumptions, some with high uncertainty around them, and the risk caucluation is hihgly simplidied, this model can: provide insigts into the possible effects that different interventions can have in reducing the spread of SARS-CoV-2, serve as a first step in the pathway of building more complex models that allow exploring the relative contribution that different transmission routes could have in spreddin the disease.
 
-Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used.
-However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines.
-
-_____
 about the paper, we can say something like "Although the relative importance of each transmission route remains poorly understood, there has been a growing consensus that contaminated fomites play a critical role in the spread of viruses." @Boone2007
@@
 QMRA can be used to incorporate different transmission routes and answer this question.
+
+QMRA can be used to assess the significance of surface mediated transmission of SARS-CoV-2 [@Haas2020], as it has been used in the past to assess the risk of other transmission routes for MERS-CoV [@Adhikari2019].
+Although not perfect, a risk assessment framework could be use to conduct a hazzard assessment.
+Since offices will need to re-open https://www.osha.gov/shpguidelines/hazard-Identification.html.
+
+Oportunities for managing the risks associated with people's interaction with frequently touched surfaces include contact reduction at a population level, for example reduce the need for contacting trafic ligth buttons by setting them into automatic mode.
+Since surface cleanning of public spaces is not feasible with the frequency needed to significanlty reduce risks, signs that remind people to clean their hands after contact could potentially be encoraged (REF).
 

### 2020-06-12 src/Risk.R

@@ -1,69 +1,0 @@
-## Set-up
-require(tidyverse)
-
-## Parameters Used
-
-# Reading the data frames 
-source ("src/process_concentrationsaliva.R") # Wölfel 2020 and Pan 2020
-source ("src/process_TEhandtosaliva.R") # Pitol 2017
-source ("src/process_TEsurfacetohand.R") # Lopez 2013
-source ("src/process_ATM_observation.R") # Observation ATM
-source ("src/process_decay.R") # van Doremalen 2020
-
-# Simulation parameters
-simNum = 100 #number of simulations
-
-# Constant parameters
-GC_TCID50 <- 0.01 # Proportion of infective gene copies [GC:TCID50]
-cough_ang <- pi/2 # 
-t <- 2 # Time 
-x <- 4
-
-# Functions (relationships between variables)
-f_Cs0 <- function (Ncough_min, Csp, Vs) {
-    Cs_0 <- (Ncough_min * Csp * Vs)
-    return(Cs_0)
-}
-
-f_Cs1 <- function(Cs_0, n, t) {
-    Cs_1 <- (Cs_0 * exp(-nt))
-    return(Cs_1)
-}
-
-f_Nf <- function(Cs_1, SAf, TEsh) {
-    Cs_1 <- (Cs_0 * exp(-nt))
-    return(Cs_1)
-}
-
-f_TE_hm <- function(t, TE_hm_d, TE_hm_w, sd_TE_hm_d, sd_TE_hm_w) {
-    m <- t/dryingtime 
-    TE_hm_mean <- m * TE_hm_d + (1-m) * TE_hm_w
-    TE_hm_sd <-  m * sd_TE_hm_d + (1-m) * sd_TE_hm_w # Check how to compute mean of sd
-    return(TE_hm_mean, TE_hm_sd)
-}
-
-
-# Parameters from distributions and datasets
-Vs <- runif(simNum, 0.0396, 0.0484) # Volume of saliva per cough, Nicas and Jones (2009)
-Ncough_min <- rnorm(simNum, 0.57, 1) # Number of coughs per minute, 
-Csp <- sample(df_Csp$C_sp, size=simNum, replace = TRUE) # Concentration of virus in sputum [GC/ml] 
[…]

### 2020-06-12 src/analyse_risk .R

@@ -16,10 +16,10 @@ simNum = 100 #number of simulations
 
 # Parameters from distributions and datasets commun to all the scenarios
 GC_TCID50 <- runif(simNum, 100, 1000) # (copies/TCID50)Proportion of infective gene copies [GC:TCID50], Ip (2015) 
-Vs <- runif(simNum, 0.0396, 0.0484) # (ml) Volume of saliva per cough, Nicas and Jones (2009)
-Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # (cough/min) Number of coughs per minute, Leung (2020)
+Vs <- runif(simNum, 0.0396, 0.0484) # (mL) Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
+#Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # (cough/min) Number of coughs per minute, Leung (2020)
 TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # Transfer Efficiency from hand to saliva, Pitol 2017
-k <-  rtri(simNum, min = 0.00135, max = 0.00459 , mode = 0.00246) # Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
+k <-  rtri(simNum, min = Q_0.5, max = Q_99.5 , mode = Q_50) # Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
 Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # (copies/ml) Concentration of saliva in the first 2 weeks after symptom onsent 
 Asf <- runif(simNum, Asf_min, Asf_max) # (cm^2) @Chabrelie2018 @Murai1996 @Peters2009a @Sahmel 2015
 FSAfm <- runif(simNum, 0.5, 0.8) # Assumed

### 2020-06-12 src/process_ATM_observation.R

@@ -19,7 +19,8 @@ t_ATM_sd <-  sd(df_ATM$t_ATM)
 # Number of ATM visits per hour (not normally distributed, Re-Sample)
 visits <- df_ATM$visits_hr
 visits <- visits[!is.na(visits)]
-
+visits_mean <- mean(visits)
+visits_sd <- sd(visits)
 # output -------------------------------------------------------------------
 
 # save file

### 2020-06-12 src/process_TEsurfacetohand.R

@@ -24,7 +24,25 @@ TE_sh_stl_sd <- df_TE_sh[14,4]
 TE_sh_pl_mean <- df_TE_sh[10,3]
 TE_sh_pl_sd <- df_TE_sh[10,4]
 
-# output -------------------------------------------------------------------
+#  -------------------------------------------------------------------
 
-# save file
-# write.csv (df_XXX, file= outfile)
+# read and process data  --------------------------------------------------
+#infile = "data/raw/TEsurhandtosurface.csv"
+
+# read data 
+#df_TE_hs <- read.csv (file=infile, header = TRUE, sep=",")
+#df_TE_hs$pt_mean <- df_TE_hs$pt_mean / 100
+#df_TE_hs$pt_sd <- df_TE_hs$pt_sd / 100
+#df_TE_hs$pt_min <- df_TE_hs$pt_min / 100
+#df_TE_hs$pt_max <- df_TE_hs$pt_max / 100
+
+# Extracting data for steel at high humidity (40-65)
+#TE_hs_stl_mean <- df_TE_hs[X,X]
+#TE_hs_stl_sd <- df_TE_hs[X,X]
+
+# Extracting data for plastic at high humidity (40-65)
+#TE_hs_pl_mean <- df_TE_hs[X,X]
+#TE_hs_pl_sd <- df_TE_hs[X,X]
+
+
+

### 2020-06-12 src/process_doseresponse.R

@@ -2,15 +2,24 @@
 # Processing the data from the Dose Response paramaters
 
 # Data obtained from QMRAwiki http://qmrawiki.org/experiments/sars
-    # Parameter developed from 2 different studies: 
-        # @
-        # @
+  # Parameter developed from 2 different studies using 2 different coronaviruses: SARS-CoV and Murine hepatitis virus (MHV-1)
+    # @DeAlbuquerque2006. A/J mice with MHV-1 virus, intranasal, survival monitored for 21 days. 
+            # Note: MHV is a commun surrogate of human coronaviruses
+    # @DeDiego2008. 4 groups of the transgenic mice, intranasal, graded doses of rSARS-CoV, survival monitored for 13 days
+  # @Watanabe2010. Combining both dose-response articles to develop the dose-response
 
-# Assuming estimate = median = mean (normal distribution)
-Q_50 <- 0.00246 
-Q_5 <-  0.00135   
-Q_95 <- 0.00459
+# Option 1, using the from 2.5th, 50th, and 97.5th percentiles as min, mode, and max
 
+# Parameter (k) for the exponential model, based on the 2 articles, using 10,000 bootstrap iterations (MLE)
+Q_0.5  <- 0.00107
+Q_2.5  <- 0.00128
+Q_5    <- 0.00135 
+Q_50   <- 0.00246 
+Q_95   <- 0.00459
+Q_97.5 <- 0.00527
+Q_99.5 <- 0.00680
+
+# Option 2, using normal distribution, calculating the SD based on 9
 # Estimating SD form percentiles 
 k_sd <- round((Q_95 - Q_5)/(2 *qnorm(0.95)), 5)
 k_mean <- Q_5

### 2020-06-12 src/risk_cough_to_hand.R

@@ -40,6 +40,7 @@ k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 i
 n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface
 TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
 angle <- 0.9 # assumed
+TEhs_stl <- rtruncnorm(simNum, 0, 1, TE_hs_stl_mean, TE_hs_stl_sd) # TE from hand to surface for stainless steel RH=[40-65%]
 
 # Creating data fram with all the variables and numbers simulated from distributions ATM   
 df_risk_ATM <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, Asf, FSAfm, TEhm, t_ATM, t_btw_ATM, n_stl, TEsh_stl))
@@ -48,7 +49,7 @@ df_risk_ATM <- as.data.frame(cbind(GC_TCID50, Vs, Ncough_min, Csp, Asf, FSAfm, T
 for (i in 1:simNum)
 {
     df_risk_ATM$Ch_0[i]  <- ((((df_risk_ATM$Csp [i]/df_risk_ATM$GC_TCID50[i])) * df_risk_ATM$Vs[i])) / ((4 *pi* x^2)/angle)
-    df_risk_ATM$Cs_0[i]  <- 
+    df_risk_ATM$Cs_0[i]  <-  
     df_risk_ATM$Cs_1[i]  <- df_risk_ATM$Cs_0[i] * exp(- df_risk_ATM$n_stl[i] * df_risk_ATM$t_btw_ATM[i])
     df_risk_ATM$Nf[i]    <- df_risk_ATM$Cs_1[i] * df_risk_ATM$Asf [i] * df_risk_ATM$TEsh_stl[i]
     df_risk_ATM$Dose[i]  <- df_risk_ATM$Nf[i] * df_risk_ATM$FSAfm [i] * df_risk_ATM$TEhm [i]

### 2020-06-12 Paper.Rmd
Adding Paper V2

@@
 Given the fact that people frequently touch their faces and the mucous membranes within the face (nose, eyes and mouth) [@Kwok2015; @Nicas2008], we have identified the indirect (surface-mediated) transfer of SARS-CoV-2 as a relevant transmission pathway.
-This risk assessment is designed to explore the risks of infection associated with the surface-mediated transmission of SARS-CoV-2.
+Therefore, this risk assessment is designed to explore the risks of infection associated with the surface-mediated transmission of SARS-CoV-2.
 
@@
 The material of the button was assumed to be either plastic or steel.
-The sequence of events leading to the viral inoculation of the susceptible individual was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger, 4) virus transfer from the finger to the mucous membranes in the face (eyes, mouth, nose).
+The sequence of events leading to the viral inoculation of the susceptible individual was considered to be the following: 1) contamination event (infected person coughing directly on the surface), 2) virus decay on the surface, 3) virus transfer from the surface to the finger, 4) virus transfer from the finger to the mucous membranes  of the nose, mouth and eyes.
 
@@
 Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input.
-The virus titers used in the risk assessment (Table 1) were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020].
-The RT-qPCR method estimates the number of copies of viral RNA in the clinical samples.
+The virus titers used in the risk assessment were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020].
+The RT-qPCR method estimates the number of copies of viral RNA present in the samples.
 Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses.
-In risk assessments where no data on the concentration of infectious viruses is available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses (REF).
+In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses (REF).
 The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2.
@@
 
-where $C(0)$ [virus/cm^2^] is the concentration of virus in the surface, $V_{s}$ [mL/cough] is the volume of saliva expelled per cough, $C_{sp}$ [virus/mL] is the concentration of infective viruses in the saliva or sputum, $x_i$[cm] is the inoculation distance or the distance between the surface and the infected individual at the time of the cough for activity $i$, and $a$ is the XXX.
+where $C(0)$ [virus/cm^2^] is the concentration of virus in the surface, $V_{s}$ [mL/cough] is the volume of saliva expelled per cough, $C_{sp}$ [virus/mL] is the concentration of infective viruses in the saliva or sputum, $x_i$ [cm] is the inoculation distance or the distance between the surface and the infected individual at the time of the cough for activity $i$, and $a$ is the XXX.
 This equation assumes that the droplets are expelled in a conical shape and a proportion of the spheric part of the cone is in contact with the surface.
@@
 
-where $C(t)$[virus/cm^2^] is the concentration of infective viruses at time $t$, $C(0)$[virus/cm^2^] is the initial concentration of virus in the surface, $t$[min] is the time lapsed after surface inoculation, and $n$[m^{-1}^] is the decay rate of the virus in the selected surface (metal or plastic).
+where $C(t)$ [virus/cm^2^] is the concentration of infective viruses at time $t$, $C(0)$ [virus/cm^2^] is the initial concentration of virus in the surface, $t$[min] is the time lapsed after surface inoculation, and $n$ [m^-1^] is the decay rate of the virus in the selected surface (metal or plastic).
 The number of viruses transferred from the surface to the finger $N_f$ [virus] was estimated using the following equation:
@@
 
-where $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands, $C(t)$ [virus/cm^2^] is the concentration of virus in the surface and $A_f$[cm^2^] is the area of the finger.
+where $C(t)$ [virus/cm^2^] is the concentration of virus in the surface, $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands, and $A_f$[cm^2^] is the area of the finger.
 For multiple touch scenarios the number of virus in the finger multiple contacts was calculated as follows:
@@
 In the case of highly virulent pathogens, where no prophylaxis is available, it is common practice to use data of animal trials to generate dose-response models.
-The dose-response relationship for SARS was obtained from the QMRA wiki <http://qmrawiki.org/experiments/sars>.
-This model is based on two studies [@DeAlbuquerque2006; @DeDiego2008] of SARS-CoV and Murine hepatitis virus (MHV-1) infection in mice.
+The dose-response relationship used in this model was obtained from the QMRA wiki <http://qmrawiki.org/experiments/sars> and is based on two studies [@DeAlbuquerque2006; @DeDiego2008] of SARS-CoV and Murine hepatitis virus (MHV-1) infection in mice.
 MHV is a commonly used surrogate for human coronaviruses.
-
-The risk of infection was estimated using the following exponential dose-response model:
+The risk of infection was estimated with the following equation:
 
@@
 | $GC:TCID_{50}$ | - | Genome copies to infectious virus conversion factor | 10^2^-10^3^ | Uniform | [@Ip2016] Mean RT-qPCR to TCID~50~ ratio of 708, 547, and 185 gene copies per TCID~50~ for seasonal A(H1N1), A(H3N2), and influenza B, respectively |
-| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
+| $C_{sp}$ | TCID~50~/mL |Concentration of infective SARS-CoV-2 in sample | $C_{sp}=\frac {C_{spGC}}{GC:TCID_{50}}$ | - | calculated|
 | $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
@@
 |$t_{btw-ATM}$ | min |Time between visits | 8.4(7.1) | Normal | This study. Otained from 20 hours of observational data |
+|$t_{btw-btn}$ | min |Time between pressing button | 0.5-20 | Uniform | Assumed |
 |$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
 |$x_{btn}$| cm | Button inoculation distance | 50 | Point value | Assumed |
-|$t_{0.5-stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
-|$t_{0.5-pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
-| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $\frac {ln(2)}{t_{0.5}}$ | - | calculated |
+|$t_{\frac {1}{2}stl}$| min | Half life of CoV-2 in steel | 338(35) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+|$t_{\frac {1}{2}pl}$| min | Half life of CoV-2 in plastic | 409(39) | Normal | @VanDoremalen2020, SARS-CoV-2 infectivity assays, 40% RH and 21-23°C |
+| $n_x$| min^-1^ |Decay rate in surface $x$ (steel or plastic) | $n_x=\frac {ln(2)}{t_\frac{1}{2}}$ | - | calculated |
 | $TE_{sh-stl}$ | % | Transfer efficiency of virus between steel to hand | 37.4(16) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
 | $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
-| $TE_{hm}$ | - | Transfer efficiency from hand to mucouse membranes | 0.201(0.063) | Normal | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
+| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
 | $A_{f}$ | cm^2^ | Finger surface area | 3.8-6.6 | Uniform | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
@@
 *Dose Response* 
-The use of animals models in QMRA is a generally accepted @Haas2014, @Haas2015. 
-It has been shown that different viruses of the family of coronaviruses have significantly different dose-response relationships, the difference in the ID~50~ or dose require to observe a 50% response can be orders of magnitude different beween different coroaviruses [@Watanabe2010].
+One of the biggest limitations of this risk assessment is the uncertainty associated with the representativeness of experimental data used to derivate the dose-response model.
+Although animals models are generally accepted to develop dose-response curvers @Haas2014, @Haas2015, tt has been shown that different viruses of the family of coronaviruses have significantly different dose-response relationships, the difference in the ID~50~ or dose require to observe a 50% response can be orders of magnitude different beween different coroaviruses [@Watanabe2010].
 Therefore, a dose-response relationship developed specifically for SARS-CoV-2 would increase the accuracy of the results of the present and future risk assessments.

### 2020-06-12 Paper_V2.Rmd
Adding Paper V2

--- 2020-06-12 Paper.Rmd
@@
 ---
-title: "Draft 1"
+title: "Draft 2"
 author: "Ana Karina"
@@
 ## Introduction
-
-close contact route (large droplet), airborne (small droplets), fomites.
-An investigation of a nosocomial outbreak of COVID-19 reported at Augustine’s Hospital in Durban, South Africa suggest that indirect contact through health care workers and fomites were the predominant transmission pathways @Lessells2020.
-
-*Risk assessment models*
-Mathematical models have been used to understand SARS-CoV and MERS-CoV @Adhikari2019 @Lei2017.
-Let et al @Lei2017  developed a model to understand the relative contribution of different transmission routes (airborne, close contact, and fomites) on the risk of infection for influenza A H1N1, SARS-CoV, and norovirus.
-The simulation results suggest that around 50% of the CoV infections were due to indirect (fomite mediated) transmission.
-A QMRA of MERS-CoV was developed by Adhikari et all @Adhikari2019 to evaluate the risk for hospital transmission of coronavirus.
-They only considered the risk of infection due from droplets and aerosolized particles, since that is the main transmission route.
-Most of the uncertainty associated with the model was due to the concentration of virus in the saliva.
-The researchers evaluated possible interventions and showed that the use of mask contributed significantly to the reduction of the risk (>90%), as well as increasing air exchange (up to 58% risk reduction).
-
-extensive environmental contamination has been found in settings with COVID-19 patients @Ong2020 @Ye2020, coronaviruses persist on inanimate surface for extended periods of time @Kampf2020
-
-If no countermeasures are taken, viruses can spread through the space populated by the human host at a high speed, as we have observed @BarOn2020.
-
-Surfaces play a significant role in the spread of respiratory viral diseases @Boone2007
-
-Potential role of aerosol transmission of SARS-CoV-2, sampling the air @Liu2020
-
-The CDC and WHO have highlighted two main transmission routes: direct transmission or person-to-person transmission (respiratory droplets produced while talking, coughing, or sneezing), and indirect transmission or surface-mediated transmission (touching a contaminated surface followed by touching mucose membranes in the face such as mouth or nose) @WHO2020 @CDC2020.
-
-Although indirect transmission through fomite mediated contacts is not the primary transmission route, it can lead to infection @Chen2004.
-A study of an outbreak of SARS CoV in Taiwan showed that, out of the 232 SARS-CoV patients in the hospital, 31 patients (13 %)  were hospital-acquired transmissions.
-Out of the 31 cases, most of the transmission occurred by direct contact with a known SARS patient, but 6/31 (~20%) of the transmissions could not be traced to direct contact with SARS patients.
-Sampling of surfaces for the RNA of coronavirus showed that surfaces such as drinking water buttons, beds and chairs were contaminated with coronavirus.
-Thus, the researchers postulated that indirect contact could have been the source of transmission @Chen2004.
-
-Viruses are known to transfer from contaminated surfaces to the skin upon contact @Julian2010 @Lopez2013 @Bidawid2004 @Rusin2002 @Ansari1991 and from the skin to the mouth and saliva of individuals @Pitol2017 @Rusin2002.
-People touch their faces frequently @Nicas2008 @Kwok2015.
-Therefore, touching contaminated surfaces offers a risk of exposure.
-Despite its importance, there is not a lot of information regarding the magnitude of risk associated with touching contaminated surfaces @BarOn2020.
-
-Hand hygiene interventions and surface disinfection have been highlighted as one of the primary strategies to control transmission @Mcclellan2020.
-Studies with coronaviruses have shown than hygiene practices such as handwashing can reduce the risk of coronavirus infection @Beale2020.
-But there is no conclusive evidence of the effectiveness of hand-washing to reduce SARS-CoV-2 transmission @Yang2020.
-
-Relevance of fomites in the context of SARS-CoV-2 (Stanford review) @Castano2020
-
 ## Materials and Methods 
@@
 ### Exposure Assessment 
+#### Scenario definition
 Only one potential exposure pathway was analysed: the indirect transmission of viruses through contaminated surfaces.
@@
 
-####  Concentration and transfer of SARS-CoV-2
+####  Virus concentration and transfer 
 Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input.
@@
 where $C(t)$ [virus/cm^2^] is the concentration of infective viruses at time $t$, $C(0)$ [virus/cm^2^] is the initial concentration of virus in the surface, $t$[min] is the time lapsed after surface inoculation, and $n$ [m^-1^] is the decay rate of the virus in the selected surface (metal or plastic).
-The number of viruses transferred from the surface to the finger $N_f$ [virus] was estimated using the following equation:
+The concentration of virus on the hand after a single touch was estimated to be a function of the transfer efficiency of viruses between the hands and the surfaces.
+The concentration of viruses in the hand after multiple touches was estimated as reported elsewhere [@Julian2009; @Julian2018], where the transfer of virus is a function of the gradient in the concentration of viruses bwteen the hand and the surface.
 
 \begin{equation}
-N_f = C(t)  TE_{sh} A_f
+C_{f_f} = C(t) TE_{sh}
 \end{equation}
 
-where $C(t)$ [virus/cm^2^] is the concentration of virus in the surface, $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands, and $A_f$[cm^2^] is the area of the finger.
-For multiple touch scenarios the number of virus in the finger multiple contacts was calculated as follows:
+or
 
+\begin{equation}
+C_{f_f} = (C(t)-C_{f_i}) TE_{sh} + C_{f_i}
+\end{equation}
+
+where $C_{f_i}$ and $C_{f_f}$ [virus/cm^2^] are the initial and final concentration of virus in the finger, $C(t)$ [virus/cm^2^] is the concentration of virus in the surface at time = $t$, $TE_{sh}$[%] is the transfer efficiency of viruses between surfaces and hands.
 The dose of viruses that entered the susceptible individual through facial membranes, $D$ [virus], was estimated as follows:
@@
 \begin{equation}
-D = N_f FSA_{fm} TE_{hm}
+D = C_{f_f} S_{fm} TE_{hm}
 \end{equation}
 
-where $N_f$[virus] is the number of virus in the finger, $FSA_{fm}$ [%] is the fractional surface area between the finger and the mucous membranes, and $TE_{hm}$ [%] is the transfer efficiency of the virus from the hand to the mucous membranes.
+where $C_{finger_f}$ [virus/cm^2^] is the final concentration of virus in the finger, $S_{f}$ [cm] is the surface area of the finger in contact with the mucous membranes, and $TE_{hm}$ [%] is the transfer efficiency of the virus from the hand to the mucous membranes.
 
@@
 
-### Factors influencing fomite mediate virus risk
-There are number of factors influencing indirect risk of SARS-CoV-2 transmission @Moriyama2020, some of these factors are highly correlated with each other.
-Seasonal and environmental factors such as temperature, humidity and sunlight are examples of them.
-Relative humidity (RH) influences the size of the droplets expelled in the cough, virus survival on surfaces @Casanova2010, and virus transfer between hands and surfaces @Lopez2013.
-Lopez et al showed that XXX.
-This risk assessment was performed using the transfer efficiency of viruses at a relative humidity of (X-X).
-If we were to exchange this variable for the percentage transfer of a low relative humidity setting we will have a lower inactivation, as the lower the RH, the higher the survival @Casanova2010, but we would have a significantly higher transfer of viruses between surfaces and skin (i.e., XX ).
-Overall, this change would predict a XX% decrease in the indirect risk of transmission with a dicrease in RH.
-Temperature and sunlight are known to correlate negatively to virus survival in surfaces @Chin2020, @Casanova2010.
-In addition to environmental factors, behavioural factors such as hygiene practices and social behaviour are known to influence fomite mediated transmission (REF).
-
-One important parameter not considered in the model is the susceptible individual immune system and how the environmental factors influence it.
-The immune system plays an important role in the probability of infection @Moriyama2020.
-
-Surfaces: SARS-CoV-2 is more stable in smooth surfaces @Chin2020.
-
-Surfaces are easy to clean.
-SARS-CoV-2 is susceptible to disintectants including household belach and ethanol @Chin2020 @Castano2020.
-Although disinfection can be achieved easily using a variaty of disinfectants, public surfaces that are frequently contacted but not frequently cleanned present a challenge.
-Strategies to minimize the contact with surfaces such as trafic light buttons could therefore be beneficial to reduce the risk of expossure.
-
-### Sensitivity analysis
-The factors that influence the most the risk of ilness where the same in all the scenations: 1) concentration of virus in the saliva, 2) ..., 3)...
-
-“A dX change in input results in a dY change in outcome”
-
-Since no randomized control trials will be runned and there are limits to the conclusions we can reach using epidemiological data, risk assesments provide an attempt to understand the risk associated with different activities.
-The results of this paper could inform policy.
-For example, trafick lights in the crossing stations could clausured momentaniely, the same than train buttons.
-
-The dependence of temperature and humidity in the transfer rate of coronavirus is unclear.
-If temperature and humidity may play a role in non-dominant transmission pathways, such as indirect transfer, it would be hard to observe epidemiological studies.
-It is known that the survival of viruses on surfaces is dependent on environmental factors such as temperature and humidity (REF).
-Similarly, the transfer rate of viruses varies with humidity (REF).
-Nevertheless, if only a small fraction of the transmission occurs through indirect transfer we wont see it.
-
-The sensitivity analysis shows that XXX parapeters are responsible for XX % of the risk variation of the model. Collecting more data on these parameters could reduce uncertainty.
-
-It is important to notice that the effectiveness of interventions such as hand washing depends not only on many factors which include: the hand-wasihng technique, the suceptibility of the virus to phisical and chemical desinfection, and the relative contribution of the indirect transfer of viruses to the overal infection risk.
-As it has been pointed out, thand hygiene should be promoted but the risk of overemphasizing it without more intormation could turn out to be negative (Yang, 2020).
-...
-("risk compensation"?).
-
-Seasonal factors such as temperature, humidity and sunlight are known to influence the spread of respiratory viruses such as influenza (REF).
-The role that those factos have on the spread of SARS-CoV-2 remains unknown (REF), althouh it is know that the temperature and humidity influences the viability of the virus (REF), as well as SARS-CoV-2 survive longer at lower temperature (REF).
-
-It has been shown that seasonal environmental factors such as temperature and humidity influence respiratory virus survival and transmission @Moriyama2020.
-
-Humidity influences the size of the droplets generated by the infected individual, the size of the droplets influence both the lifetime of the virus in the environment. 
-The deposition of droplets onto surfaces depends on the particle size.
-
-Nevertheless, our results suggest that fomite mediated transmission will be ..
-influenced by those factors.
-The extend that this influence will have in the droplet dynanimcs by influencing evaporation rate of the droplets (REF).
-transmission of CoV will depend on the relative contribution of indirect transmission pathway.
-
-### Limitations
-
-*Dose Response* 
-One of the biggest limitations of this risk assessment is the uncertainty associated with the representativeness of experimental data used to derivate the dose-response model.
-Although animals models are generally accepted to develop dose-response curvers @Haas2014, @Haas2015, tt has been shown that different viruses of the family of coronaviruses have significantly different dose-response relationships, the difference in the ID~50~ or dose require to observe a 50% response can be orders of magnitude different beween different coroaviruses [@Watanabe2010].
-Therefore, a dose-response relationship developed specifically for SARS-CoV-2 would increase the accuracy of the results of the present and future risk assessments.
-The dose response relationship used herein predicts a 50% response when the dose is 282 PFU.
-*xx*
-
-*Virus inoculation* 
-More detailed information regarding the number and size of particles expelled per couugh could be used to increase accuracy of risk assessments @Lee2019.
-This QMRA had big assumptions regarding the inoculation of viruses on the surfaces.
-Regardless of the assumptions, we expect the results not to change significantly with an increase in accurace of virus inoculation calculation, as we don't expect it to be wrong more than one order of magnitude, as compared with the variation expected in the concentration of viruses on the saliva for symptomatic patients within the first two weeks of symptom onsent, which goes from 0 to ~10^X gene copies/mL.
-*xx*
-
-*Virus concentration on infected individuals*
-
-There is a lack of data regarding the viral loads of infectious viruses in the saliva of infected individuals.
-Therefore, we translated the available concentration data, obtained using quantitative reverse-transcription polymerase chain reaction (RT-qPCR), which gives a measure of the number of copies present a viral RNA sample to infectious virus data.
-
-Care most been taken when using viral RNA data to drive conclusions about infective viruses released by the patient, as PCR data does not distinguish infectios and non-infectious virus @Atkinson2020.
-The use of conversion factor from PCR data has been used in the micribial risk assessment context, due to the limit availability of infetivity data (REF).
-In the absence of GC:PFU data on SARS-CoV-2, we used the relationship reported for Pandemic influenza virus H1N1, as both are enveloped, RNA viruses with similar sizes.
-Currently we have liminted data on the relationship between copy numbers and infectivity, which limits the extend to which the risk assessment can be used.
-However, modeling approaches such as the QMRA can help elucidate the contribution of different exposure routes to the disease transmission and therefore help the development of guidelines.
-*xx*
-
-*Transmission routes*
-We recomend to further develop risk assessment models, incorporating other transmission routes, to understand the relative countributionof each route. 
-*x*
-
-Most of the data used in the models are coming from laboratory experiments. The relevance of the laboratory conditions to real world conditions depend on the parameters.
-
-*Conclusion*
-Even though this risk assessment model relies on many assumptions, some with high uncertainty around them, and the risk caucluation is hihgly simplidied, this model can: provide insigts into the possible effects that different interventions can have in reducing the spread of SARS-CoV-2, serve as a first step in the pathway of building more complex models that allow exploring the relative contribution that different transmission routes could have in spreddin the disease.
-
-about the paper, we can say something like "Although the relative importance of each transmission route remains poorly understood, there has been a growing consensus that contaminated fomites play a critical role in the spread of viruses." @Boone2007
-
-The relative contribution of surface-mediated transmission as compared with other transmission routes such as direct transmission, air, or thorugh the feca-oral route remains to be addressed.
-QMRA can be used to incorporate different transmission routes and answer this question.
-
-QMRA can be used to assess the significance of surface mediated transmission of SARS-CoV-2 [@Haas2020], as it has been used in the past to assess the risk of other transmission routes for MERS-CoV [@Adhikari2019].
-Although not perfect, a risk assessment framework could be use to conduct a hazzard assessment.
-Since offices will need to re-open https://www.osha.gov/shpguidelines/hazard-Identification.html.
-
-Oportunities for managing the risks associated with people's interaction with frequently touched surfaces include contact reduction at a population level, for example reduce the need for contacting trafic ligth buttons by setting them into automatic mode.
-Since surface cleanning of public spaces is not feasible with the frequency needed to significanlty reduce risks, signs that remind people to clean their hands after contact could potentially be encoraged (REF).
-
 ## Supporting Information

### 2020-06-12 src/analyse_risk .R

@@ -17,11 +17,11 @@ simNum = 100 #number of simulations
 # Parameters from distributions and datasets commun to all the scenarios
 GC_TCID50 <- runif(simNum, 100, 1000) # (copies/TCID50)Proportion of infective gene copies [GC:TCID50], Ip (2015) 
 Vs <- runif(simNum, 0.0396, 0.0484) # (mL) Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
-#Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # (cough/min) Number of coughs per minute, Leung (2020)
+Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # (cough/min) Number of coughs per minute, Leung (2020)
 TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # Transfer Efficiency from hand to saliva, Pitol 2017
 k <-  rtri(simNum, min = Q_0.5, max = Q_99.5 , mode = Q_50) # Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
 Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # (copies/ml) Concentration of saliva in the first 2 weeks after symptom onsent 
-Asf <- runif(simNum, Asf_min, Asf_max) # (cm^2) @Chabrelie2018 @Murai1996 @Peters2009a @Sahmel 2015
+Af <- runif(simNum, Asf_min, Asf_max) # (cm^2) @Chabrelie2018 @Murai1996 @Peters2009a @Sahmel 2015
 FSAfm <- runif(simNum, 0.5, 0.8) # Assumed
 
 ##- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
@@ -39,14 +39,14 @@ TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from sur
 angle <- 0.1 #fix, based in the images of @2Bourouiba2014
 
 # Creating data fram with all the variables and numbers simulated from distributions ATM   
[…]

### 2020-06-30 Paper_V2.Rmd
Correcting code... adding time series

@@
 date: "4/7/2020"
-output: html_document
+#output: html_document
+output: pdf_document
 bibliography: references.bib
@@
 Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine).
+The stochastic and mechanistic model used to model the risk of surface-mediated transmission of SARS-CoV-2 infection is based on models described elsewhere [@Julian2018; @Julian2019].
 
@@
 Nevertheless, the presence of coronavirus RNA does not imply the presence of infectious viruses.
-In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses (REF).
+In risk assessments whenever data on the concentration of infectious viruses is not available, it is commonly accepted to use a gene copies to infectivity ratio to convert the concentration measured in gene copies to the concentration of infective viruses [@Chigor2014].
 The gene copy number to infectivity ratio is, to our knowledge, not available for SARS-CoV-2.
@@
 | $V_s$ | mL | Volume of saliva expelled per cough | 0.0484-0.0396 | uniform | [@Nicas2009] volume = 0.044mL, [@Adhikari2019] uniform $\pm$ 10%|
+| $N_{cgh/min}$ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |[@Leung2020], based on data of 17 individuals with seasonal CoV| 
 |$t_{ATM}$ | min |Time spent on the ATM | 0.75(0.38) | Normal | This study. Otained from 20 hours of observational data |
-|$t_{btw-ATM}$ | min |Time between visits | 8.4(7.1) | Normal | This study. Otained from 20 hours of observational data |
-|$t_{btw-btn}$ | min |Time between pressing button | 0.5-20 | Uniform | Assumed |
+|$t_{btw-ATM}$ | min |Time between visits | 0.13, 0.80 | Weibull | This study. Otained from 20 hours of observational data |
+|$t_{btw-btn}$ | min |Time between pressing button | 1-20 | Uniform | Assumed |
 |$x_{ATM}$| cm | ATM inoculation distance | 40 | Point value | This study. Assumed from the observational data  |
@@
 | $TE_{sh-pl}$ | % | Transfer efficiency of virus between plastic to hand | 79.5(21.2) | Normal | @Lopez2013, transfer efficiency of viruses (MS2) between surface and skin at 40-65% humidity | 
-| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | @Pitol2017, transfer efficiency of viruses (MS2) from hand to saliva | 
-| $A_{f}$ | cm^2^ | Finger surface area | 3.8-6.6 | Uniform | Data from @Chabrelie2018 @ @Peters2009a @Sahmel2015 | 
-| $FSA_{fm}$ | - | Fractional surface area between finger and mucous membranes | 0.5-0.8 | Uniform | Assumed | 
+| $TE_{hm}$ | % | Transfer efficiency from hand to mucouse membranes | 20(6.3) | Normal | [@Pitol2017], transfer efficiency of viruses (MS2) from hand to saliva | 
+| $A_{f}$ | cm^2^ | Finger surface area | 3.9-5.9 | Uniform | Fractional surface area of partial finger [@AuYeung2008] times average hand surface area [@EPA2011]| 
 | $k$ | PFU^-1^ | Parameter of the exponential dose-response| 0.00107, 0.00135, 0.00680 | Triangular | Data obtained from QMRAwiki, based on 2 studies @DeAlbuquerque2006, @DeDiego2008, using the from 0.5th, 50th, and 99.5th percentiles as min, mode, and max |
 |:--:|:--:|:--:|:--:|:--:|:--:|
-| N~cgh/min~ |min^-1^ | Number of coughs per minute  | 0.57(1)| Normal |@Leung2020, based on data of 17 individuals with seasonal CoV|
+The input parameters of the distributions are: Normal: mean (SD); Uniform: lower-bound - upper-bound; Triangular: min, mode, max; Weibull: scale, shape.
 
 ### Risk Characterization 
-Monte Carlo simulation method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.
+Monte Carlo simulation: method is used to incorporate uncertainty and variability of the input parameters in the risk characterization.
 

### 2020-06-30 analyse_risk_2.R

@@ -1,0 +1,89 @@
+# Conditions
+    # Material:  Metal
+    # Humidity:  High
+    # Surface:   Buttons
+
+# setup ------------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+library(truncnorm)
+library(EnvStats)
+
+# read and process data  -------------------------------------------------------
+source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020, To 2020 (first 2 weeks after symptom onset)
+source ("src/process_TEhandtosaliva.R") # Pitol 2017 (dry transfer from hand to saliva)
+source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity)
+source ("src/process_decay.R") # van Doremalen 2020
+
+# Simulation parameters:
+simNum = 10 # Number of simulations
+# ----- #
+
+GC_TCID50 <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
+Vs <- runif(simNum, 0.0396, 0.0484) # [mL] Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
+Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # [cough/min] Number of coughs per minute, Leung (2020)
+TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # [unitless] Transfer Efficiency from hand to saliva, Pitol 2017
+k <-  rtri(simNum, min = Q_0.5, max = Q_99.5 , mode = Q_50) # [unitless] Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
+Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # [copies/ml] Concentration of saliva in the first 2 weeks after symptom onsent 
+Af <- runif(simNum, Asf_min, Asf_max) # [cm^2]Fractional surface area @AuYeung2008 (adult front partial finger),hand area (male and female) @EPA2011
+
+# Metal specific parameters
+k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 in steel
+n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface
[…]

### 2020-06-30 master.R

@@ -1,3 +1,0 @@
-# Master code 
-
-source ("src/XXX.R")

### 2020-06-30 src/Risk_Stl_Btn.R

@@ -1,0 +1,100 @@
+# Conditions
+    # Material:  Metal
+    # Humidity:  High
+    # Surface:   Buttons
+
+# setup ------------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+library(truncnorm)
+library(EnvStats)
+
+# read and process data  -------------------------------------------------------
+source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020, To 2020 (first 2 weeks after symptom onset)
+source ("src/process_TEhandtosaliva.R") # Pitol 2017 (dry transfer from hand to saliva)
+source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity)
+source ("src/process_decay.R") # van Doremalen 2020
+source ("src/process_doseresponse.R") # QMRAwiki, using the from 0.5th, 50th, and 99.5 th percentiles as min, mode, and max
+source ("src/process_Asf.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
+
+# Simulation parameters:
+simNum = 10 # Number of simulations
+# ----- #
+
+GC_TCID50 <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
+Vs <- runif(simNum, 0.0396, 0.0484) # [mL] Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
+Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # [cough/min] Number of coughs per minute, Leung (2020)
+TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # [unitless] Transfer Efficiency from hand to saliva, Pitol 2017
+k <-  rtri(simNum, min = Q_0.5, max = Q_99.5 , mode = Q_50) # [unitless] Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
+Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # [copies/ml] Concentration of saliva in the first 2 weeks after symptom onsent 
[…]

### 2020-06-30 src/analyse_risk .R

@@ -21,8 +21,8 @@ Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # (cough/min) Number of coughs
 TEhm <- rtruncnorm(simNum, 0, 1, TE_hm_d, sd_TE_hm_d) # Transfer Efficiency from hand to saliva, Pitol 2017
 k <-  rtri(simNum, min = Q_0.5, max = Q_99.5 , mode = Q_50) # Dose-response parameter, taking median as mode, and 2.5, 97.5% CI as min and max
 Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # (copies/ml) Concentration of saliva in the first 2 weeks after symptom onsent 
-Af <- runif(simNum, Asf_min, Asf_max) # (cm^2) @Chabrelie2018 @Murai1996 @Peters2009a @Sahmel 2015
-FSAfm <- runif(simNum, 0.5, 0.8) # Assumed
+# Af <- runif(simNum, Asf_min, Asf_max) # (cm^2) @Chabrelie2018 @Murai1996 @Peters2009a @Sahmel 2015
+# FSAfm <- runif(simNum, 0.5, 0.8) # Assumed
 
 ##- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
 # Activity specific parameters
@@ -30,8 +30,8 @@ FSAfm <- runif(simNum, 0.5, 0.8) # Assumed
 #                           ---- ATM------   Material : steel
 
 # The TEsh and inactivation on surface were steel specific.
-t_ATM <-  rtruncnorm(simNum, 0, Inf, t_ATM_mean, t_ATM_sd)  # Time at ATM, 20 hours of observational studies
-t_btw_ATM <- rtruncnorm(simNum, 0, Inf, 60/visits_mean, 60/visits_sd) # Time between ATM visits, observational studies
+t_ATM <-  rtruncnorm(simNum, 0, Inf, t_ATM_mean, t_ATM_sd)  # [min] Time at ATM, 20 hours of observational studies
+t_btw_ATM <- rweibull(simNum, visits_shape, visits_scale)   # [min] Time between ATM visits, observational studies
 x <- 40 # Inoculation distance (cm), assumed based on observations
 k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # Half life of CoV-2 in steel
 n_stl <- log(2) / k_stl # Decay rate, function of the halflife of the virus in the surface

### 2020-06-30 src/process_ATM_observation.R

@@ -8,6 +8,7 @@ infile = "data/raw/ATM_observation.csv"
 
 # Library
 library(MASS) 
+library(fitdistrplus)
 
 # read data 
 df_ATM <- read.csv (file=infile, header = TRUE, sep=",") 
@@ -16,11 +17,21 @@ df_ATM <- read.csv (file=infile, header = TRUE, sep=",")
 t_ATM_mean <- mean(df_ATM$t_ATM)
 t_ATM_sd <-  sd(df_ATM$t_ATM)    
 
-# Number of ATM visits per hour (not normally distributed, Re-Sample)
+# Number of ATM visits per hour (not normally distributed)
 visits <- df_ATM$visits_hr
 visits <- visits[!is.na(visits)]
-visits_mean <- mean(visits)
-visits_sd <- sd(visits)
+visits <- as.numeric(visits)
+visits <- visits/60
+visits[visits == 0] <- 0.000001
+
+# Fitting a weibull distribution 
+fit.weibull <- fitdist(visits, distr = "weibull", method = "mle", lower = c(0, 0))
+weibull<-summary(fit.weibull) 
+plot(fit.weibull)
+
+visits_shape <-weibull$estimate [1]
+visits_scale <-weibull$estimate [2]
+
 # output -------------------------------------------------------------------
 
 # save file

### 2020-06-30 src/process_Asf.R

@@ -14,9 +14,10 @@ HSA_men <-  (0.107  * 10000) /2
 
 HSA <- mean(HSA_women:HSA_men) # average men/women area (cm2) 
 
-# Area of finger (uniform using min and max of front partial finger)
-#Asf_min <- HSA * FA_min
-#Asf_max <- HSA * FA_max
+# Area of finger (uniform using min and max of front partial finger). 
+#It's divided by 5 because the FA is mean to be for 5 fingers
+Asf_min <- (HSA * FA_min)/5
+Asf_max <- (HSA * FA_max)/5
 
 #  ------------------------------Method 2 ----------------------------------
 #   Data from the paper @Chabrelie2018, cm2. He used a combination of data from elsewhere
@@ -27,8 +28,8 @@ Asf_peters <- (3.55 + 4.25)/2  # Index finger distal phalanx female and male, @P
 Asf_sahmel <- 3.8 # Six participants, 3 fingers each, @Sahmel2015
 
 
-Asf_min <- Asf_sahmel
-Asd_max <- Asf_murai
+#Asf_min <- Asf_sahmel
+#Asd_max <- Asf_murai
 
 #  ------------------------------Method 3 ----------------------------------
 

### 2020-07-27 Paper_V2.Rmd
Update on prevalence, frequency, and disinfection

@@
 title: "Draft 2"
-author: "Ana Karina"
+author: "Ana Karina and Tim"
 date: "4/7/2020"
@@
 ## Introduction
+SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is transmitted via both direct (person-to-person) and indirect (via a contaminated environment) routes [@WHO2020;@CDC2020].
+Direct transmission appears to be the leading route and occurs via prolonged exposures to respiratory droplets produced while talking, coughing, and sneezing.
+Infection control recommendations, based on the assumption that direct contact transmission is the leading route, include maintaining social/physical distances, wearing masks, case isolation, contact tracing, and quarantine [@Ferretti2020].
+
+Despite knowledge that SARS-CoV-2 transmission is primarily through direct transmission, indirect transmission - or transmission occurring due to interaction with a contaminated environment – remains possible.
+People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and XXX [@Ong2020; @Ye2020].
+Infective coronavirus, including SARS-CoV-2, persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].
+Viruses readily transfer from contaminated surfaces to the skin upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991] and from the skin to the mouth and saliva of individuals [@Pitol2017; @Rusin2002].
+Taken together, this evidence suggests surface contamination poses a risk for indirect SARS-CoV-2 transmission, similar to other respiratory viruses [@Boone2007].
+
+Indirect transmission events of SARS-CoV-2, even if not the dominant transmission route, complicate infection control efforts using contact tracing.
+Contact tracing efforts focus on identifying people sharing spaces coincidentally with cases [@Ferretti2020].
+Indirect transmission could occur in the absence of shared spaces.
+As an example, transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial infection of 31 patients.
+In 6 (20%) of the patients, contact tracing failed to detect direct contacts with other SARS patients, although RNA of SARS-CoV-1 was detected throughout the hospital on drinking water buttons, beds, and chairs [@Chen2004].
+Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected.
+[@Chen2004].
+If indirect transmission is an important route for SARS-CoV-2, control strategies will necessarily have to integrate hand hygiene interventions and surface disinfection alongside contact tracing [@Mcclellan2020].
+
+Despite the potential importance of indirect transmission, it is difficult to estimate its importance relative to direct transmission [@BarOn2020].
+Quantitative Microbial Risk Analysis (QMRA) provides a framework for understanding health risks from indirect transmission and provides insights into potential impacts of infection control recommendations.
+Mechanistic models of fomite-mediated transmission events within the context of QMRA frameworks have been used to inform risks for children interacting with contaminated toys [@Julian2009], sanitation workers collecting and processing urine for nutrient recovery [@Bischel2019], transmission of norovirus within a houseboat [@Canales2019], and impact of surface disinfection [@Ryan2014a].
+QMRA has also been adapted to evaluate risks for hospital transmission of MERS-CoV through droplets and aerosolized particles.
+The analysis highlighted reductions in risk to hospital staff through mask use (>90% estimated risk reduction) and increased air exchange (up to 58% estimated risk reduction) [@Adhikari2019].
+
+In this study, a mechanistic model of indirect transmission within the QMRA framework is developed to estimate likelihood of transmission in community settings and inform guidance on surface disinfection strategies.
+Specifically, the risks of infection with SARS-CoV-2 are estimated for interactions with shared spaces (i.e., use of ATMs, buttons on cross walks and trains).
+Risk reductions are further estimated under various feasible surface disinfection strategies.
+
 ## Materials and Methods 
@@
 Two different scenarios where modelled: single touch (frequently touched buttons) and multiple touches (ATM or cash machine).
-The stochastic and mechanistic model used to model the risk of surface-mediated transmission of SARS-CoV-2 infection is based on models described elsewhere [@Julian2018; @Julian2019].
+The stochastic and mechanistic model used to model the risk of surface-mediated transmission of SARS-CoV-2 infection is based on models described elsewhere [@Julian2018; @Julian2009].
 
@@
 
+"Persons that are sneezing and/or coughing are advised by governments to stay at home."
+
+SARS-CoV-2 QMRA model design to investigate the relative contribution of different transmission routes (contact, droplet, inhalation) for healthcare workers suggested that, although fomite-mediated transmission is not the main route of expossure, it accounts for 8% of the probability of infection without the use of PPE [@2020Jones].
+
+"The guidance also includes considerations based on the current level of community transmission,"
+
 ## Supporting Information

### 2020-07-27 src/Risk_Stl_Btn_Dis.R

@@ -1,8 +1,9 @@
 # Conditions
-    # Material:     Metal
-    # Humidity:     High
-    # Surface:      Buttons
-    # Inoculation:  Cough on hands followed by hand to surface transfer
+# Material:       Metal
+# Humidity:       High
+# Surface:        Buttons
+# Inoculation:    Cough on hands followed by hand to surface transfer
+# Disinfection:   YES
 
 
 # setup ------------------------------------------------------------------------
@@ -18,19 +19,22 @@ source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity)
 source ("src/process_decay.R") # van Doremalen 2020
 source ("src/process_doseresponse.R") # QMRAwiki, using the from 0.5th, 50th, and 99.5 th percentiles as min, mode, and max
 source ("src/process_Asf.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
-
+source ("src/process_prevalence.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
+source ("src/process_disinfection.R") # Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol
+ 
 # Simulation parameters:
 simNum = 10000 # Number of simulations
 # ----- #
 
-#General parameters
-GC_TCID50 <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
+# General parameters
+GC_Inf <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
+adj_DoseResp <- runif(simNum, 0.001, 0.01) # Adjustments for intranasal administration, @Jones2020
 Vs <- runif(simNum, 0.0396, 0.0484) # [mL] Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
-Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # [cough/min] Number of coughs per minute, Leung (2020)
[…]

### 2020-07-27 src/analyse_risk .R

@@ -126,3 +126,7 @@ ggplot(data = corRes, aes(x = type, y = rho)) + geom_bar(stat = "identity")
 
 
 
+
+
+
+

### 2020-07-27 src/process_PublicTransport.R

@@ -1,0 +1,29 @@
+# Processing data on frequency of touching buttons
+# Train schedules Aprox delta time
+
+# Zurich
+S9_Zh      <- 5  #(train) https://moovitapp.com/z%C3%BCrich-3524/lines/S9/350243/4574131/en?customerId=4908&ref=2&poiType=line
+T3_Zh      <- 1  #(tram)  https://moovitapp.com/z%C3%BCrich-3524/lines/3/499595/1853852/en?customerId=4908&ref=2&poiType=line
+Bus_280_Zh <- 2  #(bus)   https://moovitapp.com/z%C3%BCrich-3524/lines/280/499696/1854180/en?customerId=4908&ref=2&poiType=line
+Zh_Gn      <- 20 #(train) https://www.sbb.ch/en/buying/pages/fahrplan/fahrplan.xhtml  (164 min 8 stops, Zurich-Geneva)
+
+# London
+Bus_201_Ln <- 1.5 #(bus)  https://tfl.gov.uk/bus/timetable/201?fromId=490010021E&direction=outbound&intervalId=0&selectedDate=mondaytofriday
+Pad_Rd_Ln  <- 3.3 #(train)  23 min, 7 stations (23/7) Paddington-> Reading 
+Ln_Br     <-  12   #(train) London to Birstol- 8 stops in 98 min
+
+# Paris
+T1_Par     <- 2   #(tram) (33 stops in 79 min) https://wikiroutes.info/es/paris#ab=48.86453-02.44266-48.94209-02.24506&var=1
+M3_Par     <- 1   #(metro) https://wikiroutes.info/es/paris#ab=48.86453-02.44266-48.90015-02.30646&var=1
+Bus_91     <- 2   #(bus) https://wikiroutes.info/es/paris#ab=48.83945-02.31911-48.85125-02.36992&var=1
+
+
+dt_PublicTranspot <- runif(simNum, 1, 20)
+
+
+# Trafic light button for pedestrians in busy London street
+# Assumptions, every time is red for pedestrians people presses it. 
+# Time based on the Traffic Signal Timing Manual (USA) https://ops.fhwa.dot.gov/publications/fhwahop08024/chapter6.htm#6.6
+# Assumtopn based on traffic lights, people touches the light once every 2 min
+dt_TrafficLight <- 2
+

### 2020-07-27 src/process_disinfection.R

@@ -1,0 +1,25 @@
+# Processing data on log reduction based on disinfection
+
+# ------------------ Hand Disinfection data -----------------------------------
+
+#   Article             virus (enveloped)     log10 Red     Sanitizer
+# __________________  _____________________  __________  ________________________
+# @Steinmann2012       vaccinia virus           4        3 Ethanol-based (45->90%)
+# @Steinmann2010       BVDV and HCV             4        2 Alcohol-based (>75%)
+# @Larson2012          Influenza                3.35     1 Alcohol Gel (62%)
+# @Rabenau2005         SARS-CoV               > 4.25     4 Alcohol-based (>75%)
+
+hand_dis <- 4.25
+
+# ----------------- Surface Disinfection data ---------------------------------
+
+#   Article             virus (enveloped)     log10 Red     Sanitizer
+# __________________  _____________________  __________  ________________________
+# @Hulkower2011         TGEV (coronavirus)     3.2        Ethanol (70%)
+#                       MHV (coronavirus)      3.9        Ethanol (70%)
+#                       TGEV (coronavirus)     0.6        Chlorine bleach (0.06%)  *Concentration too low*
+#                       MHV (coronavirus)      0.4        Chlorine bleach (0.06%)  *Concentration too low*
+# @Sattar1989           HCV-229E(human corona)  >3        Chlorine (0.1%)
+#                       HCV-229E(human corona)  >3        Chlorine (0.5%) 
+
+# surf_dis <- runif(simNum, 10^3, 10^4)

### 2020-07-27 src/process_doseresponse.R

@@ -1,16 +1,18 @@
 
-# Processing the data from the Dose Response paramaters
+# Processing the data from the Dose Response parameters
 
 # Data obtained from QMRAwiki http://qmrawiki.org/experiments/sars
-  # Parameter developed from 2 different studies using 2 different coronaviruses: SARS-CoV and Murine hepatitis virus (MHV-1)
-    # @DeAlbuquerque2006. A/J mice with MHV-1 virus, intranasal, survival monitored for 21 days. 
-            # Note: MHV is a commun surrogate of human coronaviruses
-    # @DeDiego2008. 4 groups of the transgenic mice, intranasal, graded doses of rSARS-CoV, survival monitored for 13 days
-  # @Watanabe2010. Combining both dose-response articles to develop the dose-response
+# Parameter developed from 2 different studies using 2 different coronaviruses: SARS-CoV and Murine hepatitis virus (MHV-1)
+#     @DeAlbuquerque2006. A/J mice with MHV-1 virus, intranasal, survival monitored for 21 days. MHV is a common surrogate of human coronaviruses
+#     @DeDiego2008. 4 groups of the transgenic mice, intranasal, graded doses of rSARS-CoV, survival monitored for 13 days
+# @Watanabe2010. Combining both dose-response articles to develop the dose-response
+
+# -> ->  @Jones2020, Adjustements for intranasal administration. uniform(0.01-0.001)
+
 
 # Option 1, using the from 0.5th, 50th, and 99.5 th percentiles as min, mode, and max
 
-# Parameter (k) for the exponential model, based on the 2 articles, using 10,000 bootstrap iterations (MLE)
+# Parameter (k) for the exponential model, based on the 2 articles, using 10,000 bootstrap iterations (MLE), wikiQMRA
 Q_0.5  <- 0.00107
 Q_2.5  <- 0.00128
 Q_5    <- 0.00135 
@@ -23,4 +25,6 @@ Q_99.5 <- 0.00680
 # Estimating SD form percentiles 
 k_sd <- round((Q_95 - Q_5)/(2 *qnorm(0.95)), 5)
 k_mean <- Q_5
+
+
 

### 2020-07-27 src/process_prevalence.R

@@ -1,0 +1,30 @@
+# Processing data on prevalence
+
+# Data obtained from antibody surveys in different countries. The data is meant to be
+# an estimate of the order of magnitude of the prevalence expected in the population. 
+
+#   ---   Antibody surveys   ---  
+
+         # Study           Place               % Seroprevalence  Date(latest test)   Start(estimate,Worldometers)
+    #  @Bendavid2020   Santa Clara, Cal, USA         2.8             April 4          ~ March 1
+    #  @Perez2020      Geneva, Switzerland           9.2             May 6            ~ March 1
+    #  @Pollan2020     Spain                         5               May 11           ~ February 15
+    #  @Erikstrup2020  Netherlands                   1.9             May 3            ~ March 1
+    #  @Filho2020      Rio de Janeiro, Brazil        3.8             April 27         ~ March 1     
+
+# Estimating prevalence at any given time
+# Assuming each person is able to infect for a period of 14 days
+#    Prev [%] =  Seroconversion [%] / (outbreak length [t] /duration of infection [t])
+
+Prev_Cal <- 2.8 / (35/14) 
+Prev_Swt <- 9.2 / (67/14)
+Prev_Spn <- 5   / (85/14)
+Prev_Nth <- 1.9 / (64/14)
+Prev_Brz <- 3.8 / (58/14)
+
+# Prevalence is between 0.4 (Netherlands) and 1.9 (Geneva). 
+# Based on this data, three prevalence scenarios were suggested (0.2,1 and 5%)
+
+low_prev  <- 0.002
+med_prev  <- 0.01
+high_prev <- 0.05

### 2020-07-27 testing code.R

@@ -1,0 +1,162 @@
+# Generating disinfection times, if t_dis=0 it means no disinfection, then the vector is full of 0
+if (t_dis==0) {
+    time_dis <- rep(0, length(time))
+} else {
+    dis_times <- 960 /t_dis  # 960 is the number of minutes in a working day (16 h times 60 min)
+    time_dis <- rep(c(1,rep(0,ceiling(length(time)/dis_times)-1)), ceiling(length(time)/ceiling(length(time)/dis_times)))
+    time_dis <- time_dis[1:length(time)]   
+}
+
+
+
+
+
+#------------------
+
+# Generating times, starting from 7 am, adding delta time, ultil 11 pm, then add 8 hours and start again
+a <- 7*60 # Starting time = 7 am, in minutes
+c <- 1
+time = c()
+day_number = 0 # initializing the couting of days
+days <- 7  # Simulation days
+while(a < (60*24*days)) {
+    if( (a - (day_number*24*60 + 60*23)) >= 0 ) {
+        day_number = day_number + 1
+        a = a + 8 * 60 
+    } else { 
+        a = a + 30}
+    time[c] <- a
+    c = c + 1 
+}
+
+
+t_dis <- 1
+# Generating disinfection times
+chunk <- length(time)/days
+if (t_dis==0) {
+    time_dis <- rep(0, length(time)) # No disinfection
+} else if (t_dis==1) {
+    time_dis <- rep(c(1,rep(0,(chunk-1))),days) # Morning (7am) disinfection
+} else if (t_dis==2) {
+    time_dis <- rep(c(rep(0,ceiling(chunk/2)-1),1,rep(0,floor(chunk/2))),days) # Middle of day (3pm) disinfection
+} else if (t_dis==3) {
+    time_dis <- rep(c(rep(0,(chunk-1)),1),days) # Nigth (11pm) disinfection
+} else {
+    time_dis <- rep(c(rep(0,ceiling(chunk/3)-1),1,rep(0,floor(chunk/3)),1,rep(0,ceiling(chunk/3)-1)),days) # Twice a day ~12 and ~6pm
+}
+
+
+# Generaring the vector of if infected
+prevalence <- rbinom (length(time), 1, df_stl_btn$prev[p])
+
+
+# Create the risk matrix for the conditions
+daily_risk <- cbind(time, time_dis, prevalence)
+
+
[…]

### 2020-07-30 Paper_V2.Rmd
ACS paper outlook

@@
 ## Abstract
+
+Objective. Use a stochastic mechanistic model to quantify the effect of surface [and hand] disinfection on SARS-CoV-2 infection risks. The results from this model can inform cleaning practices.
+
 ## Introduction
@@
 Despite knowledge that SARS-CoV-2 transmission is primarily through direct transmission, indirect transmission - or transmission occurring due to interaction with a contaminated environment – remains possible.
-People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and XXX [@Ong2020; @Ye2020].
-Infective coronavirus, including SARS-CoV-2, persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].
+People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and public spaces in urban areas such as bus stations and public squares [@Ong2020; @Ye2020; Abrahao2020].
+Infective coronavirus persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].
 Viruses readily transfer from contaminated surfaces to the skin upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991] and from the skin to the mouth and saliva of individuals [@Pitol2017; @Rusin2002].
@@
 In 6 (20%) of the patients, contact tracing failed to detect direct contacts with other SARS patients, although RNA of SARS-CoV-1 was detected throughout the hospital on drinking water buttons, beds, and chairs [@Chen2004].
-Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected.
-[@Chen2004].
+Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected [@Chen2004].
 If indirect transmission is an important route for SARS-CoV-2, control strategies will necessarily have to integrate hand hygiene interventions and surface disinfection alongside contact tracing [@Mcclellan2020].
@@
 
-SARS-CoV-2 QMRA model design to investigate the relative contribution of different transmission routes (contact, droplet, inhalation) for healthcare workers suggested that, although fomite-mediated transmission is not the main route of expossure, it accounts for 8% of the probability of infection without the use of PPE [@2020Jones].
+SARS-CoV-2 QMRA model design to investigate the relative contribution of different transmission routes (contact, droplet, inhalation) for healthcare workers suggested that, although fomite-mediated transmission is not the main route of expossure, it accounts for 8% of the probability of infection without the use of PPE [@Jones2020].
 
@@
 ## References
+
+## Info for the Discussion
+Frequency of hand to face contacts:
+Quantified using video recordings 10 subjects wotking on an office during 3 hr.
+The subjects touched their eyes, nostrils, and lops an average of 16(11) times an hour @Nicas2008.
+Medical students touch their faces 23 times per hour, 44% of those contacts were with a mucous membrane @Kwok2015.

### 2020-07-30 Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands/Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands.Rmd
ACS paper outlook

---
journal: jacsat
type: article
author:
  - name: Ana K. Pitol
    aff: Department of Civil and Environmental Engineering, Imperial College London, United Kingdom
    email: \email{a.pitol-garcia@imperial.ac.uk}

  - name: Timothy R. Julian
    aff: Department of Environmental Microbiology, Eawag, Dubendorf, Switzerland
    alsoaff: Department of Epidemiology and Public Health, Swiss Tropical and Public Health Institute, Basel, Switzerland
    email: \email{tim.julian@eawag.ch}

title:
  formatted: Risk reduction of SARS-CoV-2 transmission through disinfection of frequently touched surfaces. A Quantitative Microbial Risk Assessment. 
  short: An \textsf{achemso} demo
abbr: IR,NMR,UV
keywords: American Chemical Society, \LaTeX
abstract: |

  not exceeding 200 words
  A graphic for the Table of Contents (TOC) must be supplied with each Letter
  This graphic can be no wider than 8 cm and no taller than 4 cm and should be prepared following the specifications under Artwork.

bibliography: acstest.bib
csl: american-chemical-society.csl
output: rticles::acs_article
includes:
      in_header: preamble.tex
---

\begin{tocentry}
Add image (TOC)
\end{tocentry}

# Introduction
SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is transmitted via both direct (person-to-person) and indirect (via a contaminated environment) routes [@WHO2020;@CDC2020]. Direct transmission appears to be the leading route and occurs via prolonged exposures to respiratory droplets produced while talking, coughing, and sneezing. Infection control recommendations, based on the assumption that direct contact transmission is the leading route, include maintaining social/physical distances, wearing masks, case isolation, contact tracing, and quarantine [@Ferretti2020].

Despite knowledge that SARS-CoV-2 transmission is primarily through direct transmission, indirect transmission - or transmission occurring due to interaction with a contaminated environment – remains possible.   People infected with SARS-CoV-2 shed the virus into the environment, as evidenced by extensive environmental contamination detected on surfaces in cruise ships, hospitals, and public spaces in urban areas such as bus stations and public squares [@Ong2020; @Ye2020; Abrahao2020].  Infective coronavirus persists in the environment, with experimental evidence of persistence on surfaces for prolonged periods up to 9 days [@Kampf2020].  Viruses readily transfer from contaminated surfaces to the skin upon contact [@Julian2010; @Lopez2013; @Bidawid2004; @Rusin2002; @Ansari1991] and from the skin to the mouth and saliva of individuals [@Pitol2017; @Rusin2002].  Taken together, this evidence suggests surface contamination poses a risk for indirect SARS-CoV-2 transmission, similar to other respiratory viruses [@Boone2007].

Indirect transmission events of SARS-CoV-2, even if not the dominant transmission route, complicate infection control efforts using contact tracing.  Contact tracing efforts focus on identifying people sharing spaces coincidentally with cases [@Ferretti2020].  Indirect transmission could occur in the absence of shared spaces.  As an example, transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial infection of 31 patients. In 6 (20%) of the patients, contact tracing failed to detect direct contacts with other SARS patients, although RNA of SARS-CoV-1 was detected throughout the hospital on drinking water buttons, beds, and chairs [@Chen2004].  Based on these findings, the authors suggest indirect transmission as the cause of transmission amongst a subset of those infected [@Chen2004]. If indirect transmission is an important route for SARS-CoV-2, control strategies will necessarily have to integrate hand hygiene interventions and surface disinfection alongside contact tracing [@Mcclellan2020].

Despite the potential importance of indirect transmission, it is difficult to estimate its importance relative to direct transmission [@BarOn2020]. Quantitative Microbial Risk Analysis (QMRA) provides a framework for understanding health risks from indirect transmission and provides insights into potential impacts of infection control recommendations.  Mechanistic models of fomite-mediated transmission events within the context of QMRA frameworks have been used to inform risks for children interacting with contaminated toys [@Julian2009], sanitation workers collecting and processing urine for nutrient recovery [@Bischel2019], transmission of norovirus within a houseboat [@Canales2019], and impact of surface disinfection [@Ryan2014a].  QMRA has also been adapted to evaluate risks for hospital transmission of MERS-CoV through droplets and aerosolized particles.  The analysis highlighted reductions in risk to hospital staff through mask use (>90% estimated risk reduction) and increased air exchange (up to 58% estimated risk reduction) [@Adhikari2019].

In this study, a mechanistic model of indirect transmission within the QMRA framework is developed to estimate likelihood of transmission in community settings and inform guidance on surface disinfection strategies. Specifically, the risks of infection with SARS-CoV-2 are estimated for interactions with shared spaces (i.e., use of ATMs, buttons on cross walks and trains).  Risk reductions are further estimated under various feasible surface disinfection strategies.

# Results and discussion

Frequently touched surfaces such as traffic light and train buttons are of risk concern due to their frequent use by multiple people.
SARS-CoV-2 contamination has been reported in public surfaces of an urban area in Brazil. Concentration of viruses [genome units /m2] was between 20 and 2990. The surfaces were bench, ground, handrail made out metal and concrete [@Abrahao2020].

## Surface disinfection

XX% reduction

## Floats

New float types are automatically set up by the class file.  The
means graphics are included as follows (Scheme \ref{sch:example}).  As
illustrated, the float is ``here'' if possible.
\begin{scheme}
  Your scheme graphic would go here: \texttt{.eps} format\\
  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
  these can be scaled without loss of quality, and can be\\
  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
  %\includegraphics{graphic}
  \caption{An example scheme}
  \label{sch:example}
\end{scheme}

```{r,echo=FALSE,fig.cap='test'}
plot(1:10)
```

\begin{figure}
  As well as the standard float types \texttt{table}\\
  and \texttt{figure}, the class also recognises\\
  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
  \caption{An example figure}
  \label{fgr:example}
\end{figure}

Charts, figures and schemes do not necessarily have to be labelled or
captioned.  However, tables should always have a title. It is
possible to include a number and label for a graphic without any
title, using an empty argument to the \texttt{\textbackslash caption}
macro.

The use of the different floating environments is not required, but
it is intended to make document preparation easier for authors. In
general, you should place your graphics where they make logical
sense; the production process will move them if needed.

# Materials and Methods

We modeled the risk of SARS-CoV-2 infection due to contact with frequently touched buttons and estimated the effect that different surface disinfection practices have on infection risk. The sequence of events leading to the viral inoculation of the susceptible individual is considered to be the following: 1) contamination event (infected person coughing on the hand followed by a hand-to-surface contact), 2) virus decay on the surface, 3) virus transfer from the surface to finger, and 4) virus transfer from finger to facial mucous membranes. The stochastic-mechanistic model used herein is based on models described elsewhere [@Julian2018; @Julian2009]. Two scenarios were considered: risk of infection with and without surface disinfection. Additionally, different disinfection regimes were tested.

Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input. The viral loads were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020]. PCR data does not distinguish between infectious and non-infectious virus. To convert gene copies to infections virus, we used a ratio reported for influenza viruses [@Ip2016].

The proportion of viral particles emitted in a cough that is transferred to the hand when the hand is used to cover the mouth was estimated using a conical dispersion model  [@Nicas2006] -  the particles were assumed to be conically spread over a 80 angle, at a distance of 5cm. The transfer of viruses from the hand of the infected individual to the surface was estimated with: $Csurface=Chand*TEsh$, where $Chand$ is the concentration of viruses on the hand $[virus/cm^2]$, and TEsh $[%]$ is the transfer efficiency of viruses between the surface (assumed to be metal) and the hand. Once the surface was contaminated, the concentration of virus in the surface at any given time was calculated as follows $Csurf(t)= Csurf(0)$ $e^{-nt}$, where $Csurf(t)$ $[virus/cm^2]$ is the concentration of infective viruses at time $t$, $Csurf(0)$ $[virus/cm2^]$ is the initial concentration of virus in the surface, $t$ $[min]$ is the time lapsed after surface inoculation, and $n$ $[m^{-1}]$ is the decay rate of the virus in the selected surface (metal or plastic).

(Eq. \ref{eqn:example}).
\begin{equation}
  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
\end{equation}

\begin{equation}
  \begin{minipage}[c]{0.80\linewidth}
    \centering
    As illustrated here, the width of \\
    the minipage needs to allow some  \\
    space for the number to fit in to.
    %\includegraphics{graphic}
  \end{minipage}
  \label{eqn:graphic}
\end{equation}

# Experimental

\begin{acknowledgement}

The author thanks Sunil Dogga, Emmanuel Frousteyand, and Danilo Cuccato for their help with the model.

\end{acknowledgement}

\begin{suppinfo}

Experimental procedures and
characterization data for all new compounds. The class will
automatically add a sentence pointing to the information on-line:

\end{suppinfo}
## References

### 2020-07-30 Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands/Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands.tex
ACS paper outlook

\documentclass[journal=jacsat,manuscript=article]{achemso}
\usepackage[version=3]{mhchem}
\usepackage{amsmath}
\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
\author{Ana K. Pitol}
\affiliation{Department of Civil and Environmental Engineering, Imperial College
London, United Kingdom}
\email{a.pitol-garcia@imperial.ac.uk}
\author{Timothy R. Julian}
\affiliation{Department of Environmental Microbiology, Eawag, Dubendorf, Switzerland}
\alsoaffiliation{Department of Epidemiology and Public Health, Swiss Tropical and Public
Health Institute, Basel, Switzerland}
\email{tim.julian@eawag.ch}

\abbreviations{IR,NMR,UV}

\keywords{American Chemical Society, \LaTeX}

\title[An \textsf{achemso} demo]{Risk reduction of SARS-CoV-2 transmission through disinfection of
frequently touched surfaces. A Quantitative Microbial Risk Assessment.}
\makeatletter
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
% pandoc header

\begin{document}
\begin{abstract}
not exceeding 200 words A graphic for the Table of Contents (TOC) must
be supplied with each Letter This graphic can be no wider than 8 cm and
no taller than 4 cm and should be prepared following the specifications
under Artwork.
\end{abstract}
\begin{tocentry}
Add image (TOC)
\end{tocentry}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is
transmitted via both direct (person-to-person) and indirect (via a
contaminated environment) routes.\textsuperscript{1,2} Direct
transmission appears to be the leading route and occurs via prolonged
exposures to respiratory droplets produced while talking, coughing, and
sneezing. Infection control recommendations, based on the assumption
that direct contact transmission is the leading route, include
maintaining social/physical distances, wearing masks, case isolation,
contact tracing, and quarantine.\textsuperscript{3}

Despite knowledge that SARS-CoV-2 transmission is primarily through
direct transmission, indirect transmission - or transmission occurring
due to interaction with a contaminated environment -- remains possible.
People infected with SARS-CoV-2 shed the virus into the environment, as
evidenced by extensive environmental contamination detected on surfaces
in cruise ships, hospitals, and public spaces in urban areas such as bus
stations and public squares {[}4; 5; Abrahao2020{]}. Infective
coronavirus persists in the environment, with experimental evidence of
persistence on surfaces for prolonged periods up to 9
days.\textsuperscript{6} Viruses readily transfer from contaminated
surfaces to the skin upon contact\textsuperscript{7--11} and from the
skin to the mouth and saliva of individuals.\textsuperscript{10,12}
Taken together, this evidence suggests surface contamination poses a
risk for indirect SARS-CoV-2 transmission, similar to other respiratory
viruses.\textsuperscript{13}

Indirect transmission events of SARS-CoV-2, even if not the dominant
transmission route, complicate infection control efforts using contact
tracing. Contact tracing efforts focus on identifying people sharing
spaces coincidentally with cases.\textsuperscript{3} Indirect
transmission could occur in the absence of shared spaces. As an example,
transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial
infection of 31 patients. In 6 (20\%) of the patients, contact tracing
failed to detect direct contacts with other SARS patients, although RNA
of SARS-CoV-1 was detected throughout the hospital on drinking water
buttons, beds, and chairs.\textsuperscript{14} Based on these findings,
the authors suggest indirect transmission as the cause of transmission
amongst a subset of those infected.\textsuperscript{14} If indirect
transmission is an important route for SARS-CoV-2, control strategies
will necessarily have to integrate hand hygiene interventions and
surface disinfection alongside contact tracing.\textsuperscript{15}

Despite the potential importance of indirect transmission, it is
difficult to estimate its importance relative to direct
transmission.\textsuperscript{16} Quantitative Microbial Risk Analysis
(QMRA) provides a framework for understanding health risks from indirect
transmission and provides insights into potential impacts of infection
control recommendations. Mechanistic models of fomite-mediated
transmission events within the context of QMRA frameworks have been used
to inform risks for children interacting with contaminated
toys,\textsuperscript{17} sanitation workers collecting and processing
urine for nutrient recovery,\textsuperscript{18} transmission of
norovirus within a houseboat,\textsuperscript{19} and impact of surface
disinfection.\textsuperscript{20} QMRA has also been adapted to evaluate
risks for hospital transmission of MERS-CoV through droplets and
aerosolized particles. The analysis highlighted reductions in risk to
hospital staff through mask use (\textgreater90\% estimated risk
reduction) and increased air exchange (up to 58\% estimated risk
reduction).\textsuperscript{21}

In this study, a mechanistic model of indirect transmission within the
QMRA framework is developed to estimate likelihood of transmission in
community settings and inform guidance on surface disinfection
strategies. Specifically, the risks of infection with SARS-CoV-2 are
estimated for interactions with shared spaces (i.e., use of ATMs,
buttons on cross walks and trains). Risk reductions are further
estimated under various feasible surface disinfection strategies.

\hypertarget{results-and-discussion}{%
\section{Results and discussion}\label{results-and-discussion}}

Frequently touched surfaces such as traffic light and train buttons are
of risk concern due to their frequent use by multiple people. SARS-CoV-2
contamination has been reported in public surfaces of an urban area in
Brazil. Concentration of viruses {[}genome units /m2{]} was between 20
and 2990. The surfaces were bench, ground, handrail made out metal and
concrete.\textsuperscript{22}

\hypertarget{surface-disinfection}{%
\subsection{Surface disinfection}\label{surface-disinfection}}

XX\% reduction

\hypertarget{floats}{%
\subsection{Floats}\label{floats}}

New float types are automatically set up by the class file. The means
graphics are included as follows (Scheme \ref{sch:example}). As
illustrated, the float is ``here'' if possible.

\begin{scheme}
  Your scheme graphic would go here: \texttt{.eps} format\\
  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
  these can be scaled without loss of quality, and can be\\
  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
  %\includegraphics{graphic}
  \caption{An example scheme}
  \label{sch:example}
\end{scheme}

\begin{figure}
\centering
\includegraphics{Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{test}
\end{figure}

\begin{figure}
  As well as the standard float types \texttt{table}\\
  and \texttt{figure}, the class also recognises\\
  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
  \caption{An example figure}
  \label{fgr:example}
\end{figure}

Charts, figures and schemes do not necessarily have to be labelled or
captioned. However, tables should always have a title. It is possible to
include a number and label for a graphic without any title, using an
empty argument to the \texttt{\textbackslash caption} macro.

The use of the different floating environments is not required, but it
is intended to make document preparation easier for authors. In general,
you should place your graphics where they make logical sense; the
production process will move them if needed.

\hypertarget{maths}{%
\subsection{Math(s)}\label{maths}}

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

We modeled the risk of SARS-CoV-2 infection due to contact with
frequently touched buttons and estimated the effect that different
surface disinfection practices have on infection risk. The sequence of
events leading to the viral inoculation of the susceptible individual is
considered to be the following: 1) contamination event (infected person
coughing on the hand followed by a hand-to-surface contact), 2) virus
decay on the surface, 3) virus transfer from the surface to finger, and
4) virus transfer from finger to facial mucous membranes. The
stochastic-mechanistic model used herein is based on models described
elsewhere.\textsuperscript{17,23} Two scenarios were considered: risk of
infection with and without surface disinfection. Additionally, different
disinfection regimes were tested.

Viral loads in the saliva or sputum of symptomatic COVID-19 patients
within the first 14 days of symptom onset were used as model input. The
viral loads were measured using quantitative reverse-transcription
polymerase chain reaction (RT-qPCR).\textsuperscript{24--27} PCR data
does not distinguish between infectious and non-infectious virus. To
convert gene copies to infections virus, we used a ratio reported for
influenza viruses.\textsuperscript{28}

The proportion of viral particles emitted in a cough that is transferred
to the hand when the hand is used to cover the mouth was estimated using
a conical dispersion model\textsuperscript{{\textbf{???}}} - the
particles were assumed to be conically spread over a 80 angle, at a
distance of 5cm. The transfer of viruses from the hand of the infected
individual to the surface was estimated with: \(Csurface=Chand*TEsh\),
where \(Chand\) is the concentration of viruses on the hand
\([virus/cm^2]\), and TEsh \([%]
\) is the transfer efficiency of viruses between the surface (assumed to
be metal) and the hand. Once the surface was contaminated, the
concentration of virus in the surface at any given time was calculated
as follows \(Csurf(t)= Csurf(0)\) \(e^{-nt}\), where \(Csurf(t)\)
\([virus/cm^2]\) is the concentration of infective viruses at time
\(t\), \(Csurf(0)\) \([virus/cm2^]\) is the initial concentration of
virus in the surface, \(t\) \([min]\) is the time lapsed after surface
inoculation, and \(n\) \([m^{-1}]\) is the decay rate of the virus in
the selected surface (metal or plastic).

(Eq. \ref{eqn:example}). \begin{equation}
  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
\end{equation}

\begin{equation}
  \begin{minipage}[c]{0.80\linewidth}
    \centering
    As illustrated here, the width of \\
    the minipage needs to allow some  \\
    space for the number to fit in to.
    %\includegraphics{graphic}
  \end{minipage}
  \label{eqn:graphic}
\end{equation}

\hypertarget{experimental}{%
\section{Experimental}\label{experimental}}

\begin{acknowledgement}

The author thanks Sunil Dogga, Emmanuel Frousteyand, and Danilo Cuccato for their help with the model.

\end{acknowledgement}

\begin{suppinfo}

Experimental procedures and
characterization data for all new compounds. The class will
automatically add a sentence pointing to the information on-line:

\end{suppinfo}

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-WHO2020}{}%
(1) WHO. \emph{Scientific brief} \textbf{2020}, \emph{29} (March),
10--12.

\leavevmode\hypertarget{ref-CDC2020}{}%
(2) CDC. \emph{How COVID-19 Spreads}; 2020; Vol. 2019, p 2020.

\leavevmode\hypertarget{ref-Ferretti2020}{}%
(3) Ferretti, L.; Wymant, C.; Kendall, M.; Zhao, L.; Nurtay, A.;
Abeler-Dörner, L.; Parker, M.; Bonsall, D.; Fraser, C. \emph{Science}
\textbf{2020}, \emph{368} (6491), 0--8.

\leavevmode\hypertarget{ref-Ong2020}{}%
(4) Ong, S. W. X.; Tan, Y. K.; Chia, P. Y.; Lee, T. H.; Ng, O. T.; Wong,
M. S. Y.; Marimuthu, K. \emph{JAMA - Journal of the American Medical
Association} \textbf{2020}, 2--4.

\leavevmode\hypertarget{ref-Ye2020}{}%
(5) Ye, G.; Lin, H.; Chen, L.; Wang, S.; Zeng, Z.; Wang, W.; Zhang, S.;
Rebmann, T.; Li, Y.; Pan, Z.; Yang, Z.; Wang, Y.; Wang, F.; Qian, Z.;
Wan, X. \emph{medRxiv preprint} \textbf{2020}, 1--20.

\leavevmode\hypertarget{ref-Kampf2020}{}%
(6) Kampf, G.; Todt, D.; Pfaender, S.; Steinmann, E. \emph{Journal of
Hospital Infection} \textbf{2020}, \emph{104} (3), 246--251.

\leavevmode\hypertarget{ref-Julian2010}{}%
(7) Julian, T. R.; Leckie, J. O.; Boehm, A. B. \emph{Journal of Applied
Microbiology} \textbf{2010}, \emph{109}, 1868--1874.

\leavevmode\hypertarget{ref-Lopez2013}{}%
(8) Lopez, G. U.; Gerba, C. P.; Tamimi, A. H.; Kitajima, M.; Maxwell, S.
L.; Rose, J. B. \emph{Applied and Environmental Microbiology}
\textbf{2013}, \emph{79}, 5728--5734.

\leavevmode\hypertarget{ref-Bidawid2004}{}%
(9) Bidawid, S.; Malik, N.; Adegbunrin, O.; Sattar, S. A.; Farber, J. M.
\emph{J Food Prot} \textbf{2004}, \emph{67} (1), 103--109.

\leavevmode\hypertarget{ref-Rusin2002}{}%
(10) Rusin, P.; Maxwell, S.; Gerba, C. \emph{Journal of Applied
Microbiology} \textbf{2002}, \emph{93}, 585--592.

\leavevmode\hypertarget{ref-Ansari1991}{}%
(11) Ansari, S.; Springthorpe, V.; Sattar, S.; Rivard, S.; Rahman, M.
\emph{Journal of Clinical Microbiology} \textbf{1991}, \emph{29} (10),
2115--2119.

\leavevmode\hypertarget{ref-Pitol2017}{}%
(12) Pitol, A. K.; Bischel, H. N.; Kohn, T.; Julian, T. R.
\emph{Environmental Science and Technology} \textbf{2017}.

\leavevmode\hypertarget{ref-Boone2007}{}%
(13) Boone, S. A.; Gerba, C. P. \emph{Applied and Environmental
Microbiology} \textbf{2007}, \emph{73} (6), 1687--1696.

\leavevmode\hypertarget{ref-Chen2004}{}%
(14) Chen, Y. C.; Huang, L. M.; Chan, C. C.; Su, C. P.; Chang, S. C.;
Chang, Y. Y.; Chen, M. L.; Hung, C. C.; Chen, W. J.; Lin, F. Y.; Lee, Y.
T.; Chen, D. S.; Lee, Y. T.; Teng, C. M.; Yang, P. C.; Ho, H. N.; Chen,
P. J.; Chang, M. F.; Wang, J. T.; Kao, C. L.; Wang, W. K.; Hsiao, C. H.;
Hsueh, P. R. \emph{Emerging Infectious Diseases} \textbf{2004},
\emph{10} (5), 782--788.

\leavevmode\hypertarget{ref-Mcclellan2020}{}%
(15) Gottlieb, S.; Rivers, C.; Mcclellan, M. B.; Silvis, L.; Watson, C.
\emph{American Enterprise Institute} \textbf{2020}, 1--16.

\leavevmode\hypertarget{ref-BarOn2020}{}%
(16) Bar-on, Y. M.; Flamholz, A.; Phillips, R.; Milo, R. \emph{eLife}
\textbf{2020}.

\leavevmode\hypertarget{ref-Julian2009}{}%
(17) Julian, T. R.; Canales, R. a.; Leckie, J. O.; Boehm, A. B.
\emph{Risk Analysis} \textbf{2009}, \emph{29}, 617--632.

\leavevmode\hypertarget{ref-Bischel2019}{}%
(18) Bischel, H. N.; Caduff, L.; Schindelholz, S.; Kohn, T.; Julian, T.
R. \emph{Environmental Science and Technology} \textbf{2019}, \emph{53}
(12), 7055--7067.

\leavevmode\hypertarget{ref-Canales2019}{}%
(19) Canales, R. A.; Reynolds, K. A.; Wilson, A. M.; Fankem, S. L. M.;
Weir, M. H.; Rose, J. B.; Abd-Elmaksoud, S.; Gerba, C. P. \emph{Journal
of Occupational and Environmental Hygiene} \textbf{2019}, \emph{16} (1),
16--26.

\leavevmode\hypertarget{ref-Ryan2014a}{}%
(20) Ryan, M. O.; Haas, C. N.; Gurian, P. L.; Gerba, C. P.; Panzl, B.
M.; Rose, J. B. \emph{American Journal of Infection Control}
\textbf{2014}, \emph{42} (11), 1165--1172.

\leavevmode\hypertarget{ref-Adhikari2019}{}%
(21) Adhikari, U.; Chabrelie, A.; Weir, M.; Boehnke, K.; McKenzie, E.;
Ikner, L.; Wang, M.; Wang, Q.; Young, K.; Haas, C. N.; Rose, J.;
Mitchell, J. \emph{Risk Analysis} \textbf{2019}, \emph{39} (12),
2608--2624.

\leavevmode\hypertarget{ref-Abrahao2020}{}%
(22) Abrahao, J. S.; Pengo, L. S.; Rezende, I. M.; Rodrigues, R.;
Crispim, A. P. C.; Moura, C. C.; Mendonca, D. C.; Reis, E.; Souza, F.;
Oliveira, G. F. G. G. P. G.; Domingos, I. J. da S.; Boratto, P.; Silva,
P. H. B. e; Queiroz, V. F.; Silva, T. B. de S.; Oliveira, G. F. G. G. P.
G.; Alves, V. de S.; Alves, P. A.; Kroon, E. G.; Trindade, G. de S.;
Drumond, B. P.; Jônatas, A.; Abrahão, S.; Sacchetto, L.; Rezende, I. M.;
Rodrigues, R.; Paula, A.; Crispim, C.; Moura, C. C.; Correa, D.; Reis,
E.; Souza, F.; Fernanda, G.; Oliveira, G. F. G. G. P. G.; José, I.;
Domingos, S.; Boratto, P.; Henrique, P. \emph{medRxiv} \textbf{2020}.

\leavevmode\hypertarget{ref-Julian2018}{}%
(23) Julian, T. R.; Vithanage, H. S. K.; Chua, M. L.; Kuroda, M.; Pitol,
A. K.; Nguyen, P. H. L.; Canales, R. A.; Fujii, S.; Harada, H.
\emph{Science of The Total Environment} \textbf{2018}, \emph{635},
120--131.

\leavevmode\hypertarget{ref-Wolfel2020}{}%
(24) Wölfel, R.; Corman, V. M.; Guggemos, W.; Seilmaier, M.; Zange, S.;
Müller, M. A.; Niemeyer, D.; Jones, T. C.; Vollmar, P.; Rothe, C.;
Hoelscher, M.; Bleicker, T.; Brünink, S.; Schneider, J.; Ehmann, R.;
Zwirglmaier, K.; Drosten, C.; Wendtner, C. \emph{Nature} \textbf{2020},
1--14.

\leavevmode\hypertarget{ref-Pan2020}{}%
(25) Pan, Y.; Zhang, D.; Yang, P.; Poon, L. L. M.; Wang, Q. \emph{The
Lancet Infectious Diseases} \textbf{2020}, \emph{20} (4), 411--412.

\leavevmode\hypertarget{ref-Kim2020}{}%
(26) Kim, J. Y.; Ko, J. H.; Kim, Y.; Kim, Y. J.; Kim, J. M.; Chung, Y.
S.; Kim, H. M.; Han, M. G.; Kim, S. Y.; Chin, B. S. \emph{Journal of
Korean Medical Science} \textbf{2020}, \emph{35} (7), 1--7.

\leavevmode\hypertarget{ref-To2020}{}%
(27) To, K. K. W.; Tsang, O. T. Y.; Leung, W. S.; Tam, A. R.; Wu, T. C.;
Lung, D. C.; Yip, C. C. Y.; Cai, J. P.; Chan, J. M. C.; Chik, T. S. H.;
Lau, D. P. L.; Choi, C. Y. C.; Chen, L. L.; Chan, W. M.; Chan, K. H.;
Ip, J. D.; Ng, A. C. K.; Poon, R. W. S.; Luo, C. T.; Cheng, V. C. C.;
Chan, J. F. W.; Hung, I. F. N.; Chen, Z.; Chen, H.; Yuen, K. Y.
\emph{The Lancet Infectious Diseases} \textbf{2020}, \emph{20} (5),
565--574.

\leavevmode\hypertarget{ref-Ip2016}{}%
(28) Ip, D. K. M.; Lau, L. L. H.; Chan, K. H.; Fang, V. J.; Leung, G.
M.; Peiris, M. J. S.; Cowling, B. J. \emph{Clinical Infectious Diseases}
\textbf{2015}, \emph{62} (4), 431--437.
\end{document}

### 2020-07-30 src/Risk_Stl_Btn_Dis.R

@@ -12,6 +12,7 @@ library(tidyverse)
 library(truncnorm)
 library(EnvStats)
 
+
 # read and process data  -------------------------------------------------------
 source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020, To 2020 (first 2 weeks after symptom onset)
 source ("src/process_TEhandtosaliva.R") # Pitol 2017 (dry transfer from hand to saliva)
@@ -23,7 +24,7 @@ source ("src/process_prevalence.R") # Fractional surface area, @EPA2011 and @AuY
 source ("src/process_disinfection.R") # Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol
  
 # Simulation parameters:
-simNum = 10000 # Number of simulations
+simNum = 5000 # Number of simulations
 # ----- #
 
 # General parameters
@@ -40,15 +41,16 @@ FSA <- runif(simNum, Asf_min, Asf_max) # [cm^2] Fractional surface area @AuYeung
 k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # [min] Half life of CoV-2 in steel
 n_stl <- log(2) / k_stl # [min-1] Decay rate, function of the halflife of the virus in the surface
 TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
+TEhs_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from hand to surface for stainless steel RH=[40-65%]
 
 # Scenario specific parameters
 angle <- 0.8 # Fixed, based in the images of @2Bourouiba2014
 # prev <- runif(simNum, 0.002, 0.05) # For sensitivity analysis from 0.02 to 5% prevalence.
-prev <- high_prev #  low_prev = 0.002, med_prev = 0.01, high_prev = 0.05, data justified in "process_prevalence.R"
+prev <- med_prev #  low_prev = 0.002, med_prev = 0.01, high_prev = 0.05, data justified in "process_prevalence.R"
[…]

### 2020-07-30 src/plot_risk.R

@@ -1,0 +1,53 @@
+## plot _ Risk with and without disinfection ##
+
+# Conditions
+# Material:       Metal
+# Humidity:       High
+# Surface:        Buttons
+# Inoculation:    Cough on hands followed by hand to surface transfer
+# Disinfection:   YES
+
+
+# setup ------------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+library(plyr)
+library(scales)
+
+# input -------------------------------------------------------------------
+# data file
+infile = "data/processed/compilation_risk.csv"
+outfile = "figures/plot_risk.pdf"
+
+# process -----------------------------------------------------------------
+
+# read data 
+df_risk_dis <- read.csv (file=infile, header = TRUE, sep=",")
+
+# Preparing the data
+df_risk_dis <- gather(df_risk_dis, "prevalence", "risk", 2:4)
+df_risk_dis$Disinfection <- as.character(df_risk_dis$Disinfection)
+
+# Selecting only the median values
+median_risk <- filter(df_risk_dis,Percentile ==0.50)
+
+# Plot risk as a function of prevalence and disinfection
+
+#    ----------version_1 --------- 
+median_risk$prevalence <- revalue(median_risk$prevalence , c("pHigh"="5%", "pLow"="0.2%", "pMed"="1%"))
+median_risk$Disinfection <- revalue(median_risk$Disinfection , c("0"="None","1"="Once: 7am", "2"="Once: 3pm", "4"="Twice: 12pm and 6pm"))
+
+# x and y axis are transformed and formatted
+p2 <- ggplot(median_risk, aes(prevalence, risk)) + geom_point(alpha = 1, aes(shape = Disinfection, color = Disinfection)) +
+    scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                  labels = trans_format("log10", math_format(10^.x))) +
+    theme_bw()
+p2
+
+
+
+# save
+ggsave (plt_risk , filename = outfile, units = "in", dpi = 600)
+
+                
+                

### 2020-07-30 src/process_SurfaceConcentration.R

@@ -1,0 +1,27 @@
+# Processing data on concentrations of SARS-CoV-2 on public surfaces
+
+# ------------------ Surface contamination data ---------------------------------
+
+#                                                                concentration
+#   Article             method      place          surface       gene copies/m2   
+# _________________  __________  _____________  _____________    ______________
+# @Abrahao2020          qPCR     Hospital bus   Bench (metal)         620  
+#  (Brazil)                      Hospital bus   Bench (metal)         80 
+#                                Hospital bus   Bench (metal)         70 
+#                                Hospital bus   Bench (metal)         150 
+#                                Hospital bus   Ground(concrete)      20
+#                                Hospital bus   Ground(concrete)      60
+
+#                                Hospital front Ground(concrete)      2990
+
+#                                Bus terminal   Handrail (metal)      200
+#                                Bus terminal   Handrail (metal)      50   
+#                                Bus terminal   Handrail (metal)      2110
+#                                Bus terminal   Handrail (metal)      520
+#                                Bus terminal   Handrail (metal)      70
+#                                Bus terminal   Handrail (metal)      2080
+#
+#                                Public square  Table (concrete)      200
+#                                Public square  Bench (concrete)      20
+#                                Public square  Bench (concrete)      70
+#                                Public square  Bench (concrete)      70

### 2020-07-30 testing code.R

@@ -1,162 +1,0 @@
-# Generating disinfection times, if t_dis=0 it means no disinfection, then the vector is full of 0
-if (t_dis==0) {
-    time_dis <- rep(0, length(time))
-} else {
-    dis_times <- 960 /t_dis  # 960 is the number of minutes in a working day (16 h times 60 min)
-    time_dis <- rep(c(1,rep(0,ceiling(length(time)/dis_times)-1)), ceiling(length(time)/ceiling(length(time)/dis_times)))
-    time_dis <- time_dis[1:length(time)]   
-}
-
-
-
-
-
-#------------------
-
-# Generating times, starting from 7 am, adding delta time, ultil 11 pm, then add 8 hours and start again
-a <- 7*60 # Starting time = 7 am, in minutes
-c <- 1
-time = c()
-day_number = 0 # initializing the couting of days
-days <- 7  # Simulation days
-while(a < (60*24*days)) {
-    if( (a - (day_number*24*60 + 60*23)) >= 0 ) {
-        day_number = day_number + 1
-        a = a + 8 * 60 
-    } else { 
-        a = a + 30}
-    time[c] <- a
-    c = c + 1 
-}
-
-
-t_dis <- 1
-# Generating disinfection times
-chunk <- length(time)/days
-if (t_dis==0) {
-    time_dis <- rep(0, length(time)) # No disinfection
-} else if (t_dis==1) {
-    time_dis <- rep(c(1,rep(0,(chunk-1))),days) # Morning (7am) disinfection
-} else if (t_dis==2) {
-    time_dis <- rep(c(rep(0,ceiling(chunk/2)-1),1,rep(0,floor(chunk/2))),days) # Middle of day (3pm) disinfection
-} else if (t_dis==3) {
-    time_dis <- rep(c(rep(0,(chunk-1)),1),days) # Nigth (11pm) disinfection
-} else {
-    time_dis <- rep(c(rep(0,ceiling(chunk/3)-1),1,rep(0,floor(chunk/3)),1,rep(0,ceiling(chunk/3)-1)),days) # Twice a day ~12 and ~6pm
-}
-
-
-# Generaring the vector of if infected
-prevalence <- rbinom (length(time), 1, df_stl_btn$prev[p])
-
-
-# Create the risk matrix for the conditions
-daily_risk <- cbind(time, time_dis, prevalence)
-
-
-
[…]

### 2020-09-11 ES&T paper/Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands.Rmd
Updating the code

--- 2020-07-30 Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands/Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands.Rmd
@@
 \begin{tocentry}
-Add image (TOC)
+\includegraphics{Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands_files/figure-latex/TOC.pdf}
 \end{tocentry}
@@
 
+@Wilson2020 QMRA to evaluate the reduction in the risk of infection of SARS-CoV-2 for mask wearers.
+
 In this study, a mechanistic model of indirect transmission within the QMRA framework is developed to estimate likelihood of transmission in community settings and inform guidance on surface disinfection strategies.
@@
 
+# Materials and Methods
+
+We modeled the risk of SARS-CoV-2 infection due to contact with frequently touched buttons and estimated the effect that different surface disinfection practices have on infection risk.
+The sequence of events leading to the viral inoculation of the susceptible individual is considered to be the following: 1) contamination event (infected person coughing on the hand followed by a hand-to-surface contact), 2) virus decay on the surface, 3) virus transfer from the surface to finger, and 4) virus transfer from finger to facial mucous membranes.
+The stochastic-mechanistic model used herein is based on models described elsewhere [@Julian2018; @Julian2009].
+Two scenarios were considered: risk of infection with and without surface disinfection.
+Additionally, different disinfection regimes were tested.
+
+Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input.
+The viral loads were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020].
+PCR data does not distinguish between infectious and non-infectious virus.
+To convert gene copies to infections virus, we used a ratio reported for influenza viruses [@Ip2016].
+
+The proportion of viral particles emitted in a cough that is transferred to the hand when the hand is used to cover the mouth was estimated using a conical dispersion model  [@Nicas2009] -  the particles were assumed to be conically spread over a 80 angle, at a distance of 5cm.
+The transfer of viruses from the hand of the infected individual to the surface was estimated with: $Csurface=Chand*TEsh$, where $Chand$ is the concentration of viruses on the hand $[virus/cm^2]$, and TEsh $[%]$ is the transfer efficiency of viruses between the surface (assumed to be metal) and the hand.
+Once the surface was contaminated, the concentration of virus in the surface at any given time was calculated as follows $Csurf(t)= Csurf(0)$ $e^{-nt}$, where $Csurf(t)$ $[virus/cm^2]$ is the concentration of infective viruses at time $t$, $Csurf(0)$ $[virus/cm2^]$ is the initial concentration of virus in the surface, $t$ $[min]$ is the time lapsed after surface inoculation, and $n$ $[m^{-1}]$ is the decay rate of the virus in the selected surface (metal or plastic).
+
+## Alternative way of presenting the equations
+The equations can be presented as follows:
+
+(Eq. \ref{eqn:Inactivation})
+\begin{equation}
+C(t)= C(0) e^{-nt} \label{eqn:Inactivation}
+\end{equation}
+
+(Eq. \ref{eqn:Transfer_sh})
+\begin{equation}
+C_{f_f} = C(t) TE_{sh} \label{eqn:Transfer_sh}
+\end{equation}
+
+(Eq. \ref{eqn:Dose})
+\begin{equation}
+D = C_{h} S_{hm} TE_{hm} \label{eqn:Dose}
+\end{equation}
+
+(Eq. \ref{eqn:DoseResponse})
+\begin{equation}
+P_{inf}=1 - e^{-kD} \label{eqn:DoseResponse}
+\end{equation}
+
 # Results and discussion
 
-Frequently touched surfaces such as traffic light and train buttons are of risk concern due to their frequent use by multiple people.
+Frequently touched surfaces such as and train buttons and  traffic light buttons are of risk concern due to their frequent use by multiple people.
 SARS-CoV-2 contamination has been reported in public surfaces of an urban area in Brazil.
-Concentration of viruses [genome units /m2] was between 20 and 2990.
-The surfaces were bench, ground, handrail made out metal and concrete [@Abrahao2020].
+The concentration of viruses was between 20 and 2990  genome units per m$^2$.
+The surfaces were bench, ground, handrail made out metal, and concrete [@Abrahao2020].
+In this article, we estimated the risk of contacting frequently touched surfaces as a function of the prevalence of infection within the population and evaluated the reduction of risk based on different disinfection methods.
 
-## Surface disinfection
-
-XX% reduction
+“ How big the droplets need to be and how far away they can travel is still an open question"
 
@@
 
-# Materials and Methods
-
-We modeled the risk of SARS-CoV-2 infection due to contact with frequently touched buttons and estimated the effect that different surface disinfection practices have on infection risk.
-The sequence of events leading to the viral inoculation of the susceptible individual is considered to be the following: 1) contamination event (infected person coughing on the hand followed by a hand-to-surface contact), 2) virus decay on the surface, 3) virus transfer from the surface to finger, and 4) virus transfer from finger to facial mucous membranes.
-The stochastic-mechanistic model used herein is based on models described elsewhere [@Julian2018; @Julian2009].
-Two scenarios were considered: risk of infection with and without surface disinfection.
-Additionally, different disinfection regimes were tested.
-
-Viral loads in the saliva or sputum of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input.
-The viral loads were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020].
-PCR data does not distinguish between infectious and non-infectious virus.
-To convert gene copies to infections virus, we used a ratio reported for influenza viruses [@Ip2016].
-
-The proportion of viral particles emitted in a cough that is transferred to the hand when the hand is used to cover the mouth was estimated using a conical dispersion model  [@Nicas2006] -  the particles were assumed to be conically spread over a 80 angle, at a distance of 5cm.
-The transfer of viruses from the hand of the infected individual to the surface was estimated with: $Csurface=Chand*TEsh$, where $Chand$ is the concentration of viruses on the hand $[virus/cm^2]$, and TEsh $[%]$ is the transfer efficiency of viruses between the surface (assumed to be metal) and the hand.
-Once the surface was contaminated, the concentration of virus in the surface at any given time was calculated as follows $Csurf(t)= Csurf(0)$ $e^{-nt}$, where $Csurf(t)$ $[virus/cm^2]$ is the concentration of infective viruses at time $t$, $Csurf(0)$ $[virus/cm2^]$ is the initial concentration of virus in the surface, $t$ $[min]$ is the time lapsed after surface inoculation, and $n$ $[m^{-1}]$ is the decay rate of the virus in the selected surface (metal or plastic).
-
-(Eq. \ref{eqn:example}).
-\begin{equation}
-  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
-\end{equation}
-
-\begin{equation}
-  \begin{minipage}[c]{0.80\linewidth}
-    \centering
-    As illustrated here, the width of \\
-    the minipage needs to allow some  \\
-    space for the number to fit in to.
-    %\includegraphics{graphic}
-  \end{minipage}
-  \label{eqn:graphic}
-\end{equation}
-
-# Experimental
-
 \begin{acknowledgement}
-
 The author thanks Sunil Dogga, Emmanuel Frousteyand, and Danilo Cuccato for their help with the model. 
-
 \end{acknowledgement}
@@
 \begin{suppinfo}
-
 Experimental procedures and
@@
 automatically add a sentence pointing to the information on-line:
+\end{suppinfo}
 
-\end{suppinfo}
 ## References

### 2020-09-11 ES&T paper/Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands.tex
Updating the code

--- 2020-07-30 Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands/Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands.tex
@@
 
+\textbf{???} QMRA to evaluate the reduction in the risk of infection of
+SARS-CoV-2 for mask wearers.
+
 In this study, a mechanistic model of indirect transmission within the
@@
 
+\hypertarget{materials-and-methods}{%
+\section{Materials and Methods}\label{materials-and-methods}}
+
+We modeled the risk of SARS-CoV-2 infection due to contact with
+frequently touched buttons and estimated the effect that different
+surface disinfection practices have on infection risk. The sequence of
+events leading to the viral inoculation of the susceptible individual is
+considered to be the following: 1) contamination event (infected person
+coughing on the hand followed by a hand-to-surface contact), 2) virus
+decay on the surface, 3) virus transfer from the surface to finger, and
+4) virus transfer from finger to facial mucous membranes. The
+stochastic-mechanistic model used herein is based on models described
+elsewhere.\textsuperscript{17,22} Two scenarios were considered: risk of
+infection with and without surface disinfection. Additionally, different
+disinfection regimes were tested.
+
+Viral loads in the saliva or sputum of symptomatic COVID-19 patients
+within the first 14 days of symptom onset were used as model input. The
+viral loads were measured using quantitative reverse-transcription
+polymerase chain reaction (RT-qPCR).\textsuperscript{23--26} PCR data
+does not distinguish between infectious and non-infectious virus. To
+convert gene copies to infections virus, we used a ratio reported for
+influenza viruses.\textsuperscript{27}
+
+The proportion of viral particles emitted in a cough that is transferred
+to the hand when the hand is used to cover the mouth was estimated using
+a conical dispersion model\textsuperscript{28} - the particles were
+assumed to be conically spread over a 80 angle, at a distance of 5cm.
+The transfer of viruses from the hand of the infected individual to the
+surface was estimated with: \(Csurface=Chand*TEsh\), where \(Chand\) is
+the concentration of viruses on the hand \([virus/cm^2]\), and TEsh
+\([%]
+\) is the transfer efficiency of viruses between the surface (assumed to
+be metal) and the hand. Once the surface was contaminated, the
+concentration of virus in the surface at any given time was calculated
+as follows \(Csurf(t)= Csurf(0)\) \(e^{-nt}\), where \(Csurf(t)\)
+\([virus/cm^2]\) is the concentration of infective viruses at time
+\(t\), \(Csurf(0)\) \([virus/cm2^]\) is the initial concentration of
+virus in the surface, \(t\) \([min]\) is the time lapsed after surface
+inoculation, and \(n\) \([m^{-1}]\) is the decay rate of the virus in
+the selected surface (metal or plastic).
+
+\hypertarget{alternative-way-of-presenting-the-equations}{%
+\subsection{Alternative way of presenting the
+equations}\label{alternative-way-of-presenting-the-equations}}
+
+The equations can be presented as follows:
+
+(Eq. \ref{eqn:Inactivation}) \begin{equation}
+C(t)= C(0) e^{-nt} \label{eqn:Inactivation}
+\end{equation}
+
+(Eq. \ref{eqn:Transfer_sh}) \begin{equation}
+C_{f_f} = C(t) TE_{sh} \label{eqn:Transfer_sh}
+\end{equation}
+
+(Eq. \ref{eqn:Dose}) \begin{equation}
+D = C_{h} S_{hm} TE_{hm} \label{eqn:Dose}
+\end{equation}
+
+(Eq. \ref{eqn:DoseResponse}) \begin{equation}
+P_{inf}=1 - e^{-kD} \label{eqn:DoseResponse}
+\end{equation}
+
 \hypertarget{results-and-discussion}{%
@@
 
-Frequently touched surfaces such as traffic light and train buttons are
-of risk concern due to their frequent use by multiple people. SARS-CoV-2
-contamination has been reported in public surfaces of an urban area in
-Brazil. Concentration of viruses {[}genome units /m2{]} was between 20
-and 2990. The surfaces were bench, ground, handrail made out metal and
-concrete.\textsuperscript{22}
+Frequently touched surfaces such as and train buttons and traffic light
+buttons are of risk concern due to their frequent use by multiple
+people. SARS-CoV-2 contamination has been reported in public surfaces of
+an urban area in Brazil. The concentration of viruses was between 20 and
+2990 genome units per m\(^2\). The surfaces were bench, ground, handrail
+made out metal, and concrete.\textsuperscript{29} In this article, we
+estimated the risk of contacting frequently touched surfaces as a
+function of the prevalence of infection within the population and
+evaluated the reduction of risk based on different disinfection methods.
 
-\hypertarget{surface-disinfection}{%
-\subsection{Surface disinfection}\label{surface-disinfection}}
-
-XX\% reduction
+`` How big the droplets need to be and how far away they can travel is
+still an open question"
 
@@
 
-\hypertarget{maths}{%
-\subsection{Math(s)}\label{maths}}
-
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
-
-We modeled the risk of SARS-CoV-2 infection due to contact with
-frequently touched buttons and estimated the effect that different
-surface disinfection practices have on infection risk. The sequence of
-events leading to the viral inoculation of the susceptible individual is
-considered to be the following: 1) contamination event (infected person
-coughing on the hand followed by a hand-to-surface contact), 2) virus
-decay on the surface, 3) virus transfer from the surface to finger, and
-4) virus transfer from finger to facial mucous membranes. The
-stochastic-mechanistic model used herein is based on models described
-elsewhere.\textsuperscript{17,23} Two scenarios were considered: risk of
-infection with and without surface disinfection. Additionally, different
-disinfection regimes were tested.
-
-Viral loads in the saliva or sputum of symptomatic COVID-19 patients
-within the first 14 days of symptom onset were used as model input. The
-viral loads were measured using quantitative reverse-transcription
-polymerase chain reaction (RT-qPCR).\textsuperscript{24--27} PCR data
-does not distinguish between infectious and non-infectious virus. To
-convert gene copies to infections virus, we used a ratio reported for
-influenza viruses.\textsuperscript{28}
-
-The proportion of viral particles emitted in a cough that is transferred
-to the hand when the hand is used to cover the mouth was estimated using
-a conical dispersion model\textsuperscript{{\textbf{???}}} - the
-particles were assumed to be conically spread over a 80 angle, at a
-distance of 5cm. The transfer of viruses from the hand of the infected
-individual to the surface was estimated with: \(Csurface=Chand*TEsh\),
-where \(Chand\) is the concentration of viruses on the hand
-\([virus/cm^2]\), and TEsh \([%]
-\) is the transfer efficiency of viruses between the surface (assumed to
-be metal) and the hand. Once the surface was contaminated, the
-concentration of virus in the surface at any given time was calculated
-as follows \(Csurf(t)= Csurf(0)\) \(e^{-nt}\), where \(Csurf(t)\)
-\([virus/cm^2]\) is the concentration of infective viruses at time
-\(t\), \(Csurf(0)\) \([virus/cm2^]\) is the initial concentration of
-virus in the surface, \(t\) \([min]\) is the time lapsed after surface
-inoculation, and \(n\) \([m^{-1}]\) is the decay rate of the virus in
-the selected surface (metal or plastic).
-
-(Eq. \ref{eqn:example}). \begin{equation}
-  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
-\end{equation}
-
-\begin{equation}
-  \begin{minipage}[c]{0.80\linewidth}
-    \centering
-    As illustrated here, the width of \\
-    the minipage needs to allow some  \\
-    space for the number to fit in to.
-    %\includegraphics{graphic}
-  \end{minipage}
-  \label{eqn:graphic}
-\end{equation}
-
-\hypertarget{experimental}{%
-\section{Experimental}\label{experimental}}
-
 \begin{acknowledgement}
-
 The author thanks Sunil Dogga, Emmanuel Frousteyand, and Danilo Cuccato for their help with the model. 
-
 \end{acknowledgement}
@@
 automatically add a sentence pointing to the information on-line:
-
 \end{suppinfo}
@@
 
+\leavevmode\hypertarget{ref-Julian2018}{}%
+(22) Julian, T. R.; Vithanage, H. S. K.; Chua, M. L.; Kuroda, M.; Pitol,
+A. K.; Nguyen, P. H. L.; Canales, R. A.; Fujii, S.; Harada, H.
+\emph{Science of The Total Environment} \textbf{2018}, \emph{635},
+120--131.
+
+\leavevmode\hypertarget{ref-Wolfel2020}{}%
+(23) Wölfel, R.; Corman, V. M.; Guggemos, W.; Seilmaier, M.; Zange, S.;
+Müller, M. A.; Niemeyer, D.; Jones, T. C.; Vollmar, P.; Rothe, C.;
+Hoelscher, M.; Bleicker, T.; Brünink, S.; Schneider, J.; Ehmann, R.;
+Zwirglmaier, K.; Drosten, C.; Wendtner, C. \emph{Nature} \textbf{2020},
+1--14.
+
+\leavevmode\hypertarget{ref-Pan2020}{}%
+(24) Pan, Y.; Zhang, D.; Yang, P.; Poon, L. L. M.; Wang, Q. \emph{The
+Lancet Infectious Diseases} \textbf{2020}, \emph{20} (4), 411--412.
+
+\leavevmode\hypertarget{ref-Kim2020}{}%
+(25) Kim, J. Y.; Ko, J. H.; Kim, Y.; Kim, Y. J.; Kim, J. M.; Chung, Y.
+S.; Kim, H. M.; Han, M. G.; Kim, S. Y.; Chin, B. S. \emph{Journal of
+Korean Medical Science} \textbf{2020}, \emph{35} (7), 1--7.
+
+\leavevmode\hypertarget{ref-To2020}{}%
+(26) To, K. K. W.; Tsang, O. T. Y.; Leung, W. S.; Tam, A. R.; Wu, T. C.;
+Lung, D. C.; Yip, C. C. Y.; Cai, J. P.; Chan, J. M. C.; Chik, T. S. H.;
+Lau, D. P. L.; Choi, C. Y. C.; Chen, L. L.; Chan, W. M.; Chan, K. H.;
+Ip, J. D.; Ng, A. C. K.; Poon, R. W. S.; Luo, C. T.; Cheng, V. C. C.;
+Chan, J. F. W.; Hung, I. F. N.; Chen, Z.; Chen, H.; Yuen, K. Y.
+\emph{The Lancet Infectious Diseases} \textbf{2020}, \emph{20} (5),
+565--574.
+
+\leavevmode\hypertarget{ref-Ip2016}{}%
+(27) Ip, D. K. M.; Lau, L. L. H.; Chan, K. H.; Fang, V. J.; Leung, G.
+M.; Peiris, M. J. S.; Cowling, B. J. \emph{Clinical Infectious Diseases}
+\textbf{2015}, \emph{62} (4), 431--437.
+
+\leavevmode\hypertarget{ref-Nicas2009}{}%
+(28) Nicas, M.; Jones, R. M. \emph{Risk Analysis} \textbf{2009},
+\emph{29} (9), 1292--1303.
+
 \leavevmode\hypertarget{ref-Abrahao2020}{}%
-(22) Abrahao, J. S.; Pengo, L. S.; Rezende, I. M.; Rodrigues, R.;
+(29) Abrahao, J. S.; Pengo, L. S.; Rezende, I. M.; Rodrigues, R.;
 Crispim, A. P. C.; Moura, C. C.; Mendonca, D. C.; Reis, E.; Souza, F.;
@@
 Domingos, S.; Boratto, P.; Henrique, P. \emph{medRxiv} \textbf{2020}.
-
-\leavevmode\hypertarget{ref-Julian2018}{}%
-(23) Julian, T. R.; Vithanage, H. S. K.; Chua, M. L.; Kuroda, M.; Pitol,
-A. K.; Nguyen, P. H. L.; Canales, R. A.; Fujii, S.; Harada, H.
-\emph{Science of The Total Environment} \textbf{2018}, \emph{635},
-120--131.
-
-\leavevmode\hypertarget{ref-Wolfel2020}{}%
-(24) Wölfel, R.; Corman, V. M.; Guggemos, W.; Seilmaier, M.; Zange, S.;
-Müller, M. A.; Niemeyer, D.; Jones, T. C.; Vollmar, P.; Rothe, C.;
-Hoelscher, M.; Bleicker, T.; Brünink, S.; Schneider, J.; Ehmann, R.;
-Zwirglmaier, K.; Drosten, C.; Wendtner, C. \emph{Nature} \textbf{2020},
-1--14.
-
-\leavevmode\hypertarget{ref-Pan2020}{}%
-(25) Pan, Y.; Zhang, D.; Yang, P.; Poon, L. L. M.; Wang, Q. \emph{The
-Lancet Infectious Diseases} \textbf{2020}, \emph{20} (4), 411--412.
-
-\leavevmode\hypertarget{ref-Kim2020}{}%
-(26) Kim, J. Y.; Ko, J. H.; Kim, Y.; Kim, Y. J.; Kim, J. M.; Chung, Y.
-S.; Kim, H. M.; Han, M. G.; Kim, S. Y.; Chin, B. S. \emph{Journal of
-Korean Medical Science} \textbf{2020}, \emph{35} (7), 1--7.
-
-\leavevmode\hypertarget{ref-To2020}{}%
-(27) To, K. K. W.; Tsang, O. T. Y.; Leung, W. S.; Tam, A. R.; Wu, T. C.;
-Lung, D. C.; Yip, C. C. Y.; Cai, J. P.; Chan, J. M. C.; Chik, T. S. H.;
-Lau, D. P. L.; Choi, C. Y. C.; Chen, L. L.; Chan, W. M.; Chan, K. H.;
-Ip, J. D.; Ng, A. C. K.; Poon, R. W. S.; Luo, C. T.; Cheng, V. C. C.;
-Chan, J. F. W.; Hung, I. F. N.; Chen, Z.; Chen, H.; Yuen, K. Y.
-\emph{The Lancet Infectious Diseases} \textbf{2020}, \emph{20} (5),
-565--574.
-
-\leavevmode\hypertarget{ref-Ip2016}{}%
-(28) Ip, D. K. M.; Lau, L. L. H.; Chan, K. H.; Fang, V. J.; Leung, G.
-M.; Peiris, M. J. S.; Cowling, B. J. \emph{Clinical Infectious Diseases}
-\textbf{2015}, \emph{62} (4), 431--437.
 \end{document}

### 2020-09-11 ES&T paper/Supporting_Information.Rmd
Updating the code

---
journal: jacsat
type: article
author:
  - name: Ana K. Pitol
    aff: Department of Civil and Environmental Engineering, Imperial College London, United Kingdom
    email: \email{a.pitol-garcia@imperial.ac.uk}

  - name: Timothy R. Julian
    aff: Department of Environmental Microbiology, Eawag, Dubendorf, Switzerland
    alsoaff: Department of Epidemiology and Public Health, Swiss Tropical and Public Health Institute, Basel, Switzerland
    email: \email{tim.julian@eawag.ch}

title:
  formatted: Supporting Information. Risk reduction of SARS-CoV-2 transmission through disinfection of frequently touched surfaces. A Quantitative Microbial Risk Assessment. 
  short: An \textsf{achemso} demo
abbr: IR,NMR,UV
keywords: American Chemical Society, \LaTeX
bibliography: acstest.bib
csl: american-chemical-society.csl
output: rticles::acs_article
includes:
      in_header: preamble.tex
---

## Materials and Methods

### Concentration of virus in the saliva and sputum
Viral loads in the saliva, sputum, or uper respiratory track of symptomatic COVID-19 patients within the first 14 days of symptom onset were used as model input. The viral loads collected from litterature. The number of viruses in the samples were measured using quantitative reverse-transcription polymerase chain reaction (RT-qPCR)  [@Wolfel2020; @Pan2020; @Kim2020; @To2020]. The data were extracted from the figures and is meant to be an approximated number.

PCR data does not distinguish between infectious and non-infectious virus. No reliable information was found regarding the relationship between gene copy number and infectivity for SARS-CoV-2 [@Jefferson2020]. Therefore, to convert gene copies to infections virus, we used a ratio reported for influenza viruses [@Ip2016].

### Transfer efficiency of virus from hand to mouth
Data on transfer efficiency of a dry inoculum of bacteriophage MS2 from hand to saliva [@Pitol2017] was used as a proxy for transfer efficiency from hand to mucouse membranes. The data used is available at <https://zenodo.org/record/1454505#.Xo3wMZNKjBI>.

### Virus dacay on surface
Modelled as a normal distribution. 
Data obtained from the Suplementary Appendix of [@vanDoremalen2020]. Parameters for steel, using median as mean, and assuming the standard deviation based on the reported confidence intervals. 
The decay rate was calculated as follows: 
$n_x=\frac {ln(2)}{t_\frac{1}{2}}$

### Dose-response 
QMRAwiki, using the from 0.5th, 50th, and 99.5 th percentiles as min, mode, and max of a triangular distribution. 
Data obtained from QMRAwiki http://qmrawiki.org/experiments/sars, Parameter (k) for the exponential model, based on the 2 articles, using 10,000 bootstrap iterations (MLE).
Parameter developed from 2 different studies using 2 different coronaviruses: SARS-CoV and Murine hepatitis virus (MHV-1)
 @DeAlbuquerque2006. A/J mice with MHV-1 virus, intranasal, survival monitored for 21 days. MHV is a common surrogate of human coronaviruses

 @DeDiego2008. 4 groups of the transgenic mice, intranasal, graded doses of rSARS-CoV, survival monitored for 13 days

@Watanabe2010. Combining both dose-response articles to develop the dose-response

### Finger surface area in contact with mouth
Method used by Julian et. all 2018. @Julian2018
Fractional surface area taken from @AuYeung2008 (adult front partial finger)
Average hand surface area taken from EPA 2011 @EPA2011

### Prevalence data 
Data obtained from antibody surveys in different countries. The data is meant to be an estimate of the order of magnitude of the prevalence expected in the population. Estimating prevalence at any given time assuming each person is able to infect for a period of 14 days, Prev [%] =  Seroconversion [%] / (outbreak length [t] /duration of infection [t]). 
Data taken from [@Bendavid2020, @Perez2020, @Pollan2020, @Erikstrup2020, @Filho2020]. Prevalence is between 0.4 (Netherlands) and 1.9 (Geneva). Based on this data, three prevalence scenarios were suggested (0.2,1 and 5%).

### Surface disinfection
Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol. 
Article             virus (enveloped)     log10 Red     Sanitizer
 __________________  _____________________  __________  ________________________
@Hulkower2011         TGEV (coronavirus)     3.2        Ethanol (70%)
                       MHV (coronavirus)      3.9        Ethanol (70%)
                       TGEV (coronavirus)     0.6        Chlorine bleach (0.06%)  *Concentration too low*
                       MHV (coronavirus)      0.4        Chlorine bleach (0.06%)  *Concentration too low*
 @Sattar1989           HCV-229E(human corona)  >3        Chlorine (0.1%)
                       HCV-229E(human corona)  >3        Chlorine (0.5%)

 surf_dis <- runif(simNum, 10^3, 10^4)

### Hand disinfection 
Regarding Hand Hygene. It is recomended by the WHO that [@WHO2010_HandHygene]: An effective alcohol-based hand rub product should contain between 60% and 80% of alcohol.
Log reduction = 4.25 [@Rabenau2005].

### Reduction based on mask 
Method: Henderson apparatus, collison nebulizer (air particles, no droplets)
% Filtration Efficiency (fraction of total virus filtered out by the material) for bacteriophage MS2   
     Min     Material    --   Max    Material
     48.87   Scarf            89.52  Surgical Mask

## References

### 2020-09-11 ES&T paper/Supporting_Information.tex
Updating the code

--- 2020-07-30 Risk reduction of SARS-CoV-2 by disinfection of surfaces and hands/Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands.tex
@@
 
-\title[An \textsf{achemso} demo]{Risk reduction of SARS-CoV-2 transmission through disinfection of
-frequently touched surfaces. A Quantitative Microbial Risk Assessment.}
+\title[An \textsf{achemso} demo]{Supporting Information. Risk reduction of SARS-CoV-2 transmission
+through disinfection of frequently touched surfaces. A Quantitative
+Microbial Risk Assessment.}
 \makeatletter
@@
 \begin{abstract}
-not exceeding 200 words A graphic for the Table of Contents (TOC) must
-be supplied with each Letter This graphic can be no wider than 8 cm and
-no taller than 4 cm and should be prepared following the specifications
-under Artwork.
+
 \end{abstract}
-\begin{tocentry}
-Add image (TOC)
-\end{tocentry}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-SARS-CoV-2, the virus responsible for the COVID-19 pandemic, is
-transmitted via both direct (person-to-person) and indirect (via a
-contaminated environment) routes.\textsuperscript{1,2} Direct
-transmission appears to be the leading route and occurs via prolonged
-exposures to respiratory droplets produced while talking, coughing, and
-sneezing. Infection control recommendations, based on the assumption
-that direct contact transmission is the leading route, include
-maintaining social/physical distances, wearing masks, case isolation,
-contact tracing, and quarantine.\textsuperscript{3}
-
-Despite knowledge that SARS-CoV-2 transmission is primarily through
-direct transmission, indirect transmission - or transmission occurring
-due to interaction with a contaminated environment -- remains possible.
-People infected with SARS-CoV-2 shed the virus into the environment, as
-evidenced by extensive environmental contamination detected on surfaces
-in cruise ships, hospitals, and public spaces in urban areas such as bus
-stations and public squares {[}4; 5; Abrahao2020{]}. Infective
-coronavirus persists in the environment, with experimental evidence of
-persistence on surfaces for prolonged periods up to 9
-days.\textsuperscript{6} Viruses readily transfer from contaminated
-surfaces to the skin upon contact\textsuperscript{7--11} and from the
-skin to the mouth and saliva of individuals.\textsuperscript{10,12}
-Taken together, this evidence suggests surface contamination poses a
-risk for indirect SARS-CoV-2 transmission, similar to other respiratory
-viruses.\textsuperscript{13}
-
-Indirect transmission events of SARS-CoV-2, even if not the dominant
-transmission route, complicate infection control efforts using contact
-tracing. Contact tracing efforts focus on identifying people sharing
-spaces coincidentally with cases.\textsuperscript{3} Indirect
-transmission could occur in the absence of shared spaces. As an example,
-transmission of SARS-CoV-1 in a Taiwanese hospital led to nosocomial
-infection of 31 patients. In 6 (20\%) of the patients, contact tracing
-failed to detect direct contacts with other SARS patients, although RNA
-of SARS-CoV-1 was detected throughout the hospital on drinking water
-buttons, beds, and chairs.\textsuperscript{14} Based on these findings,
-the authors suggest indirect transmission as the cause of transmission
-amongst a subset of those infected.\textsuperscript{14} If indirect
-transmission is an important route for SARS-CoV-2, control strategies
-will necessarily have to integrate hand hygiene interventions and
-surface disinfection alongside contact tracing.\textsuperscript{15}
-
-Despite the potential importance of indirect transmission, it is
-difficult to estimate its importance relative to direct
-transmission.\textsuperscript{16} Quantitative Microbial Risk Analysis
-(QMRA) provides a framework for understanding health risks from indirect
-transmission and provides insights into potential impacts of infection
-control recommendations. Mechanistic models of fomite-mediated
-transmission events within the context of QMRA frameworks have been used
-to inform risks for children interacting with contaminated
-toys,\textsuperscript{17} sanitation workers collecting and processing
-urine for nutrient recovery,\textsuperscript{18} transmission of
-norovirus within a houseboat,\textsuperscript{19} and impact of surface
-disinfection.\textsuperscript{20} QMRA has also been adapted to evaluate
-risks for hospital transmission of MERS-CoV through droplets and
-aerosolized particles. The analysis highlighted reductions in risk to
-hospital staff through mask use (\textgreater90\% estimated risk
-reduction) and increased air exchange (up to 58\% estimated risk
-reduction).\textsuperscript{21}
-
-In this study, a mechanistic model of indirect transmission within the
-QMRA framework is developed to estimate likelihood of transmission in
-community settings and inform guidance on surface disinfection
-strategies. Specifically, the risks of infection with SARS-CoV-2 are
-estimated for interactions with shared spaces (i.e., use of ATMs,
-buttons on cross walks and trains). Risk reductions are further
-estimated under various feasible surface disinfection strategies.
-
-\hypertarget{results-and-discussion}{%
-\section{Results and discussion}\label{results-and-discussion}}
-
-Frequently touched surfaces such as traffic light and train buttons are
-of risk concern due to their frequent use by multiple people. SARS-CoV-2
-contamination has been reported in public surfaces of an urban area in
-Brazil. Concentration of viruses {[}genome units /m2{]} was between 20
-and 2990. The surfaces were bench, ground, handrail made out metal and
-concrete.\textsuperscript{22}
-
-\hypertarget{surface-disinfection}{%
-\subsection{Surface disinfection}\label{surface-disinfection}}
-
-XX\% reduction
-
-\hypertarget{floats}{%
-\subsection{Floats}\label{floats}}
-
-New float types are automatically set up by the class file. The means
-graphics are included as follows (Scheme \ref{sch:example}). As
-illustrated, the float is ``here'' if possible.
-
-\begin{scheme}
-  Your scheme graphic would go here: \texttt{.eps} format\\
-  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
-  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
-  these can be scaled without loss of quality, and can be\\
-  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
-  %\includegraphics{graphic}
-  \caption{An example scheme}
-  \label{sch:example}
-\end{scheme}
-
-\begin{figure}
-\centering
-\includegraphics{Risk-reduction-of-SARS-CoV-2-by-disinfection-of-surfaces-and-hands_files/figure-latex/unnamed-chunk-1-1.pdf}
-\caption{test}
-\end{figure}
-
-\begin{figure}
-  As well as the standard float types \texttt{table}\\
-  and \texttt{figure}, the class also recognises\\
-  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
-  \caption{An example figure}
-  \label{fgr:example}
-\end{figure}
-
-Charts, figures and schemes do not necessarily have to be labelled or
-captioned. However, tables should always have a title. It is possible to
-include a number and label for a graphic without any title, using an
-empty argument to the \texttt{\textbackslash caption} macro.
-
-The use of the different floating environments is not required, but it
-is intended to make document preparation easier for authors. In general,
-you should place your graphics where they make logical sense; the
-production process will move them if needed.
-
-\hypertarget{maths}{%
-\subsection{Math(s)}\label{maths}}
-
 \hypertarget{materials-and-methods}{%
@@
 
-We modeled the risk of SARS-CoV-2 infection due to contact with
-frequently touched buttons and estimated the effect that different
-surface disinfection practices have on infection risk. The sequence of
-events leading to the viral inoculation of the susceptible individual is
-considered to be the following: 1) contamination event (infected person
-coughing on the hand followed by a hand-to-surface contact), 2) virus
-decay on the surface, 3) virus transfer from the surface to finger, and
-4) virus transfer from finger to facial mucous membranes. The
-stochastic-mechanistic model used herein is based on models described
-elsewhere.\textsuperscript{17,23} Two scenarios were considered: risk of
-infection with and without surface disinfection. Additionally, different
-disinfection regimes were tested.
+\hypertarget{concentration-of-virus-in-the-saliva-and-sputum}{%
+\subsection{Concentration of virus in the saliva and
+sputum}\label{concentration-of-virus-in-the-saliva-and-sputum}}
 
@@
 viral loads were measured using quantitative reverse-transcription
-polymerase chain reaction (RT-qPCR).\textsuperscript{24--27} PCR data
-does not distinguish between infectious and non-infectious virus. To
-convert gene copies to infections virus, we used a ratio reported for
-influenza viruses.\textsuperscript{28}
+polymerase chain reaction (RT-qPCR).\textsuperscript{1--4} The data were
+extracted from the figures and is meant to be an approximate number.
 
-The proportion of viral particles emitted in a cough that is transferred
-to the hand when the hand is used to cover the mouth was estimated using
-a conical dispersion model\textsuperscript{{\textbf{???}}} - the
-particles were assumed to be conically spread over a 80 angle, at a
-distance of 5cm. The transfer of viruses from the hand of the infected
-individual to the surface was estimated with: \(Csurface=Chand*TEsh\),
-where \(Chand\) is the concentration of viruses on the hand
-\([virus/cm^2]\), and TEsh \([%]
-\) is the transfer efficiency of viruses between the surface (assumed to
-be metal) and the hand. Once the surface was contaminated, the
-concentration of virus in the surface at any given time was calculated
-as follows \(Csurf(t)= Csurf(0)\) \(e^{-nt}\), where \(Csurf(t)\)
-\([virus/cm^2]\) is the concentration of infective viruses at time
-\(t\), \(Csurf(0)\) \([virus/cm2^]\) is the initial concentration of
-virus in the surface, \(t\) \([min]\) is the time lapsed after surface
-inoculation, and \(n\) \([m^{-1}]\) is the decay rate of the virus in
-the selected surface (metal or plastic).
-
-(Eq. \ref{eqn:example}). \begin{equation}
-  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
-\end{equation}
-
-\begin{equation}
-  \begin{minipage}[c]{0.80\linewidth}
-    \centering
-    As illustrated here, the width of \\
-    the minipage needs to allow some  \\
-    space for the number to fit in to.
-    %\includegraphics{graphic}
-  \end{minipage}
-  \label{eqn:graphic}
-\end{equation}
-
-\hypertarget{experimental}{%
-\section{Experimental}\label{experimental}}
-
-\begin{acknowledgement}
-
-The author thanks Sunil Dogga, Emmanuel Frousteyand, and Danilo Cuccato for their help with the model.
-
-\end{acknowledgement}
-
-\begin{suppinfo}
-
-Experimental procedures and
-characterization data for all new compounds. The class will
-automatically add a sentence pointing to the information on-line:
-
-\end{suppinfo}
+PCR data does not distinguish between infectious and non-infectious
+virus. No reliable information was found regarding the relationship
+between gene copy number and infectivity for
+SARS-CoV-2.\textsuperscript{5} Therefore, to convert gene copies to
+infections virus, we used a ratio reported for influenza
+viruses.\textsuperscript{6}
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-WHO2020}{}%
-(1) WHO. \emph{Scientific brief} \textbf{2020}, \emph{29} (March),
-10--12.
-
-\leavevmode\hypertarget{ref-CDC2020}{}%
-(2) CDC. \emph{How COVID-19 Spreads}; 2020; Vol. 2019, p 2020.
-
-\leavevmode\hypertarget{ref-Ferretti2020}{}%
-(3) Ferretti, L.; Wymant, C.; Kendall, M.; Zhao, L.; Nurtay, A.;
-Abeler-Dörner, L.; Parker, M.; Bonsall, D.; Fraser, C. \emph{Science}
-\textbf{2020}, \emph{368} (6491), 0--8.
-
-\leavevmode\hypertarget{ref-Ong2020}{}%
-(4) Ong, S. W. X.; Tan, Y. K.; Chia, P. Y.; Lee, T. H.; Ng, O. T.; Wong,
-M. S. Y.; Marimuthu, K. \emph{JAMA - Journal of the American Medical
-Association} \textbf{2020}, 2--4.
-
-\leavevmode\hypertarget{ref-Ye2020}{}%
-(5) Ye, G.; Lin, H.; Chen, L.; Wang, S.; Zeng, Z.; Wang, W.; Zhang, S.;
-Rebmann, T.; Li, Y.; Pan, Z.; Yang, Z.; Wang, Y.; Wang, F.; Qian, Z.;
-Wan, X. \emph{medRxiv preprint} \textbf{2020}, 1--20.
-
-\leavevmode\hypertarget{ref-Kampf2020}{}%
-(6) Kampf, G.; Todt, D.; Pfaender, S.; Steinmann, E. \emph{Journal of
-Hospital Infection} \textbf{2020}, \emph{104} (3), 246--251.
-
-\leavevmode\hypertarget{ref-Julian2010}{}%
-(7) Julian, T. R.; Leckie, J. O.; Boehm, A. B. \emph{Journal of Applied
-Microbiology} \textbf{2010}, \emph{109}, 1868--1874.
-
-\leavevmode\hypertarget{ref-Lopez2013}{}%
-(8) Lopez, G. U.; Gerba, C. P.; Tamimi, A. H.; Kitajima, M.; Maxwell, S.
-L.; Rose, J. B. \emph{Applied and Environmental Microbiology}
-\textbf{2013}, \emph{79}, 5728--5734.
-
-\leavevmode\hypertarget{ref-Bidawid2004}{}%
-(9) Bidawid, S.; Malik, N.; Adegbunrin, O.; Sattar, S. A.; Farber, J. M.
-\emph{J Food Prot} \textbf{2004}, \emph{67} (1), 103--109.
-
-\leavevmode\hypertarget{ref-Rusin2002}{}%
-(10) Rusin, P.; Maxwell, S.; Gerba, C. \emph{Journal of Applied
-Microbiology} \textbf{2002}, \emph{93}, 585--592.
-
-\leavevmode\hypertarget{ref-Ansari1991}{}%
-(11) Ansari, S.; Springthorpe, V.; Sattar, S.; Rivard, S.; Rahman, M.
-\emph{Journal of Clinical Microbiology} \textbf{1991}, \emph{29} (10),
-2115--2119.
-
-\leavevmode\hypertarget{ref-Pitol2017}{}%
-(12) Pitol, A. K.; Bischel, H. N.; Kohn, T.; Julian, T. R.
-\emph{Environmental Science and Technology} \textbf{2017}.
-
-\leavevmode\hypertarget{ref-Boone2007}{}%
-(13) Boone, S. A.; Gerba, C. P. \emph{Applied and Environmental
-Microbiology} \textbf{2007}, \emph{73} (6), 1687--1696.
-
-\leavevmode\hypertarget{ref-Chen2004}{}%
-(14) Chen, Y. C.; Huang, L. M.; Chan, C. C.; Su, C. P.; Chang, S. C.;
-Chang, Y. Y.; Chen, M. L.; Hung, C. C.; Chen, W. J.; Lin, F. Y.; Lee, Y.
-T.; Chen, D. S.; Lee, Y. T.; Teng, C. M.; Yang, P. C.; Ho, H. N.; Chen,
-P. J.; Chang, M. F.; Wang, J. T.; Kao, C. L.; Wang, W. K.; Hsiao, C. H.;
-Hsueh, P. R. \emph{Emerging Infectious Diseases} \textbf{2004},
-\emph{10} (5), 782--788.
-
-\leavevmode\hypertarget{ref-Mcclellan2020}{}%
-(15) Gottlieb, S.; Rivers, C.; Mcclellan, M. B.; Silvis, L.; Watson, C.
-\emph{American Enterprise Institute} \textbf{2020}, 1--16.
-
-\leavevmode\hypertarget{ref-BarOn2020}{}%
-(16) Bar-on, Y. M.; Flamholz, A.; Phillips, R.; Milo, R. \emph{eLife}
-\textbf{2020}.
-
-\leavevmode\hypertarget{ref-Julian2009}{}%
-(17) Julian, T. R.; Canales, R. a.; Leckie, J. O.; Boehm, A. B.
-\emph{Risk Analysis} \textbf{2009}, \emph{29}, 617--632.
-
-\leavevmode\hypertarget{ref-Bischel2019}{}%
-(18) Bischel, H. N.; Caduff, L.; Schindelholz, S.; Kohn, T.; Julian, T.
-R. \emph{Environmental Science and Technology} \textbf{2019}, \emph{53}
-(12), 7055--7067.
-
-\leavevmode\hypertarget{ref-Canales2019}{}%
-(19) Canales, R. A.; Reynolds, K. A.; Wilson, A. M.; Fankem, S. L. M.;
-Weir, M. H.; Rose, J. B.; Abd-Elmaksoud, S.; Gerba, C. P. \emph{Journal
-of Occupational and Environmental Hygiene} \textbf{2019}, \emph{16} (1),
-16--26.
-
-\leavevmode\hypertarget{ref-Ryan2014a}{}%
-(20) Ryan, M. O.; Haas, C. N.; Gurian, P. L.; Gerba, C. P.; Panzl, B.
-M.; Rose, J. B. \emph{American Journal of Infection Control}
-\textbf{2014}, \emph{42} (11), 1165--1172.
-
-\leavevmode\hypertarget{ref-Adhikari2019}{}%
-(21) Adhikari, U.; Chabrelie, A.; Weir, M.; Boehnke, K.; McKenzie, E.;
-Ikner, L.; Wang, M.; Wang, Q.; Young, K.; Haas, C. N.; Rose, J.;
-Mitchell, J. \emph{Risk Analysis} \textbf{2019}, \emph{39} (12),
-2608--2624.
-
-\leavevmode\hypertarget{ref-Abrahao2020}{}%
-(22) Abrahao, J. S.; Pengo, L. S.; Rezende, I. M.; Rodrigues, R.;
-Crispim, A. P. C.; Moura, C. C.; Mendonca, D. C.; Reis, E.; Souza, F.;
-Oliveira, G. F. G. G. P. G.; Domingos, I. J. da S.; Boratto, P.; Silva,
-P. H. B. e; Queiroz, V. F.; Silva, T. B. de S.; Oliveira, G. F. G. G. P.
-G.; Alves, V. de S.; Alves, P. A.; Kroon, E. G.; Trindade, G. de S.;
-Drumond, B. P.; Jônatas, A.; Abrahão, S.; Sacchetto, L.; Rezende, I. M.;
-Rodrigues, R.; Paula, A.; Crispim, C.; Moura, C. C.; Correa, D.; Reis,
-E.; Souza, F.; Fernanda, G.; Oliveira, G. F. G. G. P. G.; José, I.;
-Domingos, S.; Boratto, P.; Henrique, P. \emph{medRxiv} \textbf{2020}.
-
-\leavevmode\hypertarget{ref-Julian2018}{}%
-(23) Julian, T. R.; Vithanage, H. S. K.; Chua, M. L.; Kuroda, M.; Pitol,
-A. K.; Nguyen, P. H. L.; Canales, R. A.; Fujii, S.; Harada, H.
-\emph{Science of The Total Environment} \textbf{2018}, \emph{635},
-120--131.
-
 \leavevmode\hypertarget{ref-Wolfel2020}{}%
-(24) Wölfel, R.; Corman, V. M.; Guggemos, W.; Seilmaier, M.; Zange, S.;
+(1) Wölfel, R.; Corman, V. M.; Guggemos, W.; Seilmaier, M.; Zange, S.;
 Müller, M. A.; Niemeyer, D.; Jones, T. C.; Vollmar, P.; Rothe, C.;
@@
 \leavevmode\hypertarget{ref-Pan2020}{}%
-(25) Pan, Y.; Zhang, D.; Yang, P.; Poon, L. L. M.; Wang, Q. \emph{The
+(2) Pan, Y.; Zhang, D.; Yang, P.; Poon, L. L. M.; Wang, Q. \emph{The
 Lancet Infectious Diseases} \textbf{2020}, \emph{20} (4), 411--412.
@@
 \leavevmode\hypertarget{ref-Kim2020}{}%
-(26) Kim, J. Y.; Ko, J. H.; Kim, Y.; Kim, Y. J.; Kim, J. M.; Chung, Y.
+(3) Kim, J. Y.; Ko, J. H.; Kim, Y.; Kim, Y. J.; Kim, J. M.; Chung, Y.
 S.; Kim, H. M.; Han, M. G.; Kim, S. Y.; Chin, B. S. \emph{Journal of
@@
 \leavevmode\hypertarget{ref-To2020}{}%
-(27) To, K. K. W.; Tsang, O. T. Y.; Leung, W. S.; Tam, A. R.; Wu, T. C.;
+(4) To, K. K. W.; Tsang, O. T. Y.; Leung, W. S.; Tam, A. R.; Wu, T. C.;
 Lung, D. C.; Yip, C. C. Y.; Cai, J. P.; Chan, J. M. C.; Chik, T. S. H.;
@@
 
+\leavevmode\hypertarget{ref-Jefferson2020}{}%
+(5) Jefferson, T.; Spencer, E.; Brassey, J.; Heneghan, C. \emph{medRxiv}
+\textbf{2020}, 2020.08.04.20167932.
+
 \leavevmode\hypertarget{ref-Ip2016}{}%
-(28) Ip, D. K. M.; Lau, L. L. H.; Chan, K. H.; Fang, V. J.; Leung, G.
-M.; Peiris, M. J. S.; Cowling, B. J. \emph{Clinical Infectious Diseases}
+(6) Ip, D. K. M.; Lau, L. L. H.; Chan, K. H.; Fang, V. J.; Leung, G. M.;
+Peiris, M. J. S.; Cowling, B. J. \emph{Clinical Infectious Diseases}
 \textbf{2015}, \emph{62} (4), 431--437.

### 2020-09-11 Temporal_Analysis/More info to add.Rmd
Updating the code

--- 2020-06-04 More info to add.Rmd
@@
 
-The subjects touched their eyes, nostrils, and lops an average of 16(11) times an hour @Nicas2008
+droplets produced by an infected host during coughing
+
+Quantified using video recordings 10 subjects wotking on an office during 3 hr. The subjects touched their eyes, nostrils, and lops an average of 16(11) times an hour @Nicas2008.
 

### 2020-09-11 Temporal_Analysis/Paper.Rmd
Updating the code

--- 2020-06-12 Paper.Rmd
@@
 
+*ATM Article* High level of microbiological contamination has been detected in ATM machines. The study highligths ATM machines as potential areas for pathogen accumulation [@Allemailem2020].
+
+"The new coronavirus SARS-CoV-2 is less deadly but far more transmissible than MERS-CoV or SARS-CoV" [Petersen2020]
+
+"The median duration of virus in the respiratory samples of patients with severe disease (21 days, 14-30 days) was significantly longer than in patients with mild disease (14 days, 10-21 days; P=0.04)" [@Zheng2020]
+
 ## Supporting Information

### 2020-09-11 Temporal_Analysis/concentration_time.R

@@ -1,0 +1,114 @@
+         #   Simulating Concentrations Through Time  #
+
+# Conditions
+# Material:       Metal
+# Humidity:       High
+# Surface:        Buttons
+# Inoculation:    Cough on hands followed by hand to surface transfer
+# Disinfection:   YES - Surface
+
+
+# setup ------------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+library(truncnorm)
+library(EnvStats)
+
+# read and process data  -------------------------------------------------------
+source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020, To 2020 (first 2 weeks after symptom onset)
+source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity)
+source ("src/process_decay.R") # van Doremalen 2020
+source ("src/process_prevalence.R") # Data obtained from antibody surveys in different countries.
+
+# Simulation parameters:
+simNum = 1000 # Number of simulations
+
+# General parameters
+Vs <- runif(simNum, 0.0396, 0.0484) # [mL] Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
+Csp <- sample(df_Csp$Csp, size=simNum, replace = TRUE) # [copies/ml] Concentration of saliva in the first 2 weeks after symptom onsent 
+
+# Metal specific parameters
+k_stl <- rtruncnorm(simNum, 0, Inf, k_stl_mean, k_stl_sd) # [min] Half life of CoV-2 in steel
+n_stl <- log(2) / k_stl # [min-1] Decay rate, function of the halflife of the virus in the surface
+TEsh_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from surface to hand for stainless steel RH=[40-65%]
+TEhs_stl <- rtruncnorm(simNum, 0, 1, TE_sh_stl_mean, TE_sh_stl_sd) # TE from hand to surface for stainless steel RH=[40-65%]
+
+# Scenario specific parameters
+angle <- 0.8 # Fixed, based in the images of @2Bourouiba2014
+# prev <- runif(simNum, 0.002, 0.05) # For sensitivity analysis from 0.02 to 5% prevalence.
[…]

### 2020-09-11 src/Risk_Stl_Btn_HandDis.R
Update on the code, three risk models

@@ -3,7 +3,7 @@
 # Humidity:       High
 # Surface:        Buttons
 # Inoculation:    Cough on hands followed by hand to surface transfer
-# Intervention:   Surface Disinfection
+# Intervention:   Hand Disinfection
 
 
 # setup ------------------------------------------------------------------------
@@ -21,16 +21,10 @@ source ("src/process_doseresponse.R") # QMRAwiki, using the from 0.5th, 50th, an
 source ("src/process_Asf.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
 source ("src/process_prevalence.R") # Data obtained from antibody surveys in different countries.
 source ("src/process_disinfection.R") # Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol
-# source ("src/process_mask") # Reduction in inoculation on hands when mask is used
 
- 
 # Simulation parameters:
 simNum = 50000 # Number of simulations
 # ----- #
-
-# Paramaters (non used in this version of the model)
-# adj_DoseResp <- runif(simNum, 0.001, 0.01) # Adjustments for intranasal administration, @Jones2020
-# Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # [cough/min] Number of coughs per minute, Leung (2020)
 
 # General parameters
 GC_Inf <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
@@ -54,17 +48,20 @@ ratio <- x * tan (angle) # [cm] Ratio of circular plane on a cone at a distance 
 area_inoculation <- pi * ratio^2 # [cm2] circular area from a conical distribution at a distance x
 
 # Scenario specific parameters
-prev <- low_prev #  low_prev = 0.002, med_prev = 0.01, high_prev = 0.05, data justified in "process_prevalence.R"
[…]

### 2020-09-11 src/Risk_Stl_Btn_Mask.R

@@ -3,8 +3,7 @@
 # Humidity:       High
 # Surface:        Buttons
 # Inoculation:    Cough on hands followed by hand to surface transfer
-# Intervention:   Surface Disinfection
-
+# Intervention:   Face Mask
 
 # setup ------------------------------------------------------------------------
 # external libraries
@@ -21,9 +20,9 @@ source ("src/process_doseresponse.R") # QMRAwiki, using the from 0.5th, 50th, an
 source ("src/process_Asf.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
 source ("src/process_prevalence.R") # Data obtained from antibody surveys in different countries.
 source ("src/process_disinfection.R") # Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol
-# source ("src/process_mask") # Reduction in inoculation on hands when mask is used
+source ("src/process_mask.R") # Reduction in inoculation on hands when mask is used
 
- 
+
 # Simulation parameters:
 simNum = 50000 # Number of simulations
 # ----- #
@@ -54,12 +53,14 @@ ratio <- x * tan (angle) # [cm] Ratio of circular plane on a cone at a distance 
 area_inoculation <- pi * ratio^2 # [cm2] circular area from a conical distribution at a distance x
 
 # Scenario specific parameters
-prev <- low_prev #  low_prev = 0.002, med_prev = 0.01, high_prev = 0.05, data justified in "process_prevalence.R"
-t_dis <- 3 # Disinfection strategy ([0= No dis], [1=7am], [2=12pm], [3=7am and 12pm], [4=12 and 6pm])
+prev <- med_prev #  low_prev = 0.002, med_prev = 0.01, high_prev = 0.05, data justified in "process_prevalence.R"
+compliance <- 0  # use 0, 0.25, 0.50, 0.75
 
[…]

### 2020-09-11 src/eliminate/Risk_Hand_Dis.R

@@ -1,0 +1,173 @@
+# Conditions
+# Material:       Metal
+# Humidity:       High
+# Surface:        Buttons
+# Inoculation:    Cough on hands followed by hand to surface transfer
+# Disinfection:   YES - Surface
+
+
+# setup ------------------------------------------------------------------------
+# external libraries
+library(tidyverse) 
+library(truncnorm)
+library(EnvStats)
+
+
+# read and process data  -------------------------------------------------------
+source ("src/process_concentrationsaliva.R") # Wölfel 2020, Pan 2020, Kim 2020, To 2020 (first 2 weeks after symptom onset)
+source ("src/process_TEhandtosaliva.R") # Pitol 2017 (dry transfer from hand to saliva)
+source ("src/process_TEsurfacetohand.R") # Lopez 2013 (high himidity)
+source ("src/process_decay.R") # van Doremalen 2020
+source ("src/process_doseresponse.R") # QMRAwiki, using the from 0.5th, 50th, and 99.5 th percentiles as min, mode, and max
+source ("src/process_Asf.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
+source ("src/process_prevalence.R") # Fractional surface area, @EPA2011 and @AuYeung2008 (adult front partial finger)
+source ("src/process_disinfection.R") # Log reduction by disinfection, @Hulkower2011, @Sattar1989 for Chlorine and Ethanol
+
+# Simulation parameters:
+simNum = 100 # Number of simulations
+# ----- #
+
+# General parameters
+GC_Inf <- runif(simNum, 100, 1000) # [copies/TCID50] Proportion of infective gene copies [GC:TCID50], Ip (2015) 
+adj_DoseResp <- runif(simNum, 0.001, 0.01) # Adjustments for intranasal administration, @Jones2020
+Vs <- runif(simNum, 0.0396, 0.0484) # [mL] Volume of saliva per cough, Nicas and Jones (2009), mean 0.044ml /pm 10%
+# Ncough_min <- rtruncnorm(simNum, 0, Inf, 0.57, 1) # [cough/min] Number of coughs per minute, Leung (2020)
[…]

### 2020-09-11 src/eliminate/plot_risk.R

@@ -41,8 +41,32 @@ median_risk$Disinfection <- revalue(median_risk$Disinfection , c("0"="None","1"=
 p2 <- ggplot(median_risk, aes(prevalence, risk)) + geom_point(alpha = 1, aes(shape = Disinfection, color = Disinfection)) +
     scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                   labels = trans_format("log10", math_format(10^.x))) +
+                  scale_size_manual(values=c(2,7,7,7))+
     theme_bw()
 p2
+
+
+
+# --------- TEST PLOTS
+temporal <- read.csv (file="data/processed/temporal.csv", header = TRUE, sep=",")
+
+p3 <- ggplot(temporal, aes(Disinfection, pMed)) + geom_point(alpha = 1, size=8, aes(color = intervention)) +
+    scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                  labels = trans_format("log10", math_format(10^.x))) +
+    theme_bw()
+p3
+
+p4 <- ggplot(temporal, aes(Disinfection, pLow)) + geom_point(alpha = 1, size=8, aes(color = intervention)) +
+    scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                  labels = trans_format("log10", math_format(10^.x))) +
+    theme_bw()
+p4
+
+p5 <- ggplot(temporal, aes(Disinfection, pHigh)) + geom_point(alpha = 1, size=8, aes(color = intervention)) +
+    scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                  labels = trans_format("log10", math_format(10^.x))) +
+    theme_bw()
+p5
 
 
 

### 2020-09-11 src/plot_intervention.R

@@ -1,0 +1,275 @@
+#   Set up
+
+library(tidyverse) 
+library(ggpubr)
+
+
+# -----------------
+# - - - - - 
+# CONTROL     LOW
+
+surf_Low_0 <-  read.csv (file="data/processed/surf_pLow_d0.csv", header = TRUE, sep=",")
+cnt <- surf_Low_0$x
+
+
+# SURFACE
+surf_Low_1 <-  read.csv (file="data/processed/surf_pLow_d1.csv", header = TRUE, sep=",")
+surf_Low_2 <-  read.csv (file="data/processed/surf_pLow_d2.csv", header = TRUE, sep=",")
+surf_Low_3 <-  read.csv (file="data/processed/surf_pLow_d3.csv", header = TRUE, sep=",")
+
+s_a <- surf_Low_1$x
+s_b <- surf_Low_2$x
+s_c <- surf_Low_3$x
+
+surf_Low <- as.data.frame(rbind(s_a,s_b,s_c)) 
+surf_Low$intervention <- "surface"
+surf_Low$condition    <- c("Once (.7am)", "Once (12pm)", "Twice (7am,12pm)")
+surf_Low$prevalence   <- "Low Prevalence"
+
+# HAND
+hand_Low_1 <-  read.csv (file="data/processed/hand_pLow_d1.csv", header = TRUE, sep=",")
+hand_Low_2 <-  read.csv (file="data/processed/hand_pLow_d2.csv", header = TRUE, sep=",")
+hand_Low_3 <-  read.csv (file="data/processed/hand_pLow_d3.csv", header = TRUE, sep=",")
+
+h_a <- hand_Low_1$x
+h_b <- hand_Low_2$x
+h_c <- hand_Low_3$x
+
+hand_Low <- as.data.frame(rbind(h_a,h_b,h_c)) 
+hand_Low$intervention <- "hand"
+hand_Low$condition    <- c("Comp 25%", "Comp 50%", "Comp 75%")
+hand_Low$prevalence   <- "Low Prevalence"
+
+# MASK
+mask_Low_1 <-  read.csv (file="data/processed/mask_pLow_d1.csv", header = TRUE, sep=",")
+mask_Low_2 <-  read.csv (file="data/processed/mask_pLow_d2.csv", header = TRUE, sep=",")
[…]

### 2020-09-11 src/process_TEhandtosaliva.R

@@ -14,18 +14,11 @@ df_TE_hm <- read.csv (file=infile, header = TRUE, sep=",")
 
 #-----------------
 
-# dt [min] = drying time for average droplet size, assumtion. 
-# TE_hm_d [%] = transfer efficiency from hand to mucous membrane (dry)
-# TE_hm_w [%] = transfer efficiency from hand to mucous membrane (wet)
-# TE_hm [%]= transfer efficiency from hand to mucous membrane (dry and wet)
-
 TE_hm_d <- mean(df_TE_hm$pt_dry)/100 # mean TEhm when inoculum is dry
 sd_TE_hm_d <- sd(df_TE_hm$pt_dry)/100 # sd TEhm when inoculim is dry
 
-TE_hm_w <- mean(df_TE_hm$pt_wet)/100  # mean TEhm when inoculum is wet
-sd_TE_hm_w <- sd(df_TE_hm$pt_wet)/100  # sd TEhm when inoculum is wet
-
-# Creating a function to estimate the TEhm as a function of time after inoculation (t)
+# TE_hm_w <- mean(df_TE_hm$pt_wet)/100  # mean TEhm when inoculum is wet
+# sd_TE_hm_w <- sd(df_TE_hm$pt_wet)/100  # sd TEhm when inoculum is wet
 
 # output -------------------------------------------------------------------
 

### 2020-09-11 src/process_TEsurfacetohand.R

@@ -24,25 +24,5 @@ TE_sh_stl_sd <- df_TE_sh[14,4]
 TE_sh_pl_mean <- df_TE_sh[10,3]
 TE_sh_pl_sd <- df_TE_sh[10,4]
 
-#  -------------------------------------------------------------------
-
-# read and process data  --------------------------------------------------
-#infile = "data/raw/TEsurhandtosurface.csv"
-
-# read data 
-#df_TE_hs <- read.csv (file=infile, header = TRUE, sep=",")
-#df_TE_hs$pt_mean <- df_TE_hs$pt_mean / 100
-#df_TE_hs$pt_sd <- df_TE_hs$pt_sd / 100
-#df_TE_hs$pt_min <- df_TE_hs$pt_min / 100
-#df_TE_hs$pt_max <- df_TE_hs$pt_max / 100
-
-# Extracting data for steel at high humidity (40-65)
-#TE_hs_stl_mean <- df_TE_hs[X,X]
-#TE_hs_stl_sd <- df_TE_hs[X,X]
-
-# Extracting data for plastic at high humidity (40-65)
-#TE_hs_pl_mean <- df_TE_hs[X,X]
-#TE_hs_pl_sd <- df_TE_hs[X,X]
-
 
 

### 2020-09-11 src/process_concentrationsaliva.R

@@ -1,11 +1,16 @@
-# Processing data of concentration of SARS-CoV-2 samples in sputum (taken as saliva), and samples
+# Processing data of concentration of SARS-CoV-2 samples in sputum, and samples
 # taken from the uper respiratory track (urt). 
 
 # Data obtained from: 
     #"Viral load of SARS-CoV-2 in clinical samples" (Pan, 2020) 
-    #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020)
+    #"Virological assessment of hospitalized cases of coronavirus disease 2019" (Wölfel,2020); Figure 2_ Data
     #"Viral load kinetics of SARS-CoV-2 infection in first two patients in Korea" (Kim, 2020)
     #"Temporal profiles of viral load in posterior oropharyngeal saliva samples and serum antibody responses during infection by SARS-CoV-2: an observational cohort study" (To, 2020)
+
+
+# Assumptions: Since concentrations on sputum, uper respiratiry track (urt), and saliva were similar, we used 
+             # from sputum, urt and saliva in the concentration dataset.
+             # From the study @To2020, initial and peak concentrations were both used 
 
 # setup -------------------------------------------------------------------
 # external libraries

### 2020-09-11 src/process_convergence.R

@@ -1,0 +1,35 @@
+# Convergence compilations
+
+# Conditions
+# Material:       Metal
+# Humidity:       High
+# Surface:        Buttons
+# Inoculation:    Cough on hands followed by hand to surface transfer
+
+# Disinfection:  0
+# Prevalence: Medium
+
+
+# setup -------------------------------------------------------------------
+# external libraries
+ library(tidyverse) 
+
+# read and process data  --------------------------------------------------
+
+infile1 = "data/processed/compilation_risk.csv" # Compilation of all risks for 10,000 simulations
+infile2 = "data/processed/convergence_risk.csv" # Risk for baseline scenario for 100,000 simulations
+infile3 = "data/processed/convergence_risk_20.csv" # Risk for baseline scenario for 20,000 simulations
+infile4 = "data/processed/convergence_risk_50.csv" # Risk for baseline scenario for 50,000 simulations
+
+# outfile = "data/processed/TEhandtosaliva.csv"
+
+# read data 
+
+df_convergence_1 <- read.csv (file=infile1, header = TRUE, sep=",")
+df_convergence_2 <- read.csv (file=infile2, header = TRUE, sep=",")
+df_convergence_3 <- read.csv (file=infile3, header = TRUE, sep=",")
+df_convergence_4 <- read.csv (file=infile4, header = TRUE, sep=",")
+
+
+
+

### 2020-09-11 src/process_disinfection.R

@@ -11,6 +11,14 @@
 
 hand_dis <- 4.25
 
+
+
+# Regarding Hand Hygene. It is recomended by the WHO that: 
+  #  @WHO2010_HandHygene
+  #  An effective alcohol-based hand rub product should contain between 60% and 80% of alcohol 
+  #  its efficacy should be proven according to the European Norm 1500. These products can be purchased on the market,
+  #  but can also be produced locally in pharmacies using the formula and instructions provided by WHO.
+
 # ----------------- Surface Disinfection data ---------------------------------
 
 #   Article             virus (enveloped)     log10 Red     Sanitizer

### 2020-09-11 src/process_handinoculationarea.R

@@ -1,0 +1,12 @@
+
+
+# Processing the surface area inculated by the cough 
+
+
+simNum <- 100
+
+
+# r = r =  L tan (angle)
+# Angle = 50 to 75 degrees
+
+

### 2020-09-11 src/process_mask.R

@@ -1,0 +1,11 @@
+# Collecting data for droplet reduction by face-masks
+
+#   @Davies2013
+#   Method: Henderson apparatus, collison nebulizer (air particles, no droplets)
+#   % Filtration Efficiency (fraction of total virus filtered out by the material) for bacteriophage MS2   
+#     Min     Material    --   Max    Material
+#     48.87   Scarf            89.52  Surgical Mask
+ 
+mask_min <- 0.4887
+mask_max <- 0.8952
+
